---
title: "GMBC_data analysis"
author: "AL"
date: "2024-10-22"
output: pdf_document
--- 
       
# DATA READING AND CLEANING
  
    
Set Up 
```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

#General
library(tidyverse)
library(stringr)
library(glue)
library(openxlsx)

#Plotting
library(ggplot2)
library(ggtext)
library(gridExtra)
library(grid)
library(cowplot)

#Models
library(Publish)
library(purrr)
library(broom)

#Stats
library(tableone)
library(rstatix)
library(Maaslin2)

#WGS
library(vegan)

```
  
```{r estrobolome targets, control ids, WGS filter}

## Estrobolome data
estrobolome_targets <- read_csv("Estrobolome targets_2025-09-04.csv")

## List of Plate_IDs for Control Samples
technical_ids <- c("GMBC_00307","GMBC_00412","GMBC_00408", #plate1 012_T1-T3
                   "GMBC_00617","GMBC_00911","GMBC_00807", #plate2 079_T1-T3
                   "GMBC_01204","GMBC_01216","GMBC_01120", #plate3 023_T1-T3 (part 2)
                  "GMBC_01401","GMBC_01315","GMBC_01415",#plate4 012_T4-T6
                  "GMBC_01403","GMBC_01310","GMBC_01402", #plate4 079_T4-T6
                  "GMBC_01409","GMBC_01314","GMBC_01407") #plate4 023_T4-T6.P1 (part 1)

control_ids <- c("GMBC_00508","GMBC_00609","GMBC_00723","GMBC_00416", #plate1 NC_01-04
                  "GMBC_00212","GMBC_00801","GMBC_00305","GMBC_00902", #plate2 NC_05-08
                  "GMBC_01002","GMBC_01114","GMBC_01201", #plate3 NC_09-11
                 "GMBC_01414","GMBC_01317") #plate4 NC_12,NC_13


## Filter WGS data by a meta (epi) variable 
wgs_filter <- function(data, variable, var_value) {
  #convert alpha to a symbol for NSE
  alpha <- rlang::ensym(variable)
  #data modifications
  gmbc_data %>% 
    filter(!is.na(plate_ID)) %>%
    #pull out needed meta from gmbc_data
    select(plate_ID, !!alpha) %>%
    #add onto the metaphlan data
    merge(x=data, y=., by="plate_ID", all.x=TRUE) %>%
    #filter by the added meta variable
    filter(!!alpha==var_value) %>% 
    #drop meta var so only meteagenomic
    select(-!!alpha)
}


## Filter metabolite data by collection part
metabo_filter.p1 <- function(metabo_data) {
  metabo_data %>%
    merge(gmbc_data %>% select(study_id, collection_part), by = "study_id", all.x = TRUE) %>%
    filter(collection_part == "1") %>%
    select(-collection_part)
}


``` 
   
gmbc_data (REDCap, CDHQIII)             
```{r gmbc_data read in, impute}

#Read in gmbc_data from Data Cleaning.Rmd  
gmbc_data <- read.csv("gmbc_data_2025-08-22.csv") %>% 
  #filter for only part 1
  filter(collection_part == 1)

## Converting Data to Factor
gmbc_data <- gmbc_data %>%
  mutate(
    #selecting redcap variables only
      across(
      .cols = study_id:t_biol.occur,
      .fns = ~ if (!(cur_column() %in% 
              #numerical variables (for NOT in this list)
              c("s_age","s_height_cm","s_bmi","d_household_adults", "d_household_children","METmin",
              "rh_menarchy_age","s_menopause_age","rh_contraceptive_totalyr","rh_pregnancy_num",
              "rh_pregnancy_live","rh_pregnancy_age_first","rh_breastfeed_months",
               "rh_pregnancy_age_last","t_diameter.avg") 
        )) {
          as.factor(.)
        } else {
          .
        }
    ),
  #re-leveling factors
  ethnic_collapsed = factor(ethnic_collapsed, 
      levels = c("White","Asian","Other")),
  d_education_collapsed = factor(d_education_collapsed, 
    levels = c("<= High school","Post-high school training","Bachelor's","Postgraduate")),
  d_income_collapsed = factor(d_income_collapsed, 
      levels = c("49k","50k","100k","150k","PNA")) 
  )

## Updating GMBC variables for adjustment
gmbc_data <- gmbc_data %>% 
  mutate(
    #updates for control who became a case
    plate_ID = case_when(
      study_id == "GMBC_048" ~ "GMBC_00803",
      study_id == "GMBC_048_E2" ~ "GMBC_01318",
      TRUE ~ plate_ID ),
    #age at first pregnancy
    rh_pregnancy_age_first_cat = case_when(
        (rh_pregnancy_age_first >= 30) ~ as.factor(3), #geriatric pregnancy
        (rh_pregnancy_age_first < 30 & rh_pregnancy_age_first >= 25) ~ as.factor(2),
        (rh_pregnancy_age_first < 25) ~ as.factor(1),
        TRUE ~ as.factor(0) ),
    #smoking never / ever
    sm_summary = case_when(
        (sm_summary == 0) ~ as.factor(0),
        (sm_summary %in% c(1,2)) ~ as.factor(1),
        TRUE ~ NA ),
    #education level
    d_education_collapsed = case_when(
      (d_education_collapsed %in% c("<= High school","Post-high school training") ) ~ as.factor("< Bachelor's"),
      TRUE ~ d_education_collapsed ),
    d_education_collapsed = factor(d_education_collapsed), #refactor to collapse levels
    #ethnicity
    ethnic_collapsed = case_when(
      ethnic_collapsed %in% c("Asian","Other") ~ as.factor("Non-White"),
      TRUE ~ ethnic_collapsed),
    ethnic_collapsed = factor(ethnic_collapsed), #refactor to collapse levels
    
  #Dietary Vars (DENSITY ADJUSTMENT)
  Total.dietary.fibre.g_density = (Total.dietary.fibre.g / Energy.kcal) * 1000,
  Total.saturated.fatty.acids.g_density = 
    (Total.saturated.fatty.acids.g / Energy.kcal) * 1000,
  Total.sugar.g_density = (Total.sugar.g / Energy.kcal) * 1000
)


``` 
 
MetaPhlAn4 (stool)         
```{r metaphlan data read and cleaning}

## Read in the data
metaphlan4_all <- 
  #force three cols to properly read in
  read.table("profiled_metagenome.tsv", sep = "\t", fill = TRUE, col.names = c("plate_ID", "clade_name", "clade_taxid", "relative_abundance", "coverage", "estimated_number_of_reads_from_the_clade"), quote = "", comment.char = "") %>%
  #drop front matter
  slice(-(1:329)) %>% 
  mutate(name = str_extract(clade_name, "[^|]+$"),
         #need to convert to numeric
         across(relative_abundance:estimated_number_of_reads_from_the_clade, as.numeric)) %>% 
  #derive rank
  mutate(rank = case_when(
    grepl("k__", name) ~ "kingdom",
    grepl("p__", name) ~ "phylum",
    grepl("c__", name) ~ "class",
    grepl("o__", name) ~ "order",
    grepl("f__", name) ~ "family",
    grepl("g__", name) ~ "genus",
    grepl("s__", name) ~ "species",
    grepl("t__", name) ~ "subspecies",
    TRUE ~ "other"
  ))


## Data Selection

#Data selection
metaphlan4 <- metaphlan4_all %>% 
  # Drop the negative and technical controls
  filter(!plate_ID %in% control_ids) %>%
  filter(!plate_ID %in% technical_ids) %>% 
  # SELECT PART 1 SAMPLES ONLY
  wgs_filter(., collection_part, 1)

#The third percentile of data = 0.00078 (which is almost 0.001% as used in Robertson 2023), 4=0.00104, 5=0.00132
min_abundance_threshold <- quantile(metaphlan4 %>% pull(relative_abundance), probs = 0.03)
min_abundance_threshold
#Calculating frequency of taxa present
filtered_taxa <- metaphlan4 %>%
  #binary presence/absence matrix
  mutate(presence = ifelse(is.na(relative_abundance), FALSE, relative_abundance > min_abundance_threshold)) %>% #0
  #calc presence frequency for each taxon
  group_by(name) %>%
  summarize(
    presence_count = sum(presence), #243 is the max (should be 241?)
    frequency = presence_count / n_distinct(metaphlan4$plate_ID) #presence/total count to get %
  )
#Filter taxa based on prevalence thresholds
filtered_taxa10 <- filtered_taxa %>% filter(frequency >= 0.1) %>% pull(name)
#Data Filtering
metaphlan4_long <- metaphlan4 %>% 
  filter(name %in% filtered_taxa10) %>% 
  arrange(plate_ID)


## Rename the subspecies to include the species in the name (more informative)
metaphlan4_long <- metaphlan4_long %>% 
  mutate(name = if_else( rank == "subspecies", #if subspecies level
                         str_extract(clade_name, "s__[^|]+\\|t__[^|]+"), #rename to include species
                         name) ) #otherwise leave


## Transforming to wider form
metaphlan4_wide <- metaphlan4_long %>% 
  filter(rank %in% c("species","genus","subspecies")) %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s

## Transforming to wider form for SPECIES data
metaphlan4_wide.s <- metaphlan4_long %>% 
  filter(rank=="species") %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s

## Transforming to wider form for GENUS data
metaphlan4_wide.g <- metaphlan4_long %>% 
  filter(rank=="genus") %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s

## Transforming to wider form for SUBSPECIES data
metaphlan4_wide.t <- metaphlan4_long %>% 
  filter(rank=="subspecies") %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s


```   

HUMAnN3 (stool)
```{r humann.path read and clean}

## Read in cleaned data set
humann.path_nopipe <- read.csv("humann.path_nopipe.csv") %>% 
  # Drop the negative and technical controls
  filter(!plate_ID %in% control_ids) %>%
  filter(!plate_ID %in% technical_ids) %>% 
  # select baseline samples only
  wgs_filter(., collection_part, 1)  %>% 
  rename(relative_abundance = value)


## Data Filtering

#third percentile of data
min_abundance_threshold <- humann.path_nopipe %>% 
  filter(!name %in% c("UNMAPPED","UNINTEGRATED")) %>% 
  pull(relative_abundance) %>% 
  quantile(., probs = 0.03)

#calculating frequency of taxa present
filtered_path5 <- humann.path_nopipe %>%
  #binary presence/absence matrix
  mutate(presence = ifelse(is.na(relative_abundance), FALSE, relative_abundance > min_abundance_threshold)) %>%
  #calc presence frequency for each taxon
  group_by(name) %>%
  summarize(
    presence_count = sum(presence),
    #presence/total count to get %
    frequency = presence_count / n_distinct(humann.path_nopipe$plate_ID)
  ) %>%
  #filter taxa based on the 5% threshold
  filter(frequency >= 0.05) %>%
  #pull out names
  pull(name)


#apply
humann.path_long <- humann.path_nopipe %>% 
  #remove unintegrated and unmapped samples
  filter(!name %in% c("UNMAPPED","UNINTEGRATED")) %>% 
  #filtering dataframe for >5% frequency
  filter(name %in% filtered_path5)

## Transforming to wider form
humann.path_wide <- humann.path_long %>% 
  pivot_wider(id_cols=c(plate_ID), #sample stays as row
              names_from=name, #name becomes column with rel abun as value
              values_from=relative_abundance,
              values_fill = 0)  #covert NA relative abundance to 0s


``` 

```{r humann.enzymes_nopipe read and clean}

## Read in cleaned data set
humann.enzymes_nopipe <- read.csv("humann.enzymes_nopipe.csv") %>% 
  # Drop the negative and technical controls
  filter(!plate_ID %in% control_ids) %>%
  filter(!plate_ID %in% technical_ids) %>% 
  # select baseline samples only
  wgs_filter(., collection_part, 1)  %>% 
  rename(relative_abundance = value)
  

## Data Filtering

#identify third percentile of data
min_abundance_threshold <- quantile(humann.enzymes_nopipe %>% pull(relative_abundance), probs = 0.03)

#calculate frequency of taxa present
filtered_enzyme5 <- humann.enzymes_nopipe %>%
  #binary presence/absence matrix
  mutate(presence = ifelse(is.na(relative_abundance), FALSE, relative_abundance > min_abundance_threshold)) %>%
  #calc presence frequency for each taxon
  group_by(name) %>%
  summarize(
    presence_count = sum(presence),
    #presence/total count to get %
    frequency = presence_count / n_distinct(humann.enzymes_nopipe$plate_ID)
  ) %>% 
  #filter taxa based on the 5% threshold
  filter(frequency >= 0.05) %>%
  #pull out names
  pull(name)

#applying filter for >5% frequency
humann.enzymes_long <- humann.enzymes_nopipe %>%
  filter(name %in% filtered_enzyme5)

## Transforming to wider form
humann.enzymes_wide <- humann.enzymes_long %>% 
  pivot_wider(id_cols = c(plate_ID), #sample stays as row
              names_from = name, #Enzyme name has special characters that crash
              values_from = relative_abundance,
              values_fill = 0)  #covert NA relative abundance to 0s


```

Metabolomics (blood & stool)   
```{r metabo plasma - reading and cleaning}

#Read in metabolomics abundances
plasma_abund <- read.csv("GMBC_Metabolite_Results.plasma_2025-04-30.csv") %>% 
  # SELECT PART 1 SAMPLES ONLY
  metabo_filter.p1()

## Calculating prevalence frequency of metabolites present
filtered_metabo10 <- plasma_abund %>%
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relab") %>%
  mutate(presence = ifelse(is.na(relab), FALSE, relab > 0)) %>%
  group_by(Metabolite) %>%
  summarize(frequency = sum(presence) / n_distinct(plasma_abund$study_id)) %>%
  filter(frequency >= 0.10) %>%
  pull(Metabolite)


## Data Filtering >10% prevalence frequency
metabo.plasma_wide <- plasma_abund %>%
  select(study_id, all_of(filtered_metabo10)) %>% 
  merge(., gmbc_data %>% select(study_id), by="study_id", all.x=T)

## Long form data
metabo.plasma_long <- metabo.plasma_wide %>% 
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relative_abundance")


``` 
  
```{r metabo stool - reading and cleaning}

#Clean up the study codes
stool_abund <- read.csv("GMBC_Metabolite_Results.stool_2025-04-30.csv") %>% 
  # SELECT PART 1 SAMPLES ONLY
  metabo_filter.p1()


## Calculating prevalence frequency of metabolites present
filtered_metabo10 <- stool_abund %>%
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relab") %>%
  mutate(presence = ifelse(is.na(relab), FALSE, relab > 0)) %>%
  group_by(Metabolite) %>%
  summarize(frequency = sum(presence) / n_distinct(stool_abund$study_id)) %>%
  filter(frequency >= 0.10) %>%
  pull(Metabolite)


## Data Filtering >10% prevalence frequency (metabo_wide is filtered)
metabo.stool_wide <- stool_abund %>%
  select(study_id, all_of(filtered_metabo10)) %>% 
  merge(., gmbc_data %>% select(study_id), by="study_id", all.x=T)

## Long form data
metabo.stool_long <- metabo.stool_wide %>% 
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relative_abundance")


``` 
   

   
# ANALYSES FUNCTIONS 

```{r alpha and beta diversity - functions (4) }

#alpha is within sample diversity - what and how much is there in a sample
#beta is between sample diversity - how similar is one sample to another
 
## Calculate Alpha Diversity with Sig Tests (Wilcoxon Rank Sum)
run_alpha_diversity <- function(data_long, metaphlan=FALSE) {
  
  ## Calculate alpha diversity as new cols
  alpha_data <- data_long %>% 
    #conditional filtering, if metaphlan==T select species, default is FALSE
    { if (metaphlan) filter(., rank == "subspecies") else . } %>%
    #{ if (metaphlan) filter(., rank == "genus") else . } %>%
    group_by(plate_ID) %>% #all species in a sample
    #calculate alpha metrics
    summarize(shannon = diversity(relative_abundance, "shannon"),
              richness = specnumber(relative_abundance)) %>% 
    #merge alpha metrics onto gmbc_data
    merge(gmbc_data %>% select(plate_ID, case_con),
          ., by="plate_ID", all.x = TRUE)
  
  ## Sig testing for case VS controls (wilcoxon rank sum) 
  
    #SHANNON
    #calculate group medians alpha div
    meds <- alpha_data %>% 
      #filter(shannon > 0) %>% #only include where alpha data
      group_by(case_con) %>% #group by cancer status
      summarise(median_alpha = median(shannon), .groups = "drop") #calculate group medians
    #calculating p values using wilcoxon rank sum test
    p.shannon <- wilcox.test(shannon ~ case_con, data = alpha_data) #%>%filter(shannon>0))
    #print out metrics
    print(glue("The group medians of SHANNON diversity:
             Controls = {meds$median_alpha[meds$case_con == 0]}
             Cases = {meds$median_alpha[meds$case_con == 1]}
             Abs median difference: {abs(diff(meds$median_alpha))}
             The p value (Wilcoxon): {p.shannon$p.value}"))

    #RICHNESS
    #calculate group medians alpha div
    meds <- alpha_data %>% 
      #filter(richness > 0) %>% #only include where alpha data
      group_by(case_con) %>% #group by cancer status
      summarise(median_alpha = median(richness), .groups = "drop") #calculate group medians
    #calculating p values using wilcoxon rank sum test
    p.richness <- wilcox.test(richness ~ case_con, data = alpha_data)#%>%filter(richness>0))
    #print out metrics
    print(glue("The group medians of RICHNESS:
             Controls = {meds$median_alpha[meds$case_con == 0]}
             Cases = {meds$median_alpha[meds$case_con == 1]}
             Abs median difference: {abs(diff(meds$median_alpha))}
             The p value (Wilcoxon): {p.richness$p.value}"))
  
  return(list(alpha_data = alpha_data, 
            p_shannon = p.shannon$p.value, 
            p_richness = p.richness$p.value))
}

## Plot Individual Alpha Plots for Shannon + Richness
plot_alpha_diversity <- function(data, alpha, y_title, p.val, x.size, y.size) {
  
  alpha <- rlang::ensym(alpha) # Convert alpha to a symbol for NSE
  
  alpha.plot <- data %>% 
  filter(!is.na(plate_ID)) %>% 
  ggplot(aes(x=as.factor(case_con), y=!!alpha, color=as.factor(case_con))) + 
    geom_boxplot(show.legend = FALSE) +
    geom_jitter(width = 0.2, alpha=0.3, show.legend = FALSE) + 
    labs(x=element_blank(), y=y_title) +
    scale_color_manual(values = c("0"="darkgray", "1"="blue")) +
    theme(panel.background = element_blank(),
          plot.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x = element_blank(), 
          axis.ticks.x = element_blank()) + 
    theme(panel.border = element_rect(color = "black", fill = NA), 
          panel.spacing = unit(0.5, "lines")) +
    annotate("text", x=x.size, y=y.size, label = paste0("p = ", round(p.val, digits=4)),
             color = "black", size = 3)
  
  return(alpha.plot)
}

## Run Beta Diversity (Bray-Curtis Distances), Sig test PERMANOVA, PCA, plot
plot_beta_diversity <- function(data_wide, x.size, y.size) {

  ## Calculate Bray-Curtis Distance Matrix
  dist_mat1 <- data_wide %>% 
    arrange(plate_ID) %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc matrix of distances using vegan function


  ## Significant testing between groups (PERMANOVA)
    #pulling out grouping variable
    groups1 <- merge(
        data_wide %>% select(plate_ID),
        gmbc_data %>% select(plate_ID, case_con, collection_part),
        by="plate_ID"
      ) %>% 
      arrange(plate_ID) %>% 
      pull(case_con) %>% #pull out the case_con variables
      factor(levels = c("0", "1")) #set the factor levels
    #apply permanova function
    beta_p <- adonis2(dist_mat1 ~ groups1)$`Pr(>F)`[1]
    beta_R2 <- adonis2(dist_mat1 ~ groups1)$`R2`[1]


  ## Calculating PCOAs
    #apply pcoa
    beta_pcoa <- cmdscale(dist_mat1, #pcoa is list with $points (PCs per sample)
                       k = 8, #N-1 PCs (default 2)
                       eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                       add = TRUE) #added add=T from Sandeep
    #convert eigenvalues to percentage
    beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)
    #turn pcoa points from list to dataframe
    beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
      rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
      rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
      merge(gmbc_data, ., by="plate_ID", all.x = TRUE) #merge PC values onto data

    #calculate the median coordinates for each group
    medians <- beta_pcoa.df %>%
      filter(!is.na(plate_ID)) %>%
      group_by(case_con) %>%
      summarize(med_PC1 = median(PC1), med_PC2 = median(PC2))

  ## Plotting PCOAs
  beta_plot <- beta_pcoa.df %>% 
    filter(!is.na(plate_ID)) %>% 
    ggplot(aes(x = PC1, y = PC2, color = case_con)) + 
    geom_point() +
    stat_ellipse(aes(x = PC1, y = PC2), level = 0.95) +
    scale_color_manual(values = c("0" = "darkgray", "1" = "blue")) + 
    labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
         y = paste0(
                "<span style='font-size:10pt'>Beta Diversity</span><br>",
                "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"
              ),
           #glue("PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)"),
         color = "Case Status") + 
    theme(legend.position = "none",
          axis.title.y = element_markdown(),
          panel.background = element_blank(),
          plot.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.border = element_rect(color = "black", fill = NA),
          panel.spacing = unit(0.5, "lines")) +
    annotate("text", x = x.size, y = y.size,
             label = glue("p = {round(beta_p, 3)}\nR2 = {round(beta_R2, digits = 3)}"),
             color = "black", size = 3) +
    #add lines from each point to the median of its group with matching colors
    geom_segment(data = beta_pcoa.df %>% 
                   filter(!is.na(plate_ID)) %>% 
                   left_join(medians, by = "case_con"),
                 aes(x = med_PC1, y = med_PC2, 
                     xend = PC1, yend = PC2, 
                     color = case_con),
                 linewidth = 0.2, alpha = 0.5) +
    #add median points with matching colors
    geom_point(data = medians, 
               aes(x = med_PC1, y = med_PC2, color = case_con), 
               size = 3, shape = 4)  # Shape 4 is an 'x' marker

  return(beta_plot)
  }

## Plot Alpha and Beta Diversity - Unified Plot
plot_alpha_beta <- function(alpha_data, beta_plot, alpha_plot1, alpha_plot2, letter1, letter2, letter3, filename) {
  
  # Open a PNG device
  png(filename, width = 4 * 600, height = 4 * 600, res = 600, bg = "white")
  
  ##Pull out a Legend
    #build shannon plot
    ap1 <- alpha_data %>% 
      filter(!is.na(plate_ID)) %>% 
      mutate(case_con = factor(case_con, levels = c(0, 1), labels = c('Control', 'Case'))) %>% 
      ggplot(aes(x=case_con, y=shannon, color=case_con)) + 
        scale_color_manual(values = c("Control"="darkgray", "Case"="blue")) +
        geom_boxplot() +
        labs(color = "Cancer Status") + theme(panel.background = element_blank())
    #pulling out legend
    tmp <- ggplot_gtable(ggplot_build(ap1))
    leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    legend <- tmp$grobs[[leg]]
  
  ##Build the Bottom Row at a different width
  bottom_row <- arrangeGrob(legend, beta_plot, widths = c(1, 2))
  
  ##Arrange Plots as a Grid 
  grid.arrange(
        alpha_plot1,
        alpha_plot2,
        bottom_row,
        layout_matrix = rbind(c(1, 2),
                              c(3, 3)),
        heights = c(1, 1.75))

  ##Adding in A, B, C Text
  grid.text(letter1, x = unit(0.46, "npc"), y = unit(0.71, "npc"), 
            gp = gpar(fontsize = 15, fontface = "bold"))
  grid.text(letter2, x = unit(0.96, "npc"), y = unit(0.71, "npc"), 
            gp = gpar(fontsize = 15, fontface = "bold"))
  grid.text(letter3, x = unit(0.96, "npc"), y = unit(0.14, "npc"), 
            gp = gpar(fontsize = 15, fontface = "bold"))

  dev.off()
  
}


```
   
```{r data transform - function }

transform_metagenomics <- function(data_wide, merge = TRUE) {
  
  #Calculate the pseudocount as half the smallest non-zero value across all features
  pseudocount <- 0.5 * min(data_wide %>% 
                             select(-plate_ID) %>% 
                             unlist() %>% 
                             .[. > 0], 
                           na.rm = TRUE)
  
  #Transform metagenomic data
  metaphlan.models <- data_wide %>% 
    #CLR transformation
     mutate(across(-plate_ID, 
                   ~ log(. + pseudocount) - mean(log(. + pseudocount), 
                                                    na.rm = TRUE)))# %>% 
    
  # Merge with metadata OR skip
  if (merge) {
    metaphlan.models <- merge(metaphlan.models, gmbc_data, by="plate_ID", all.x = TRUE)
  } else {
    metaphlan.models <- metaphlan.models
  }
  
  return(metaphlan.models)
  
}

transform_metabo <- function(data_wide, remove_outliers = TRUE, merge = TRUE) {
  
  # Step 1: Calculate the pseudocount as half the smallest non-zero value
  pseudocount <- 0.5 * min(
    data_wide %>%
      select(-study_id) %>%
      unlist() %>%
      .[. > 0],
    na.rm = TRUE
  )
  
  # Step 2: Log2
  data_log <- data_wide %>%
    mutate(across(-study_id, ~ log2(.x + pseudocount)))

  # Step 3: Apply OR Skip - Mask outliers in raw data (|z| > 4)
  if (remove_outliers) {
    data_clean <- data_log  # copy of raw data
    for (col in names(data_clean)[-1]) {
      #zscore scale
      data_logz <- data_clean %>%
        mutate(across(-study_id, ~ scale(.x)[, 1])) 
      #pull out scores
      z_scores <- data_logz[[col]]
      data_clean[[col]][abs(z_scores) > 4] <- NA
    }
  } else {
    data_clean <- data_log
  }

  # Step 5: Merge with metadata OR skip
  if (merge) {
    data_final <- merge(data_clean, gmbc_data, by = "study_id", all.x = TRUE)
  } else {
    data_final <- data_clean
  }
  
  return(data_final)
  
}

transform_metabo2 <- function(data_wide, remove_outliers = TRUE, transform = TRUE, merge = TRUE) {
  
  # Step 1: Calculate the pseudocount as half the smallest non-zero value
  pseudocount <- 0.5 * min(
    data_wide %>%
      select(-study_id) %>%
      unlist() %>%
      .[. > 0],
    na.rm = TRUE
  )
  
  # Step 2: Log2 + z-score for outlier detection
  data_logz <- data_wide %>%
    mutate(across(-study_id, ~ log2(.x + pseudocount))) %>% 
    mutate(across(-study_id, ~ scale(.x)[, 1]))

  # Step 3: Apply OR Skip - Mask outliers in raw data (|z| > 4)
  if (remove_outliers) {
    data_clean <- data_wide  # copy of raw data
    for (col in names(data_clean)[-1]) {
      z_scores <- data_logz[[col]]
      data_clean[[col]][abs(z_scores) > 4] <- NA
    }
  } else {
    data_clean <- data_wide
  }
  
  # Step 4: Conditionally transform based on masked raw stats OR skip
  if (transform) {
    data_transformed <- data_clean %>%
      mutate(across(-study_id, ~ {
        x <- .x
        x_log <- log2(x + pseudocount)
        if (sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) < 0.55) {
          scale(x_log)[, 1]
        } else {
          x_log #otherwise leave as log2
        }
      }))
  } else {
    data_transformed <- data_clean
  }
  
  # Step 5: Merge with metadata OR skip
  if (merge) {
    data_final <- merge(data_transformed, gmbc_data, by = "study_id", all.x = TRUE)
  } else {
    data_final <- data_transformed
  }
  
  return(data_final)
  
}

```
  
```{r adjustment sets}

get_covariates <- function(){
  
  confounder_vars <- c("s_age","s_bmi","ethnic_collapsed","d_education_collapsed", "hormone_current.p1","sm_summary","pa_cat","Alcohol.g"
  )
  
  diet_vars <- c(confounder_vars,
    #energy-adjusted values
    "Total.dietary.fibre.g_density",
    "Total.saturated.fatty.acids.g_density",
    "Total.sugar.g_density"
  )
  
  adjustment_sets <- list(unadj = c(),
                          confounder_vars = confounder_vars,
                          diet_vars = diet_vars)

  return(adjustment_sets)

}

```

Logistic Regression       
```{r iterative modeling - function }
  
## Build the function to apply across multiple variables 
iterative_modeling <- function(column_name, data_set, model_vars) {
  
  #drop any NA for the variable of interest
  data_set <- data_set %>%
    drop_na(!!sym(column_name)) #turn to symbol to ref
  
  # Calculate sample size N
  N <- nrow(data_set)
  
  #construct the formula with the changing column as the first predictor and covariates
  formula <- as.formula(
    paste0(
      #include backticks to allow special characters in the model formula
      "case_con ~ `", column_name, "`",
      #if covariates, add in, otherwise if unadjusted model add nothing
      if (length(model_vars) > 0) paste0("+", paste(model_vars, collapse = " + ")) else ""
      )
    )
  
  #fit the model
  model <- glm(formula, 
               data = data_set, 
               family = binomial)
  
  #print model summary for first taxa, enzyme, metabolite
  if (column_name %in% c("s__Anaerotruncus_rubiinfantis|t__SGB25416","s__Escherichia_coli","g__Alistipes","`Genistein`","`1.1.1.62`","Genistein")) { 
    print(formula)
    print(summary(model))
  }
  
  #pull out results
  result <- tidy(model, conf.int = TRUE)  %>%
    
    #filter to pull out row of interest
    filter(term == paste0("`", column_name, "`") | term == column_name) %>% 
    
    #calc HR and CIs
    mutate( OR = exp(estimate), #hazard Ratio
            conf_low = exp(conf.low), #confidence Interval Lower Bound
            conf_high = exp(conf.high), #confidence Interval Upper Bound
            OR_CI = paste0(round(OR, 2), " (",
                                round(conf_low, 2), "–",
                                round(conf_high, 2), ")")
            ) %>%
    
    #select relevant terms
    select(term, OR, conf_low, conf_high, OR_CI, p.value)
  
  #add col for sample size
  result$N <- N
  
  #return result
  return(result) 
  
}

```     

```{r iterative modeling - apply } 

library(writexl)
## PRE DEFINE VARIABLES

#predictor_vars
#adjustment_sets 
#dataset.model



## APPLY THE FUNCTION 

apply_iterative.modeling <- function(predictor_vars, adjustment_sets, dataset.model, file_name) {
  
  #Create empty results list
  df_results <- list()
  #Create list to store estrobolome-only results across all models
  estro_summary_list <- list()
    
  # Clean estrobolome list (remove NA)
  estrobolome_list <- c(estrobolome_targets$metaphlan_GMBC,
                        estrobolome_targets$Enzyme_code,
                        estrobolome_targets$Pathway_code)
  estrobolome_list <- Filter(Negate(is.na), estrobolome_list)
  
  #Loop through models for each adjustment level
  for (i in names(adjustment_sets) ) {
    
    #select what to include in the adjustment set
    model_vars = adjustment_sets[[i]]
    
    #Remove NA vars
    dataset.model <- dataset.model %>% 
      filter(if_all(all_of(model_vars), ~ !is.na(.)))
    
    #Apply the function to each column in the selected range and combine results
    log_model.results <- map_dfr(predictor_vars, function(column_name) {
      iterative_modeling(column_name = column_name, 
                         data_set = dataset.model, 
                         model_vars = model_vars)
    }) 
    
    #Calculate a q value (FDR adjustment for multiple testing)
    log_model.results <- log_model.results %>% 
      mutate(q.value = round(p.adjust(p.value, method = "BH"), 3)) %>% 
      arrange(p.value) %>% 
      select(term, OR_CI, p.value, q.value, N, OR, conf_low, conf_high)
    
    # === Filter for estrobolome targets and tag with adjustment ===
    estro_subset <- log_model.results %>%
      filter(str_detect(term, paste(estrobolome_list, collapse = "|"))) %>%
      mutate(adjustment = i)
    
    #Clean up the terms
    log_model.results <- log_model.results %>%
      mutate(term = gsub("\\.", " ", term)) %>%
      mutate(term = gsub("`", "", term))
    estro_subset <- estro_subset %>%
      mutate(term = gsub("\\.", " ", term)) %>%
      mutate(term = gsub("`", "", term))
    
    # Store estrobolome results for this adjustment level
    estro_summary_list[[i]] <- estro_subset
    
    #Store the result in the list with a dynamic sheet name
    df_results[[paste0("Sheet_", i)]] <- log_model.results
    
  } 
  
  # Combine estrobolome summary across adjustment levels
  estrobolome_summary <- bind_rows(estro_summary_list)
      
  # Add estrobolome summary as a sheet in the output
  df_results[["Estrobolome_summary"]] <- estrobolome_summary
  
  #Save to Excel (multiple sheets)
  write_xlsx(df_results, sprintf("Logreg_%s_%s.xlsx", file_name, as.character(Sys.Date())))

}


```
 
 


# SAMPLE CHARACTERISTICS

```{r tableone}

#Variables of interest
table1_vars <- c(
  #demographics
  "s_age","s_height_cm","s_bmi",
  "ethnic_collapsed","d_education_collapsed",
  "health_condition","h_hbp","h_diabetes",
  "family_history",
  "rh_menarchy_age","s_menopause_age",
  "rh_contraceptive", "rh_contraceptive_totalyr",
  "rh_pregnancy_num","rh_pregnancy_live","rh_pregnancy_age_first",
  "rh_pregnancy_age_first_cat","rh_pregnancy_age_last","rh_breastfeed_months",
  "rh_hysterectomy", "rh_ovary",
  "hormone_ever","hormone_current.p1",
  "pa_cat","sm_summary",
  #Diet
  "Alcohol.g","Total.dietary.fibre.g",
  "Total.dietary.fibre.g_density",
  "Total.saturated.fatty.acids.g",
  "Total.saturated.fatty.acids.g_density",
  "Total.sugar.g",
  "Total.sugar.g_density", 
  #Tumor
  "t_er","t_pr","t_her2","t_stage","t_grade","t_foci","t_diameter.avg","t_location",
  "t_rads.occur","t_chemo.occur","t_endo.occur","t_biol.occur","t_bisp.occur",
  "t_rads.days_completed", "t_chemo.days_completed", "t_endo.days_medicated"
)


#hist(gmbc_data$Total.saturated.fatty.acids.g)

#Create list of nonnormal vals to run wilcoxon test on (only works for numerical vals, not categorical)
nonnormal_vars <- c(
  "s_age","s_height_cm","s_bmi","rh_menarchy_age",
  "s_menopause_age","rh_pregnancy_live","rh_breastfeed_months",
  "Alcohol.g","Total.dietary.fibre.g_density",
  "Total.saturated.fatty.acids.g_density","Total.sugar.g_density"
)

#running tableone function
gmbc.table1 <- gmbc_data %>% 
  filter(collection_part==1) %>% 
  filter(!is.na(Energy.kcal)) %>% #missing diet removed
  CreateTableOne(data=., 
                 vars= table1_vars,
                 strata="case_con", 
                 test=T, 
                 addOverall=T)
print(gmbc.table1, showAllLevels=T, nonnormal = nonnormal_vars)



```


# METAGENOMICS

## Quality and Controls

```{r metaphlan quality reads}

##### SAMPLE AND CONTROL QUALITY #####

metaphlan4_quality <- 
  #force three cols to properly read in
  read.table("profiled_metagenome_stats.tsv", sep = "\t", fill = TRUE, col.names = c("plate_ID", "reads_processed", "estimated_reads_mapped_to_known_clades"), quote = "", comment.char = "") %>% 
  #drop front matter
  slice(-(1:657)) %>% 
  mutate(
    reads_processed = as.numeric(reads_processed),
    estimated_reads_mapped_to_known_clades = as.numeric(estimated_reads_mapped_to_known_clades),
    non_annotated_reads = reads_processed - estimated_reads_mapped_to_known_clades,
    non_annotated_reads.perc = non_annotated_reads/reads_processed*100
    ) 


# Sample quality
metaphlan4_quality %>% 
  #drop the negative and technical controls
  filter(!plate_ID %in% control_ids) %>%
  filter(!plate_ID %in% technical_ids) %>% 
  #fiilter part 1 only
  merge(., gmbc_data %>% select(plate_ID, collection_part), by="plate_ID") %>% 
  filter(collection_part == 1) %>% 
  #summarize calcs
  summarize(
    reads_processed.mean = mean(reads_processed),
    reads_processed.sd = sd(reads_processed),
    reads_pr.min = min(reads_processed),
    reads_pr.max = max(reads_processed),
    non_annotated_reads.perc.mean = mean(non_annotated_reads.perc),
    non_annotated_reads.perc.sd = sd(non_annotated_reads.perc),
    non_annot.perc.min = min(non_annotated_reads.perc),
    non_annot.perc.max = max(non_annotated_reads.perc),
  )



# Negative control quality
metaphlan4_quality %>% 
  #filter in negative controls
  filter(plate_ID %in% control_ids) %>%
  #summarize calc
  summarize( no_controls = n(), #number of rows
             reads_processed.max = max(reads_processed),
             reads_processed.min = min(reads_processed),
             reads_processed.mean = mean(reads_processed),
             reads_processed.sd = sd(reads_processed))
  

# Technical Control Quality
metaphlan4_quality %>% 
  #filter in technical controls
  filter(plate_ID %in% technical_ids) %>% 
  #greating grouping var
  mutate(groups = case_when(
    plate_ID %in% c("GMBC_00307","GMBC_00412","GMBC_00408") ~ "012_P1", #plate1 012_T1-T3
    plate_ID %in% c("GMBC_00617","GMBC_00911","GMBC_00807") ~ "079_P1", #plate2 079_T1-T3
    plate_ID %in% c("GMBC_01204","GMBC_01216","GMBC_01120") ~ "023_P2", #plate3 023_T1-T3 (part 2)
    plate_ID %in% c("GMBC_01401","GMBC_01315","GMBC_01415") ~ "012_P2", #plate4 012_T4-T6
    plate_ID %in% c("GMBC_01403","GMBC_01310","GMBC_01402") ~ "079_P2", #plate4 079_T4-T6
    plate_ID %in% c("GMBC_01409","GMBC_01314","GMBC_01407") ~ "023_P1", #plate4 023_T4-T6.P1 (part 1)
    TRUE ~ NA )) %>%  
  group_by(groups) %>% 
  #summarize calc
  summarize( no_controls = n(), #number of rows
             reads_processed.max = max(reads_processed),
             reads_processed.min = min(reads_processed),
             reads_processed.mean = mean(reads_processed),
             reads_processed.sd = sd(reads_processed)
             )

##### TECHNICAL CONTROLS TAXONOMY #####

## Total Species Count ####

#select technical data
metaphlan4_all %>% 
  #select tehcnical Controls
  filter(plate_ID %in% technical_ids) %>% 
  #pivot to wider form
  filter(rank %in% c("species")) %>% #,"genus","subspecies")) %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0) %>%   #covert NA relative abundance to 0s
  #calculated row sums
  mutate(total_species_count = rowSums(select(., -1) > 0, na.rm = TRUE)) %>% 
  #greating grouping var
  mutate(groups = case_when(
    plate_ID %in% c("GMBC_00307","GMBC_00412","GMBC_00408") ~ "012_P1", #plate1 012_T1-T3
    plate_ID %in% c("GMBC_00617","GMBC_00911","GMBC_00807") ~ "079_P1", #plate2 079_T1-T3
    plate_ID %in% c("GMBC_01204","GMBC_01216","GMBC_01120") ~ "023_P2", #plate3 023_T1-T3 (part 2)
    plate_ID %in% c("GMBC_01401","GMBC_01315","GMBC_01415") ~ "012_P2", #plate4 012_T4-T6
    plate_ID %in% c("GMBC_01403","GMBC_01310","GMBC_01402") ~ "079_P2", #plate4 079_T4-T6
    plate_ID %in% c("GMBC_01409","GMBC_01314","GMBC_01407") ~ "023_P1", #plate4 023_T4-T6.P1 (part 1)
    TRUE ~ NA )) %>%  
  group_by(groups) %>%
  #calculate mean and SD of species count per group
  summarise(
    max_species_count = max(total_species_count, na.rm = TRUE),
    min_species_count = min(total_species_count, na.rm = TRUE), 
    mean_species_count = mean(total_species_count, na.rm = TRUE),
            sd_species_count = sd(total_species_count, na.rm = TRUE),
            n = n())



```

```{r Relative abundance plot}

## Phylum Relative abundances ####

phylum_long <- metaphlan4_all %>% 
  #select tehcnical Controls
  filter(plate_ID %in% technical_ids) %>% 
  #select only phylum level
  filter(rank=="genus") %>% 
  #greating grouping var
  mutate(groups = case_when(
    plate_ID %in% c("GMBC_00307","GMBC_00412","GMBC_00408") ~ "012_P1", #plate1 012_T1-T3
    plate_ID %in% c("GMBC_00617","GMBC_00911","GMBC_00807") ~ "079_P1", #plate2 079_T1-T3
    plate_ID %in% c("GMBC_01204","GMBC_01216","GMBC_01120") ~ "023_P2", #plate3 023_T1-T3 (part 2)
    plate_ID %in% c("GMBC_01401","GMBC_01315","GMBC_01415") ~ "012_P2", #plate4 012_T4-T6
    plate_ID %in% c("GMBC_01403","GMBC_01310","GMBC_01402") ~ "079_P2", #plate4 079_T4-T6
    plate_ID %in% c("GMBC_01409","GMBC_01314","GMBC_01407") ~ "023_P1", #plate4 023_T4-T6.P1 (part 1)
    TRUE ~ NA )) %>%  
  #collection parts
  mutate(collection_part = ifelse(str_detect(groups, "P1"), 1, 2)) %>% 
  # Aggregate and calculate proportions
  group_by(groups, collection_part, plate_ID, name) %>%
  summarize(total_abundance = sum(relative_abundance), .groups = 'drop') %>%
  group_by(groups, collection_part, plate_ID) %>%
  mutate(proportion = total_abundance / sum(total_abundance) * 100) %>%
  #drop where less than 5%
  filter(total_abundance >= 3) %>%
  rename(phylum = name) %>%   # Rename 'name' to 'phylum'
  mutate(groups2 = case_when (
            str_detect(groups, "012") ~ "012",
            str_detect(groups, "079") ~ "079",
            str_detect(groups, "023") ~ "023")) %>% 
  mutate(
    groups2 = factor(groups2, levels = c("012","023","079"),
                     labels = c("Participant 12","Participant 23","Participant 79")),
    collection_part = factor(collection_part, levels = c("1","2"),
                             labels = c("Baseline", "Follow Up"))
  ) %>% 
  mutate(
    phylum = ifelse(phylum == "g__Lachnospiraceae_unclassified", "Lachnospiraceae uncl.", phylum),
    phylum = gsub("^(g__|s__)", "", phylum),
    phylum = gsub("_", " ", phylum) 
    )


library(cowplot)
library(ggplot2)

p1 <- ggplot(
   phylum_long %>% filter(collection_part == "Baseline"), 
   aes(x=plate_ID, y=proportion, fill=phylum)) + 
   geom_bar(stat = "identity", position = "fill") + 
   facet_grid(cols = vars(groups2), scales = "free_x") + 
   labs(x = NULL, y = "Relative Abundance (%)", fill = "Genus") + #"Baseline", 
   theme(axis.text.x = element_blank(), 
         axis.ticks.x = element_blank(), 
         panel.background = element_blank(), 
         plot.background = element_blank(), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         strip.background = element_blank(), 
         legend.position = "none") + 
     theme(plot.margin = margin(t = 5, r = 5, b = 5, l = 25))

p2 <- ggplot(
  phylum_long %>% filter(collection_part == "Follow Up"), 
  aes(x=plate_ID, y=proportion, fill=phylum)) + 
  geom_bar(stat = "identity", position = "fill") + 
  facet_grid(cols = vars(groups2), scales = "free_x") + 
  labs(x = "Technical Replicates (n=3) per Collection Point", #NULL,#"Follow Up", 
       y = "Relative Abundance (%)", fill = "Genus") + 
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(), 
        panel.background = element_blank(), 
        plot.background = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        strip.text.x = element_blank(), 
        strip.background = element_blank(), 
        legend.position = "none") + 
  theme(plot.margin = margin(t = 5, r = 5, b = 5, l = 25))


# Remove y-axis labels from individual plots
p1_clean <- p1 + theme(
  legend.position = "none",
  axis.title.y = element_blank(),
  plot.margin = margin(t = 5, r = 10, b = 5, l = 7)
)
p2_clean <- p2 + theme(
  legend.position = "none",
  axis.title.y = element_blank(),
  plot.margin = margin(t = 5, r = 10, b = 5, l = 7)
)

# Stack the two plots vertically
plots <- plot_grid(
  p1_clean, p2_clean,
  ncol = 1,
  align = "v"
)

# Add shared right-side labels for Baseline / Follow Up
ylabels_right <- plot_grid(
  ggdraw() + draw_label("Baseline", angle = -90, x = 0.05, y = 0.5, size = 10),
  ggdraw() + draw_label("Follow Up", angle = -90, x = 0.05, y = 0.5, size = 10),
  ncol = 1
)

# Add shared y-axis label on the left
shared_ylabel <- ggdraw() + draw_label(
  "Relative Abundance (%)", angle = 90, x = 0.5, y = 0.5, size = 10
)

# Combine left y-label, stacked plots, and right labels
main <- plot_grid(
  shared_ylabel, plots, ylabels_right,
  ncol = 3,
  rel_widths = c(0.01, 1, 0.01)  # left label narrow, plots wide, right labels narrow
)

# Shared legend
legend_plot <- ggplot(phylum_long, aes(x = plate_ID, y = proportion, fill = phylum)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(fill = "Genus") +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 6),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.4, "lines")
  )
shared_legend <- get_legend(legend_plot)

# Add shared legend to the far right
rel_abund.plot <- plot_grid(
  main, shared_legend,
  ncol = 2,
  rel_widths = c(1, 0.31)
)

# Display final plot
rel_abund.plot


```

```{r Bray Curtis Plot}

## Bray Curtis ####

##Calculate Bray-Curtis Distance Matrix
tech.dist_mat <- metaphlan4_all %>% 
  #select tehcnical Controls
  filter(plate_ID %in% technical_ids) %>% 
  #select species level data
  filter(rank %in% c("species")) %>% #,"genus","subspecies")) %>%
  #pivot to wider form
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0) %>%  #covert NA relative abundance to 0
  #bray curtis prep
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray")

##Extract Within-Sample Distances
#convert distance object to long format
tech.dist_df <- as.matrix(tech.dist_mat) %>%
  as.data.frame() %>%
  rownames_to_column("sample1") %>%
  pivot_longer(-sample1, names_to = "sample2", values_to = "bray_curtis") %>%
  filter(sample1 != sample2)  # remove self-comparisons
#extract a common sample group identifier (adjust pattern as needed)
tech.dist_df <- tech.dist_df %>%
  mutate(
    sample_group1 = case_when(
      sample1 %in% c("GMBC_00307","GMBC_00412","GMBC_00408") ~ "012_P1", #plate1 012_T1-T3
      sample1 %in% c("GMBC_00617","GMBC_00911","GMBC_00807") ~ "079_P1", #plate2 079_T1-T3
      sample1 %in% c("GMBC_01204","GMBC_01216","GMBC_01120") ~ "023_P2", #plate3 023_T1-T3 (part 2)
      sample1 %in% c("GMBC_01401","GMBC_01315","GMBC_01415") ~ "012_P2", #plate4 012_T4-T6
      sample1 %in% c("GMBC_01403","GMBC_01310","GMBC_01402") ~ "079_P2", #plate4 079_T4-T6
      sample1 %in% c("GMBC_01409","GMBC_01314","GMBC_01407") ~ "023_P1", #plate4 023_T4-T6.P1 (part 1)
      TRUE ~ NA ),
    sample_group2 = case_when(
      sample2 %in% c("GMBC_00307","GMBC_00412","GMBC_00408") ~ "012_P1", #plate1 012_T1-T3
      sample2 %in% c("GMBC_00617","GMBC_00911","GMBC_00807") ~ "079_P1", #plate2 079_T1-T3
      sample2 %in% c("GMBC_01204","GMBC_01216","GMBC_01120") ~ "023_P2", #plate3 023_T1-T3 (part 2)
      sample2 %in% c("GMBC_01401","GMBC_01315","GMBC_01415") ~ "012_P2", #plate4 012_T4-T6
      sample2 %in% c("GMBC_01403","GMBC_01310","GMBC_01402") ~ "079_P2", #plate4 079_T4-T6
      sample2 %in% c("GMBC_01409","GMBC_01314","GMBC_01407") ~ "023_P1", #plate4 023_T4-T6.P1 (part 1)
      TRUE ~ NA )#
  ) %>% 
  filter(sample_group1 == sample_group2)
#summarize within each technical replicate group
tech.dist_df <- tech.dist_df %>%
  filter(sample1 < sample2) %>%
  group_by(sample_group1) %>%
  summarise(
    mean_bc = mean(bray_curtis),
    sd_bc = sd(bray_curtis),
    max_bc = max(bray_curtis),
    n_pairs = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_bc)) %>%
  mutate(label = glue("Bray-Curtis dist: {round(mean_bc, 3)} (±{round(sd_bc, 3)})"))


## Calculation PCOA

# Calculating PCOAs
beta_pcoa <- cmdscale(tech.dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  #creating grouping var
  mutate(
    groups = case_when(
      plate_ID %in% c("GMBC_00307","GMBC_00412","GMBC_00408") ~ "012_P1", #plate1 012_T1-T3
      plate_ID %in% c("GMBC_00617","GMBC_00911","GMBC_00807") ~ "079_P1", #plate2 079_T1-T3
      plate_ID %in% c("GMBC_01204","GMBC_01216","GMBC_01120") ~ "023_P2", #plate3 023_T1-T3 (part 2)
      plate_ID %in% c("GMBC_01401","GMBC_01315","GMBC_01415") ~ "012_P2", #plate4 012_T4-T6
      plate_ID %in% c("GMBC_01403","GMBC_01310","GMBC_01402") ~ "079_P2", #plate4 079_T4-T6
      plate_ID %in% c("GMBC_01409","GMBC_01314","GMBC_01407") ~ "023_P1", #plate4 023_T4-T6.P1 (part 1)
      TRUE ~ NA ),
    collection_part = ifelse(str_detect(groups, "P1"), 1, 2),
    groups2 = case_when (
            str_detect(groups, "012") ~ "012",
            str_detect(groups, "079") ~ "079",
            str_detect(groups, "023") ~ "023"),
    groups2 = factor(groups2, levels = c("012","023","079"),
                     labels = c("Participant 12","Participant 23","Participant 79")),
    collection_part = factor(collection_part, levels = c("1","2"),
                             labels = c("Baseline", "Follow Up"))
  )



# Join BC distance labels to centroid positions
centroid_labels <- beta_pcoa.df %>%
  group_by(groups = groups) %>%  # Adjust if your grouping var is named differently
  summarize(
    mean_PC1 = mean(PC1, na.rm = TRUE),
    mean_PC2 = mean(PC2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(tech.dist_df, by = c("groups" = "sample_group1")) %>% 
  mutate(
    mean_PC1 = case_when(
      groups=="012_P1" ~ -0.18,
      groups=="012_P2" ~ -0.28,
      groups=="079_P1" ~  -0.03,
      groups=="079_P2" ~  0.00,
      groups=="023_P1" ~ -0.09,
      groups=="023_P2" ~ -0.19,
      TRUE ~ mean_PC1
    ),
    mean_PC2 = case_when(
      groups=="012_P1" ~ 0.24,
      groups=="012_P2" ~ 0.285,
      groups=="079_P1" ~ 0.05,
      groups=="079_P2" ~ -0.01,
      groups=="023_P1" ~ -0.32,
      groups=="023_P2" ~ -0.285,
      TRUE ~ mean_PC2
    )
  )


## Plotting PCOAs
beta_plot <- beta_pcoa.df %>% 
  filter(!is.na(plate_ID)) %>% 
  ggplot(aes(x = PC1, y = PC2, color = groups2, shape = collection_part)) + 
  geom_point(size=4) +
  labs(x = glue("PCoA 1"),
       y = glue("PCoA 2"),
       color = "Participant",
       shape = "Collection Point") + 
  theme(#legend.position = "none",
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines"),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.position = c(0.2, 0.5),    # inside top right (x,y in npc)
        #legend.box = "horizontal",
        #legend.justification = c("left", "top"),
        legend.background = element_rect(color = "black", fill = "white"),
        legend.key.size = unit(0.4, "lines"),
        #legend.spacing.y = unit(0.1, "cm"), 
        #legend.spacing.x = unit(0.1, "cm"), 
       ) +
       guides(
         color = guide_legend(order=1),
         shape = guide_legend(order=2)
         ) # make it compact


final.beta_plot <- beta_plot +
    geom_text(
    data = centroid_labels,
    aes(x = mean_PC1, y = mean_PC2, label = label),
    inherit.aes = FALSE,  # 👈 prevents using aes from main ggplot
    color = "black", size = 3, hjust = 0, vjust = -0.5,
    show.legend = FALSE
  ) 



```
 
```{r quality figs - combined plot }

library(cowplot)

png("metaphlan_quality_2025-08-26.png", 
    width = 9 * 600, height = 3.5 * 600, 
    res = 600, bg = "white")

print(
  # Arrange side-by-side with custom widths
  plot_grid(
    final.beta_plot,
    rel_abund.plot,
    nrow = 1,
    rel_widths = c(1, 1.31),
    labels = c("A", "B"),           # built-in labels
    label_size = 15,
    label_fontface = "bold",
    label_x = 0.01,                 # fine-tune horizontal offset
    label_y = 0.99                  # fine-tune vertical offset
  )
)

dev.off()




```


## Summary
  
Taxonomic Abundance (MetaPhlAn4)
```{r metaphlan summary and estrobolome targets}


## PRE-Filtering Unique Species and Prevalent Phylums

#overall (pre filtering) - after removal of controls
metaphlan4 %>% filter(rank == "genus") %>% distinct(name) %>% nrow()
metaphlan4 %>% filter(rank == "species") %>% distinct(name) %>% nrow()
metaphlan4 %>% filter(rank == "subspecies") %>% distinct(name) %>% nrow()

#kingdom level data - summing relative abundances for each kingdom across all samples
metaphlan4 %>%
  filter(rank == "phylum") %>% 
  group_by(name) %>%
  summarise(total_abundance = sum(relative_abundance, na.rm = TRUE)) %>%
  mutate(proportion = total_abundance / sum(total_abundance))


## POST-Filtering Unique and Prevalent Species (Robertson 2023)

uni <- metaphlan4_long %>% filter(rank=="genus") %>% distinct(name)
glue("The number of unique genera in P1 samples after filtering: {nrow(uni)}")
uni.1 <- uni %>% distinct() %>% filter(grepl("GGB",name))
glue("Of these, {nrow(uni.1)} included GGB genera ({round(nrow(uni.1)/nrow(uni),3)*100}%)")

uni.species <- metaphlan4_long %>% filter(rank == "species") %>% distinct(name)
glue("The number of unique species in P1 samples after filtering: {nrow(uni.species)}")
uni.species1 <- uni.species %>% distinct() %>% filter(grepl("GGB",name))
glue("Of these, {nrow(uni.species1)} included GGB_SGB species ({round(nrow(uni.species1)/nrow(uni.species),3)*100}%)")

metaphlan4_long %>% filter(rank=="subspecies") %>% distinct(name) %>%  nrow()

#kingdom level data - summing relative abundances for each kingdom across all samples
metaphlan4_long %>%
  filter(rank == "phylum") %>% 
  group_by(name) %>%
  summarise(total_abundance = sum(relative_abundance, na.rm = TRUE)) %>%
  mutate(proportion = total_abundance / sum(total_abundance))

#F/B Ratio by case_con status
metaphlan4_long %>%
  filter(rank == "phylum") %>%
  merge(gmbc_data %>% select(plate_ID, case_con), by = "plate_ID", all.x = TRUE) %>%
  group_by(case_con, name) %>%
  summarise(total_abundance = sum(relative_abundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = name, values_from = total_abundance) %>%
  mutate(FB_ratio = `p__Firmicutes` / `p__Bacteroidetes`) %>% 
  t() %>% 
  view()


#predominant species - sum relabun for each species then sort desc
metaphlan4_wide.s %>% 
  select(!plate_ID) %>% 
  colSums() %>% 
  sort(decreasing = TRUE) %>% 
  head(5)


### RUN THE SUMMARY TABLE
describe_data(metaphlan4_wide, "plate_ID","metaphlan")

### Estrobolome Target Coverage

glue("The number of genera-level estrobolome targets: {nrow(estrobolome_targets %>% filter(rank=='genus'))}")
glue("The number of species-level estrobolome targets: {nrow(estrobolome_targets %>% filter(rank=='species'))}")

#PRIOR to data filtering
estrobolome_prefilt <- metaphlan4 %>%  
  filter(name %in% estrobolome_targets$metaphlan_GMBC) %>%
  distinct(name)

#found
glue("Genera targets found prior to filtering: {nrow(estrobolome_prefilt %>% filter(grepl('g__', name)))}")
glue("Species targets found prior to filtering: {nrow(estrobolome_prefilt %>% filter(grepl('s__', name)))}")
#not found in metaphlan
estrobolome_targets %>% filter(rank=="genus") %>% filter(!metaphlan_GMBC %in% estrobolome_prefilt$name) %>% pull(name)
estrobolome_targets %>% filter(rank=="species") %>% filter(!metaphlan_GMBC %in% estrobolome_prefilt$name) %>% pull(name)


#POST data filtering
estrobolome_postfilt <- metaphlan4_long %>%
  filter(name %in% estrobolome_targets$metaphlan_GMBC) %>%
  distinct(name)

#remaining
glue("Genera targets remaining after filtering: {nrow(estrobolome_postfilt %>% filter(grepl('g__', name)))}")
glue("Species targets remaining after filtering): {nrow(estrobolome_postfilt %>% filter(grepl('s__', name)))}")
#removed
estrobolome_prefilt %>% filter(grepl('g__', name)) %>% filter(!name %in% estrobolome_postfilt$name) %>% pull(name)
estrobolome_prefilt %>% filter(grepl('s__', name)) %>% filter(!name %in% estrobolome_postfilt$name)%>% pull(name)


```

Pathway Abundance (HUMAnN3)
```{r humann.path summary}

## RAW DATA DESCRIPTION
glue("The max relabun raw data: {max(humann.path_nopipe$relative_abundance)} (unintegrated)")
glue("The min relabun raw data: {min(humann.path_nopipe$relative_abundance)}")
glue("The max relabun raw data (removing unmapped, unintegrated): {
     humann.path_nopipe %>% filter(!name %in% c('UNINTEGRATED','UNMAPPED')) %>% pull(relative_abundance) %>% max()}")
glue("The avg relabun for unintegrated data (raw): {
     mean(humann.path_nopipe %>% filter(name == 'UNINTEGRATED') %>% pull(relative_abundance)) }")
glue("The avg relabun for unmapped data (raw): {
     mean(humann.path_nopipe %>% filter(name == 'UNMAPPED') %>% pull(relative_abundance)) }")
glue("The number of unique pathways in data (raw): {nrow( humann.path_nopipe %>% distinct(name) ) }")


## FILTERED DATA
glue("The max relabun filtered data: {max(humann.path_long$relative_abundance)}")
glue("The min relabun filtered data: {min(humann.path_long$relative_abundance)}")
glue("The number of unique pathways in filtered data: {nrow( humann.path_long %>% distinct(name) ) }")

#predominant pathways - sum relabun for each path then sort desc
humann.path_wide %>% 
  select(!plate_ID) %>% 
  colSums() %>% 
  sort(decreasing = TRUE) %>% 
  head(5)


### RUN THE SUMMARY TABLE
describe_data(humann.path_wide, "plate_ID","humann3_path")

## Estrobolome Targets

#overall number
glue("The number of pathway estrobolome targets: {nrow(estrobolome_targets %>% filter(!is.na(Pathway_code)))}")

#found in raw data
estrobolome.path_prefilt <- humann.path_nopipe %>%
  filter(sapply(name, function(x) any(str_detect(x, estrobolome_targets$Pathway_code)))) %>%
  distinct(name) %>% 
  filter(!grepl("GALACT", name))
estrobolome.path_prefilt$name

#found in filtered p1 data
estrobolome.path_postfilt <- humann.path_long %>%
  filter(sapply(name, function(x) any(str_detect(x, estrobolome_targets$Pathway_code)))) %>%
  distinct(name) %>% 
  filter(!grepl("GALACT", name))
estrobolome.path_postfilt$name

#those not found
estrobolome_targets %>% 
    filter(!sapply(Pathway_code, function(x) any(str_detect(x, estrobolome.path_prefilt$name)))) %>%
    distinct(Pathway_code, Pathway_name)

``` 

Enzyme Abundance (HUMAnN3)
```{r humann.enzymes summary}

## RAW DATA DESCRIPTION

glue("The max relabun raw data: {max(humann.enzymes_nopipe$relative_abundance)}")
glue("The min relabun raw data: {min(humann.enzymes_nopipe$relative_abundance)}")
glue("The number of unique enzymes in raw samples: {nrow( humann.enzymes_nopipe %>% distinct(name) ) }")

## FILTERED DATA
glue("The max relabun filtered data: {max(humann.enzymes_long$relative_abundance)}")
glue("The min relabun filtered data: {min(humann.enzymes_long$relative_abundance)}")
glue("The number of unique enzymes in filtered data: {nrow(humann.enzymes_long %>% distinct(name))}")

#predominant enzymes - sum relabun for each path then sort desc
humann.enzymes_wide %>% 
  select(!plate_ID) %>% 
  colSums() %>% 
  sort(decreasing = TRUE) %>% 
  head(5) %>% 
  as.data.frame() %>% 
  rownames_to_column("name") %>% 
  merge(., humann.enzymes_nopipe %>% select(name, Enzyme_name),
        by = "name") %>% 
  unique() %>% 
  mutate(names = paste0(Enzyme_name," (", name, ")")) %>% 
  select(-name, -Enzyme_name)

### RUN THE SUMMARY TABLE
describe_data(humann.enzymes_wide, "plate_ID","humann3_enzymes")

## Estrobolome Targets

#overall number
glue("The number of enzyme estrobolome targets: {nrow(estrobolome_targets %>% filter(!is.na(Enzyme_code)))}")

#found in raw p1 data
estrobolome.enzyme_prefilt <- humann.enzymes_nopipe %>%
  filter(name %in% estrobolome_targets$Enzyme_code) %>%
  distinct(name, Enzyme_name)
glue("The number of estrobolome enzyme targets found in baseline raw data: {nrow(estrobolome.enzyme_prefilt)}")
#those found
estrobolome.enzyme_prefilt %>% distinct(Enzyme_name,name) %>% pull(Enzyme_name, name)
#those not found
estrobolome_targets %>% filter(!Enzyme_code %in% estrobolome.enzyme_prefilt$name) %>% distinct(Enzyme_name) %>% pull(Enzyme_name)


#found in filtered p1 data
estrobolome.enzyme_postfilt <- humann.enzymes_long %>%
  filter(name %in% estrobolome_targets$Enzyme_code) %>%
  distinct(name)
glue("The number of estrobolome enzyme targets found in baseline filtered data: {nrow(estrobolome.enzyme_postfilt)}")
#not found
setdiff(estrobolome.enzyme_prefilt$name, estrobolome.enzyme_postfilt$name)

```

 
## Analyses

Alpha and Beta Diversity 
```{r alpha and beta diversity - apply metaphlan}

#run alpha diversity
alpha_results <- run_alpha_diversity(metaphlan4_long, metaphlan=TRUE)

#plot alpha and beta diversity together
final_plot <- plot_alpha_beta(
  alpha_results$alpha_data, 
  #run beta diversity and get plot
  beta_plot = plot_beta_diversity(metaphlan4_wide.t, -0.43, -0.35),
  #plot alpha diversity - shannon
  alpha_plot1 = plot_alpha_diversity(alpha_results$alpha_data, shannon, "Shannon Index",
                                     alpha_results$p_shannon, 1.5, 2.5),
  #plot alpha diversity - richness
  alpha_plot2 = plot_alpha_diversity(alpha_results$alpha_data, richness, "Species Richness",
                                     alpha_results$p_richness, 1.5, 60),
  #"A","B","C",
  "","","",
  "alpha-beta_metaphlan_2025-08-18.png"
)


```

```{r alpha and beta diversity - apply humann.path}

#run alpha diversity
alpha_results <- run_alpha_diversity(humann.path_long)

#plot alpha and beta diversity together
final_plot.path <- plot_alpha_beta(
  alpha_results$alpha_data, 
  #run beta diversity and get plot
  beta_plot = plot_beta_diversity(humann.path_wide, 0.1, 0.11),
  #plot alpha diversity - shannon
  alpha_plot1 = plot_alpha_diversity(alpha_results$alpha_data, shannon, "Shannon Index",
                                     alpha_results$p_shannon, 1.5, 5.55),
  #plot alpha diversity - richness
  alpha_plot2 = plot_alpha_diversity(alpha_results$alpha_data, richness, "Pathway Richness",
                                     alpha_results$p_richness, 1.5, 410),
  #"A","B","C",
  "","","",
  "alpha-beta_humann.path_2025-08-18.png"
)

```

```{r alpha and beta diversity - apply humann.enzymes}

#run alpha diversity
alpha_results <- run_alpha_diversity(humann.enzymes_long)

#plot alpha and beta diversity together
final_plot.enzymes <- plot_alpha_beta(
  alpha_results$alpha_data, 
  #run beta diversity and get plot
  beta_plot = plot_beta_diversity(humann.enzymes_wide, 0.07, 0.16),
  #plot alpha diversity - shannon
  alpha_plot1 = plot_alpha_diversity(alpha_results$alpha_data, shannon, "Shannon Index",
                                     alpha_results$p_shannon, 1.5, 6.59),
  #plot alpha diversity - richness
  alpha_plot2 = plot_alpha_diversity(alpha_results$alpha_data, richness, "Enzyme Richness",
                                     alpha_results$p_richness, 1.5, 1850),
  #"D","E","F",
  "","","",
  "alpha-beta_humann.enzymes_2025-08-18.png"
)

```

```{r alpha/beta combined path + enzymes plot }

library(magick)

# Read the two PNG files
img0 <- image_read("alpha-beta_metaphlan_2025-08-18.png")
img1 <- image_read("alpha-beta_humann.path_2025-08-18.png")
img2 <- image_read("alpha-beta_humann.enzymes_2025-08-18.png")

# Combine side by side
combined <- image_append(c(img0, img1, img2), stack=TRUE)  # horizontal by default

# Save the combined image
image_write(combined, "combined_plot.png")




```

```{r Combined plot}

library(cowplot)

png("diversity_2025-11-13.png", 
    width = 6.5 * 600, height = 6.5 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #row 1, metaphlan
    plot_grid(
      plot_grid(metaphlan.shannon, metaphlan.richness, nrow = 2, rel_heights = c(1, 1)),
      metaphlan.beta_plot,
      nrow = 1,
      ncol = 2,
      rel_widths = c(1,1.5)
    ),
    #row 2, pathways
    plot_grid(
      plot_grid(path.shannon, path.richness, nrow = 2, rel_heights = c(1, 1)),
      path.beta_plot,
      nrow = 1,
      ncol = 2,
      rel_widths = c(1,1.5)
    ),
    #arguments
    nrow = 2,
    ncol = 2,
    rel_heights = c(1,1)
  )
)



grid.text("Species", x = unit(0.45, "npc"), y = unit(0.985, "npc"), #0.099 x
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Pathways", x = unit(0.45, "npc"), y = unit(0.675, "npc"), #0.348
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Enzymes", x = unit(0.45, "npc"), y = unit(0.342, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Cases", x = unit(0.15, "npc"), y = unit(0.993, "npc"), #0.099 x
          gp = gpar(fontsize = 9))

grid.text("Controls", x = unit(0.305, "npc"), y = unit(0.993, "npc"), #0.099 x
          gp = gpar(fontsize = 9))


dev.off()


```


Estrobolome - Regression Modelling
```{r LogReg - estrobolome targets all}

## Combine all estrobolome targets datasets
#target list
estrobolome_list <- c(estrobolome_targets$metaphlan_GMBC,
                      estrobolome_targets$Enzyme_code,
                      estrobolome_targets$Pathway_code)
estrobolome_list <- Filter(Negate(is.na), estrobolome_list)
#select from each datatype
estro.t <- metaphlan4_wide.t %>% 
  transform_metagenomics(merge=FALSE) %>% 
  select(plate_ID, matches(paste(estrobolome_list, collapse = "|")))
estro.g <- metaphlan4_wide.g %>% 
  transform_metagenomics(merge=FALSE) %>% 
  select(plate_ID, matches(paste(estrobolome_list, collapse = "|")))
estro.p <- humann.path_wide %>% 
  transform_metagenomics(merge=FALSE) %>% 
  select(plate_ID, matches(paste(estrobolome_list, collapse = "|"))) %>% 
  select(!contains("GALACT"))
estro.e <- humann.enzymes_wide %>% 
  transform_metagenomics() %>% 
  select(plate_ID, matches(paste(estrobolome_list, collapse = "|")),
         case_con, "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
    "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
    "family_history","health_condition","rh_pregnancy_live","rh_pregnancy_age_first_cat",
    "rh_breastfeed_months","rh_contraceptive","hormone_ever",
    "Total.dietary.fibre.g_density",
    "Total.saturated.fatty.acids.g_density",
    "Total.sugar.g_density" )
#merge all together
estro.full <- estro.t %>% 
  left_join(estro.g, by="plate_ID") %>% 
  left_join(estro.p, by="plate_ID") %>% 
  left_join(estro.e, by="plate_ID")


#run: iterative modeling - apply
apply_iterative.modeling(
  predictor_vars <- estro.full %>% select(2:72) %>% colnames(), 
  adjustment_sets <- get_covariates(),
  dataset.model <- estro.full,
  file_name <- "estrobolome.targets_clr"
)

```

```{r LogReg unadjusted - estrobolome targets all volcano plot}

library(readxl)

## UNADJUSTED
results <- read_excel("Logreg_estrobolome.targets_clr_2025-09-05.xlsx", sheet = "Sheet_unadj") %>%
  mutate(
    sig = ifelse(q.value < 0.25, "Yes (q<0.25)", "No (q≥0.25)"), # & p.value < 0.05
    log10_FDR = -log10(q.value)
  )

## SCATTER WITH BAR PLOT

# Step 1: Generate histogram data from the OR column
hist_data <- ggplot_build(
  ggplot(results, aes(x = OR)) +
    geom_histogram(bins = 100)
)$data[[1]]

# Step 2: Rescale histogram y values to negative
hist_data$y <- -hist_data$count / max(hist_data$count) / 8  # adjust height if needed

# Step 3: Final plot
volcano.e.0 <- ggplot() +
  # Histogram bars below y = 0
  geom_col(data = hist_data, aes(x = x, y = y), fill = "black", width = hist_data$width[1]) +
  
  # Scatter points
  geom_point(data = results, aes(x = OR, y = log10_FDR, shape = shape, color = sig),
             alpha = 0.7, size = 2) +

  # Scales and lines
  scale_color_manual(values = c("Yes (q<0.25)" = "red", "No (q≥0.25)" = "grey70")) +
  scale_shape_manual(values = c("Species" = 16,
                                 "Genus" = 17,
                                 "Pathway" = 4,
                                 "Enzyme" = 15)) +
  #geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_segment(aes(x = 1, xend = 1, y = 0, yend=1.26),
             linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha=1) +

  # Adjust y-axis to make room for bars
  coord_cartesian(ylim = c(-0.07, NA)) +
  scale_x_continuous(limits = c(0.75, 1.25)) +
  #scale_x_continuous(limits = c(NA, 1.5)) +
  labs(
    x = "Odds Ratio",
    y = "-log10(FDR)",
    color = "Significance",
    shape = "Group"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    legend.position = "none"
  )

volcano.e.0
```
  
```{r LogReg adjusted - estrobolome targets all volcano plot}

library(readxl)

## CONFOUNDER
results <-  read_excel("Logreg_estrobolome.targets_clr_2025-09-05.xlsx", sheet = "Sheet_confounder_vars") %>%
  mutate(
    sig = ifelse(q.value < 0.25, "Yes (q<0.25)", "No (q≥0.25)"),
    log10_FDR = -log10(q.value)
  ) 

#duplicate row 1, give big OR so outside of range, make POS for the LEGEND
results <- rbind(results, results[1, ])
results[nrow(results), "OR"] <- 5
results[nrow(results), "sig"] <- "Yes (q<0.25)"


## SCATTER WITH BAR PLOT

# Step 1: Generate histogram data from the OR column
hist_data <- ggplot_build(
  ggplot(results, aes(x = OR)) +
    geom_histogram(bins = 250)
)$data[[1]]

# Step 2: Rescale histogram y values to negative
hist_data$y <- -hist_data$count / max(hist_data$count) / 8  # adjust height if needed

# Step 3: Final plot
volcano.e.1 <- ggplot() +
  # Histogram bars below y = 0
  geom_col(data = hist_data, aes(x = x, y = y), fill = "black", width = hist_data$width[1]) +
  
  # Scatter points
  geom_point(data = results, aes(x = OR, y = log10_FDR, shape = shape, color = sig),
             alpha = 0.7, size = 2) +

  # Scales and lines
  scale_color_manual(values = c("Yes (q<0.25)" = "red", "No (q≥0.25)" = "grey70")) +
  scale_shape_manual(values = c("Species" = 16,
                                 "Genus" = 17,
                                 "Pathway" = 4,
                                 "Enzyme" = 15)) +
  #geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_segment(aes(x = 1, xend = 1, y = 0, yend=1.26),
             linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha=1) +

  # Adjust y-axis to make room for bars
  coord_cartesian(ylim = c(-0.07, NA)) +
  scale_x_continuous(limits = c(0.75, 1.25)) +
  #scale_x_continuous(limits = c(NA, 1.5)) +
  labs(
    x = "Odds Ratio",
    y = NULL,#"-log10(FDR)",
    color = "Significant",
    shape = "Feature"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    legend.text = element_text(size = 6.8),
    legend.title = element_text(size = 8),
    legend.position = c(1, 0.98),    # inside top right (x,y in npc)
    legend.box = "horizontal",
    legend.justification = c("right", "top"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.7, "lines"),
    #legend.spacing.y = unit(0.1, "cm"), 
    #legend.spacing.x = unit(0.1, "cm"), 
   ) +
   guides(
     color = guide_legend(order=1),
     shape = guide_legend(ncol = 2, byrow = TRUE, order=2)
     ) # make it compact

volcano.e.1
```

```{r LogReg - estrobolome combine volcano plot }

library(cowplot)

png("volcano_estrobolome_2025-09-05.png", 
    width = 6.5 * 600, height = 3.25 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    volcano.e.0, volcano.e.1,
    nrow = 1,
    rel_widths = c(1, 1)
  )
)

grid.text("A", x = unit(0.08, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("B", x = unit(0.55, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

dev.off()


```


XGBoost Top Features - Regression Modelling
```{r LogReg - XGBoost subset}

## Combine selected features and datasets
#target list
estrobolome_list <- c(
  "g__Blautia",
  "g__Alistipes",
  "g__Hungatella",
  "g__Candidatus_Avimonas",
  "g__GGB3057",
  "g__Faecalicatena",
  "g__GGB79634",
  "g__Christensenellaceae_unclassified",
  "g__GGB13404",
  "g__Intestinimonas",
  "g__GGB79734",
  "g__GGB9063",
  "g__Lachnoclostridium",
  "g__GGB9635",
  "g__GGB80140",
  "s__Anaerotruncus_rubiinfantis|t__SGB25416",
  "s__Alistipes_shahii|t__SGB2295",
  "s__Lachnospiraceae_bacterium_AM48_27BH|t__SGB4706",
  "s__Alistipes_finegoldii|t__SGB2301",
  "s__Blautia_obeum|t__SGB4811",
  "s__Sutterella_wadsworthensis|t__SGB9286",
  "s__Alistipes_putredinis|t__SGB2318",
  "s__Mediterraneibacter_massiliensis|t__SGB4595",
  "s__Candidatus_Avimonas_narfia|t__SGB14941",
  "s__Mediterraneibacter_butyricigenes|t__SGB25493",
  "s__Blautia_SGB4815|t__SGB4815",
  "s__Faecalibacillus_intestinalis|t__SGB6754",
  "s__Clostridium_fessum|t__SGB4705",
  "s__Candidatus_Cibionibacter_quicibialis|t__SGB15286",
  "s__Anaerobutyricum_hallii|t__SGB4532",
  "s__GGB3057_SGB4059|t__SGB4059",
  "PWY-7992: superpathway of menaquinol-8 biosynthesis III",
  "PWY-7992: superpathway of menaquinol-8 biosynthesis III",
  "RIBOSYN2-PWY: flavin biosynthesis I (bacteria and plants)",
  "PWY-7883: anhydromuropeptides recycling II",
  "PWY-5130: 2-oxobutanoate degradation I",
  "PWY-7761: NAD salvage pathway II (PNC IV cycle)",
  "PWY-6876: isopropanol biosynthesis (engineered)",
  "PWY-6901: superpathway of glucose and xylose degradation",
  "P42-PWY: incomplete reductive TCA cycle",
  "PWY-8131: 5'-deoxyadenosine degradation II",
  "METHGLYUT-PWY: superpathway of methylglyoxal degradation",
  "PWY-7013: (S)-propane-1,2-diol degradation",
  "PWY-6588: pyruvate fermentation to acetone",
  "PWY-7371: 1,4-dihydroxy-6-naphthoate biosynthesis II",
  "PWY-5973: cis-vaccenate biosynthesis",
  "PWY-5920: superpathway of heme b biosynthesis from glycine",
  "1.8.98.1",
  "3.5.1.81",
  "2.7.1.113",
  "2.5.1.120",
  "3.4.13.21",
  "1.2.99.5",
  "6.3.4.15",
  "1.4.99.6",
  "3.4.22.8",
  "4.2.1.115",
  "2.4.1.319",
  "2.4.1.329",
  "2.1.1.173",
  "1.10.9.1",
  "5.1.3.8",
  "2.1.1.61"
)

#select from each datatype
estro.t <- metaphlan4_wide.g %>% #t
  transform_metagenomics() %>% #merge=FALSE
  select(plate_ID, all_of(intersect(estrobolome_list, colnames(.)))) 
estro.p <- humann.path_wide %>%
  transform_metagenomics(merge=FALSE) %>%
  select(plate_ID, all_of(intersect(estrobolome_list, colnames(.))))
estro.e <- humann.enzymes_wide %>%
  transform_metagenomics() %>%
  select(plate_ID, all_of(intersect(estrobolome_list, colnames(.))),
         case_con, "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
    "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
    "Total.dietary.fibre.g_density",
    "Total.saturated.fatty.acids.g_density",
    "Total.sugar.g_density"
)
#merge all together
estro.full <- estro.t %>%
  left_join(estro.p, by="plate_ID") %>%
  left_join(estro.e, by="plate_ID")


#run: iterative modeling - apply
apply_iterative.modeling(
  predictor_vars <- estrobolome_list, 
  adjustment_sets <- get_covariates(),
  dataset.model <- estro.full,
  file_name <- "top.features_clr"
)

```


XGBoost Top Features - Relative Abundance Plots       
```{r genus - relabun plot}

## top 15 features list
features <- c(
  "g__Blautia",
  "g__Alistipes",
  "g__Hungatella",
  "g__Candidatus_Avimonas",
  "g__GGB3057",
  "g__Faecalicatena",
  "g__GGB79634",
  "g__Christensenellaceae_unclassified",
  "g__GGB13404",
  "g__Intestinimonas",
  "g__GGB79734",
  "g__GGB9063",
  "g__Lachnoclostridium",
  "g__GGB9635",
  "g__GGB80140"
)

## Metaphlan data
maaslin2_data.plot <- metaphlan4_long %>% 
  filter(name %in% features) %>% 
  select(plate_ID, name, relative_abundance) %>% 
  merge(., gmbc_data %>% select(plate_ID, single_id, collection_part, case_con)) %>% 
  mutate(case_con = factor(case_con, 
                           levels = c(0, 1), 
                           labels = c('Control', 'Case'))) %>% 
  #clean up names
  mutate(
    name = name %>%
      str_remove("^g__") %>%         # remove leading s__
      str_replace_all("_", " ")       # replace underscores with spaces
  ) %>% 
  mutate(name = ifelse(name == "Christensenellaceae unclassified", 
                       "Christensenell. uncl.", name)) %>% 
  #filter out zero vals
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance + 1e-6))
  

## Clean up features list the same way as in the data
features_clean <- features %>%
  str_remove("^g__") %>%
  str_replace_all("_", " ") %>%
  ifelse(. == "Christensenellaceae unclassified", "Christensenell. uncl.", .)

## Apply the factor ordering
maaslin2_data.plot <- maaslin2_data.plot %>%
  mutate(name = factor(name, levels = rev(features_clean)))  # rev() = top-to-bottom order


## Plotting
rel_abun.genus <- ggplot(maaslin2_data.plot, aes(y=name, x=log_rel_abundance, color=case_con)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.12) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(name, case_con), fill = case_con),
               position = position_dodge(width = 0.8),
               alpha = 0.4,
               outlier.shape = NA) +
  labs(y=NULL, x="log10(Relative Abundance)") +
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  scale_fill_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_text(), #element_markdown(),
        axis.text.x = element_text(size = 8),
        legend.position = "none"
        # legend.text = element_text(size = 6),
        # legend.position = c(0.92, 0.5),
        # legend.background = element_rect(color="black", fill = NA),
        # legend.margin = margin(t=1, r=3, b=3),
        # legend.key.size = unit(0.7, "lines"),
        )


```
  
```{r species - relabun plot}

## top 15 features list
features <- c(
  "s__Anaerotruncus_rubiinfantis|t__SGB25416",
  "s__Alistipes_shahii|t__SGB2295",
  "s__Lachnospiraceae_bacterium_AM48_27BH|t__SGB4706",
  "s__Alistipes_finegoldii|t__SGB2301",
  "s__Blautia_obeum|t__SGB4811",
  "s__Sutterella_wadsworthensis|t__SGB9286",
  "s__Alistipes_putredinis|t__SGB2318",
  "s__Mediterraneibacter_massiliensis|t__SGB4595",
  "s__Candidatus_Avimonas_narfia|t__SGB14941",
  "s__Mediterraneibacter_butyricigenes|t__SGB25493",
  "s__Blautia_SGB4815|t__SGB4815",
  "s__Faecalibacillus_intestinalis|t__SGB6754",
  "s__Clostridium_fessum|t__SGB4705",
  "s__Candidatus_Cibionibacter_quicibialis|t__SGB15286",
  "s__Anaerobutyricum_hallii|t__SGB4532"
)

## Metaphlan data
maaslin2_data.plot <- metaphlan4_long %>% 
  filter(name %in% features) %>% 
  select(plate_ID, name, relative_abundance) %>% 
  merge(., gmbc_data %>% select(plate_ID, single_id, collection_part, case_con)) %>% 
  mutate(case_con = factor(case_con, 
                           levels = c(0, 1), 
                           labels = c('Control', 'Case'))) %>% 
  #clean up names
  mutate(
    name = name %>%
      str_remove("^s__") %>%                   # remove leading s__
      str_remove("\\|t__SGB[0-9]+") %>%        # remove "|t__SGB####"
      str_replace_all("_", " ")                # replace underscores with spaces
  ) %>% 
  mutate(name = ifelse(name == "Lachnospiraceae bacterium AM48 27BH", 
                       "L. bact. AM48-27BH", name)) %>% 
  mutate(name = ifelse(name == "Candidatus Cibionibacter quicibialis", 
                       "C. C. quicibialis", name)) %>% 
  mutate(name = ifelse(name == "Candidatus Avimonas narfia", 
                       "C. Avimonas narfia", name)) %>% 
  mutate(name = ifelse(name == "Mediterraneibacter butyricigenes", 
                       "M. butyricigenes", name)) %>% 
  mutate(name = ifelse(name == "Mediterraneibacter massiliensis", 
                       "M. massiliensis", name)) %>% 
  mutate(name = ifelse(name == "Anaerotruncus rubiinfantis", 
                       "A. rubiinfantis", name)) %>% 
  mutate(name = ifelse(name == "Sutterella wadsworthensis", 
                       "S. wadsworthensis", name)) %>% 
  mutate(name = ifelse(name == "Faecalibacillus intestinalis", 
                       "F. intestinalis", name)) %>% 
  mutate(name = ifelse(name == "Anaerobutyricum hallii",
                       "A. hallii", name)) %>%
  #filter out zero vals
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance + 1e-6))


## Clean up features list the same way as in the data
features_clean <- features %>%
  str_remove("^s__") %>%
  str_remove("\\|t__SGB[0-9]+") %>%
  str_replace_all("_", " ") %>%
  ifelse(. == "Lachnospiraceae bacterium AM48 27BH", "L. bact. AM48-27BH", .) %>%
  ifelse(. == "Candidatus Cibionibacter quicibialis", "C. C. quicibialis", .) %>%
  ifelse(. == "Candidatus Avimonas narfia", "C. Avimonas narfia", .) %>%
  ifelse(. == "Mediterraneibacter butyricigenes", "M. butyricigenes", .) %>%
  ifelse(. == "Mediterraneibacter massiliensis", "M. massiliensis", .) %>% 
  ifelse(. == "Anaerotruncus rubiinfantis", "A. rubiinfantis", .) %>% 
  ifelse(. == "Sutterella wadsworthensis", "S. wadsworthensis", .) %>% 
  ifelse(. == "Faecalibacillus intestinalis", "F. intestinalis", .) %>%
  ifelse(. == "Anaerobutyricum hallii", "A. hallii", .)

## Apply factor levels to enforce order top-to-bottom
maaslin2_data.plot <- maaslin2_data.plot %>%
  mutate(name = factor(name, levels = rev(features_clean)))

## Plotting
rel_abun.species <- ggplot(maaslin2_data.plot, aes(y=name, x=log_rel_abundance, color=case_con)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.12) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(name, case_con), fill = case_con),
               position = position_dodge(width = 0.8),
               alpha = 0.4,
               outlier.shape = NA) +
  labs(y=NULL, x="log10(Relative Abundance)") +
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  scale_fill_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_markdown(),
        axis.text.x = element_text(size = 8),
        legend.position = "none"
        #legend.background = element_rect(color="black", fill = NA),
        #legend.margin = margin(t=1, r=3, b=3)
        )


```

```{r pathways - relabun plot} 

## top 15 features list
features <- c(
  "PWY-7992: superpathway of menaquinol-8 biosynthesis III",
  "RIBOSYN2-PWY: flavin biosynthesis I (bacteria and plants)",
  "PWY-7883: anhydromuropeptides recycling II",
  "PWY-5130: 2-oxobutanoate degradation I",
  "PWY-7761: NAD salvage pathway II (PNC IV cycle)",
  "PWY-6876: isopropanol biosynthesis (engineered)",
  "PWY-6901: superpathway of glucose and xylose degradation",
  "P42-PWY: incomplete reductive TCA cycle",
  "PWY-8131: 5'-deoxyadenosine degradation II",
  "METHGLYUT-PWY: superpathway of methylglyoxal degradation",
  "PWY-7013: (S)-propane-1,2-diol degradation",
  "PWY-6588: pyruvate fermentation to acetone",
  "PWY-7371: 1,4-dihydroxy-6-naphthoate biosynthesis II",
  "PWY-5973: cis-vaccenate biosynthesis",
  "PWY-5920: superpathway of heme b biosynthesis from glycine"
)


## Metaphlan data
maaslin2_data.plot <- humann.path_long %>% 
  filter(name %in% features) %>% 
  select(plate_ID, name, relative_abundance) %>% 
  merge(., gmbc_data %>% select(plate_ID, single_id, collection_part, case_con)) %>% 
  mutate(case_con = factor(case_con, 
                           levels = c(0, 1), 
                           labels = c('Control', 'Case'))) %>% 
  #clean up names
  mutate(
    name = name %>%
      str_remove(":.*$")   # remove everything from ":" to end
  ) %>% 
  mutate(
    name = case_when(
      name == "PWY-7992" ~ "menaquinol-8 B (7992)",
      name == "RIBOSYN2-PWY" ~ "flavin B (ribosyn2)",
      name == "PWY-7883" ~ "murein recycling (7883)",
      name == "PWY-5130" ~ "2-oxobutanoate D (5130)",
      name == "PWY-7761" ~ "NAD salvage (7761)",
      name == "PWY-6876" ~ "isopropanol B (6876)",
      name == "PWY-6901" ~ "glucose D (6901)",
      name == "P42-PWY" ~ "reduct TCA cyc (P42)",
      name == "PWY-8131" ~ "5'-deoxyA D (8131)",
      name == "METHGLYUT-PWY" ~ "methylglyoxal D",
      name == "PWY-7013" ~ "propane-1,2-diol D (7013)",
      name == "PWY-6588" ~ "pyruvate ferm (6588)",
      name == "PWY-7371" ~ "dihyd-naphthoate B (7371)",
      name == "PWY-5973" ~ "cis-vaccenate B (5973)",
      name == "PWY-5920" ~ "heme b B (5920)",
      TRUE ~ name
    ),
    name2 = factor(name, levels = rev(c(
      "menaquinol-8 B (7992)",
      "flavin B (ribosyn2)",
      "murein recycling (7883)",
      "2-oxobutanoate D (5130)",
      "NAD salvage (7761)",
      "isopropanol B (6876)",
      "glucose D (6901)",
      "reduct TCA cyc (P42)",
      "5'-deoxyA D (8131)",
      "methylglyoxal D",
      "propane-1,2-diol D (7013)",
      "pyruvate ferm (6588)",
      "dihyd-naphthoate B (7371)",
      "cis-vaccenate B (5973)",
      "heme b B (5920)"
    )))
  ) %>% 
  #filter out zero vals
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance + 1e-6)) 


## Plotting
rel_abun.path <- ggplot(maaslin2_data.plot, aes(y=name2, x=log_rel_abundance, color=case_con)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.1) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(name, case_con), fill = case_con),
               position = position_dodge(width = 0.8),
               alpha = 0.4,
               outlier.shape = NA) +
  labs(y=NULL, x="log10(Relative Abundance)") + #
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  scale_fill_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_text(hjust = 1),
        axis.text.x = element_text(size = 8),
        legend.position = "none"
        #legend.background = element_rect(color="black", fill = NA),
        #legend.margin = margin(t=1, r=3, b=3)
        )


```
        
```{r enzymes - relabun plot}

## top 15 features list
features <- c(
  "1.8.98.1",
  "3.5.1.81",
  "2.7.1.113",
  "2.5.1.120",
  "3.4.13.21",
  "1.2.99.5",
  "6.3.4.15",
  "1.4.99.6",
  "3.4.22.8",
  "4.2.1.115",
  "2.4.1.319",
  "2.4.1.329",
  "2.1.1.173",
  "1.10.9.1",
  "5.1.3.8"
)


## Metaphlan data
maaslin2_data.plot <- humann.enzymes_long %>% 
  filter(name %in% features) %>% 
  select(plate_ID, name, relative_abundance) %>% 
  mutate(name = case_when(
   name == "1.2.99.5" ~ "1.2.7.12",
   name == "1.10.9.1" ~ "7.1.1.6",
   TRUE ~ name
  )) %>% 
  #clean up names
  #mutate(name = paste0("EC ", name)) %>% 
  mutate(name2 = case_when(
      name == "1.8.98.1" ~ "DiMP:CoB-CoM Hdr(1.8.98.1)",
      name == "3.5.1.81" ~ "Daa deacylase (3.5.1.81)", #N-acyl-
      name == "2.7.1.113" ~ "DeoxyG kinase (2.7.1.113)",
      name == "2.5.1.120" ~ "ADF synthase (2.5.1.120)",
      name == "3.4.13.21" ~ "Dipeptidase E (3.4.13.21)",
      name == "1.2.7.12" ~ "F-formyl DH (1.2.7.12)",
      name == "6.3.4.15" ~ "B–BCCP ligase (6.3.4.15)",
      name == "1.4.99.6" ~ "D-Arg DH (1.4.99.6)",
      name == "3.4.22.8" ~ "Clostripain (3.4.22.8)",
      name == "4.2.1.115" ~ "UDP-GlcNAc DH (4.2.1.115)",
      name == "2.4.1.319" ~ "β1,4-MOP (2.4.1.319)",
      name == "2.4.1.329" ~ "S6FPase (2.4.1.329)",
      name == "2.1.1.173" ~ "23S G2445 MTase(2.1.1.173)",
      name == "7.1.1.6" ~ "PQ-PC reductase (7.1.1.6)",
      name == "5.1.3.8" ~ "GlcNAc 2-epimerase(5.1.3.8)",
    TRUE ~ name),
    name2 = factor(name2, levels = rev(c(
      "DiMP:CoB-CoM Hdr(1.8.98.1)",
      "Daa deacylase (3.5.1.81)",
      "DeoxyG kinase (2.7.1.113)",
      "ADF synthase (2.5.1.120)",
      "Dipeptidase E (3.4.13.21)",
      "F-formyl DH (1.2.7.12)",
      "B–BCCP ligase (6.3.4.15)",
      "D-Arg DH (1.4.99.6)",
      "Clostripain (3.4.22.8)",
      "UDP-GlcNAc DH (4.2.1.115)",
      "β1,4-MOP (2.4.1.319)",
      "S6FPase (2.4.1.329)",
      "23S G2445 MTase(2.1.1.173)",
      "PQ-PC reductase (7.1.1.6)",
      "GlcNAc 2-epimerase(5.1.3.8)"
    )))
  ) %>% 
  merge(., gmbc_data %>% select(plate_ID, single_id, collection_part, case_con)) %>% 
  mutate(case_con = factor(case_con, 
                           levels = c(0, 1), 
                           labels = c('Control', 'Case'))) %>% 
  #filter out zero vals
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance + 1e-6))


## Plotting
rel_abun.enzymes <- ggplot(maaslin2_data.plot, aes(y=name2, x=log_rel_abundance, color=case_con)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.1) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(name, case_con), fill = case_con),
               position = position_dodge(width = 0.8),
               alpha = 0.4,
               outlier.shape = NA) +
  labs(y=NULL, x="log10(Relative Abundance)") +
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  scale_fill_manual(name = NULL, values = c("Case" = "blue4", "Control" = "grey30")) +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_text(), #element_markdown(),
        axis.text.x = element_text(size = 8),
        legend.position = "none"
        #legend.background = element_rect(color="black", fill = NA),
        #legend.margin = margin(t=1, r=3, b=3)
        )


```
     
```{r combined plot}

library(cowplot) 

#combine bottom row first
top_row <- plot_grid(
  rel_abun.genus, 
  rel_abun.path, 
  ncol = 2, 
  rel_widths = c(1, 1.05) 
)

bottom_row <- plot_grid(
  rel_abun.species, rel_abun.enzymes,
  ncol = 2,
  rel_widths = c(1, 1.15)   # make left one wider
)


png("metagenome_relabun_2025-10-03.png", 
    width = 8 * 600, height = 9 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  #combine everything
  plot_grid(
    top_row,
    bottom_row,
    nrow = 2
  )
)


# ENZYMES
grid.text("Genera",    x = 0.1, y = 0.53, gp = gpar(fontsize = 15, fontface = "bold"))
grid.text("Pathways", x = 0.59, y = 0.53, gp = gpar(fontsize = 15, fontface = "bold"))
grid.text("Species",  x = 0.1, y = 0.03, gp = gpar(fontsize = 15, fontface = "bold"))
grid.text("Enzymes",  x = 0.59, y = 0.03, gp = gpar(fontsize = 15, fontface = "bold"))



dev.off()

```


# METABOLOMICS 

OPLS-DA
```{r OPLS-DA - plasma}

library(ropls)


## Prepare Data

#dataset
oplsda <- metabo.plasma_wide %>% 
  transform_metabo(., remove_outliers = F, merge = F) %>% 
  merge(., gmbc_data %>% select(study_id, case_con))
# predictors (metabolites)
X <- oplsda %>% select(-c(case_con,study_id)) %>% as.matrix()
# outcome variable (e.g., case vs. control)
Y <- as.factor(oplsda$case_con)  



## Run OPLS-DA
opls_model <- opls(X, Y, predI = 1, orthoI = NA, permI = 100)  
  # predI = predictive components, use 1 to separate Y groups
  # orthoI = orthogonal components, NA lets ropls select best no. via CV
  # permI = permutation testing, confirms statistical validity of the model
  # R2Y = goodness of fit 
  # Q2 = cross-validated predictability


## Important Variables

vipVn.plasma <- getVipVn(opls_model) %>%   # Variable Importance in Projection (VIP) scores
  as.data.frame() %>% 
  arrange(desc(.))
#write.csv(vipVn.plasma, "vip.csv", row.names = TRUE)
#head(vipVn.plasma, 10)


## Scores Plotting

#built in plot (zoom in)
# plot(opls_model, typeVc = "x-score")  # score plot


## Scores Plotting

#predictive component scores
#extract predictive scores (p1) and orthogonal scores (o1)
p_scores <- getScoreMN(opls_model)
o_scores <- opls_model@orthoScoreMN  # slot with orthogonal components
#including group labels
scores <- data.frame(
    p1 = p_scores[, 1],
    o1 = o_scores[, 1],
    case_con = Y
  )
#get explained variance for predictive components
expl_var <- opls_model@modelDF[1]

#calculate the median coordinates for each group
medians <- scores %>%
  group_by(case_con) %>% 
  summarize(
    med_comp1 = median(p1), 
    med_comp2 = median(o1),
    .groups = "drop"
  )


#build the plot
opls.plasma_plot <- ggplot(scores, aes(x = p1, y = o1, color = case_con)) +
  geom_point(alpha=0.6) +
  stat_ellipse(aes(group = case_con), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = c("0" = "darkgray", "1" = "blue")) +
  theme_minimal() +
  labs(   
    x = paste0("Predictive Component (", round(expl_var$R2X[1], 2)*100, "% exp var)"),
    y = paste0("Orthogonal Component (", round(expl_var$R2X[2], 2)*100, "% exp var)"),
    color = "Case_Status"
  ) +
  theme(legend.position = "none",
        panel.background = element_rect(fill = "white", color = NA),
        plot.background = element_rect(fill = "white", color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines")) +
  
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = scores %>% 
                 left_join(medians, by = "case_con"),
               aes(x = med_comp1, y = med_comp2, xend = p1, yend = o1,
                   color = case_con),
               linewidth = 0.2, alpha = 0.4, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_comp1, y = med_comp2, color = case_con), 
             size = 3, shape = 4)  # Shape 4 is an 'x' marker


## Save file
ggsave(
  filename = "opls_2025-11-13.png",   # file name
  plot = opls.plasma_plot,                   # which plot to save
  width = 4,                  # width
  height = 4,                 # height
  units = "in",               # units: "in", "cm", or "mm"
  dpi = 600                   # resolution
)


```

```{r OPLS-DA - stool}

library(ropls)


## Prepare Data

#dataset
oplsda.stool <- metabo.stool_wide %>%
  transform_metabo(., remove_outliers = F, merge = F) %>% 
  merge(., gmbc_data %>% select(study_id, case_con))
# predictors (metabolites)
X <- oplsda.stool %>% select(-c(case_con,study_id)) %>% as.matrix()
# outcome variable (e.g., case vs. control)
Y <- as.factor(oplsda.stool$case_con)  


## Run PLS-DA (FORCE PLS-DA, no orthogonal components)
#opls_model.stool <- opls(X, Y, predI = 1, orthoI = NA, permI = 100) 
opls_model.stool <- opls(X, Y, predI = 2, orthoI = 0, permI = 100)  
  # predI = predictive components, use 1 to separate Y groups
  # orthoI = orthogonal components, NA lets ropls select best no. via CV
  # permI = permutation testing, confirms statistical validity of the model
  # R2Y = goodness of fit 
  # Q2 = cross-validated predictability


## Important Variables

vipVn.stool <- getVipVn(opls_model.stool) %>%   # Variable Importance in Projection (VIP) scores
  as.data.frame() %>% 
  arrange(desc(.))
#write.csv(vipVn.stool, "vip.csv", row.names = TRUE)
#head(vipVn.stool, 10)



#predictive component scores
#extract predictive scores (p1) and orthogonal scores (o1)
x_scores <- getScoreMN(opls_model.stool)

#including group labels
scores <- data.frame(
    p1 = x_scores[, 1],
    p2 = x_scores[, 2],
    case_con = Y
  )
#get explained variance for predictive components
expl_var <- opls_model.stool@modelDF[1]

#calculate the median coordinates for each group
medians <- scores %>%
  group_by(case_con) %>% 
  summarize(
    med_comp1 = median(p1), 
    med_comp2 = median(p2),
    .groups = "drop"
  )


#build the plot
pls.stool_plot <- ggplot(scores, aes(x = p1, y = p2, color = case_con)) +
  geom_point() +
  stat_ellipse(aes(group = case_con), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = c("0" = "darkgray", "1" = "blue")) +
  theme_minimal() +
  labs(   
    x = paste0("Latent Variable 1 (", round(expl_var$R2X[1], 2)*100, "% exp var)"),
    y = paste0("Latent Variable 2 (", round(expl_var$R2X[2], 2)*100, "% exp var)"),
    color = "Case Status"
  ) +
  theme(legend.position = c(1,1),#"right",
        legend.justification = c("right", "top"),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.size = unit(0.7, "lines"),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines")) +
  
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = scores %>% 
                 left_join(medians, by = "case_con"),
               aes(x = med_comp1, y = med_comp2, xend = p1, yend = p2,
                   color = case_con),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_comp1, y = med_comp2, color = case_con), 
             size = 3, shape = 4)  # Shape 4 is an 'x' marker

pls.stool_plot

```
 
```{r combined OPLS + PLS plot}

library(cowplot)


png("metabo_opls_2025-08-18.png", 
    width = 6.5 * 600, height = 3.25 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    opls.plasma_plot, pls.stool_plot,
    nrow = 1,
    rel_widths = c(1, 1)
  )
)
  
grid.text("A", x = unit(0.1, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("B", x = unit(0.6, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

dev.off()


```
 
  
Logistic Regression Models   
```{r plasma logreg - apply}

#then run: iterative modeling - apply
apply_iterative.modeling(
  predictor_vars <- metabo.plasma_wide %>% select(-study_id) %>% colnames, 
  adjustment_sets <- get_covariates(),
  dataset.model <- metabo.plasma_wide %>% transform_metabo(),
  file_name <- "metabo.plasma_log2.outlier"
)


```
    
```{r plasma logreg - volcano plot}

library(readxl)

##CLASS
data_class <- read_excel("GMBC1_Data_Summary_2025-07-30.xlsx", sheet = "1.Metabo_describe")
colnames(data_class) <- as.character(unlist(data_class[1, ]))
data_class <- data_class[-1, ]
data_class <- data_class %>% 
  select(Class, Metabolite)  %>% 
  mutate(Class = ifelse(Class=="Uremic toxin", "Other", Class))

## UNADJUSTED
metabo_results <- read_excel("Logreg_metabo.plasma_log2.outlier_2025-07-24.xlsx", sheet = "Sheet_unadj") %>% select(-last_col(), -last_col(1))


# Data preparation
results <- metabo_results %>% 
  merge(., data_class, by.x="term", by.y="Metabolite", all.x=TRUE) %>% 
  #create an index number for each class
  mutate(Class_Index = as.numeric(factor(Class))) %>% 
  #mutate if sig or not
  mutate(sig = q.value < 0.25) %>% 
  #remove problematic metabolite from plot
  filter(!term %in% c("LysoPC(17:0)","LysoPE(15:0)","LysoPC(16:0)"))


## Histgram
#build
hist_data <- ggplot_build(
  ggplot(results, aes(x = OR)) +
    geom_histogram(bins = 65)
)$data[[1]]
#rescale y values to negative
hist_data$y <- -hist_data$count / max(hist_data$count) / 10  # adjust height if needed

# Plot
volcano.p.0 <- ggplot() +
  #histogram (bars below 0)
  geom_col(data = hist_data, aes(x = x, y = y), fill = "black", width = hist_data$width[1]) +
  #geom points
  geom_point(data = results, aes(x=OR, y=-log10(q.value), color=Class, alpha = sig),
             size = 2) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.2), guide = "none") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha=1) +
  labs(
    x = NULL,#"Odds Ratio",
    y = "-log10(FDR)",
    color = "Class"
  ) +
  scale_x_continuous(limits = c(0.16, 3.33)) +
  #scale_y_continuous(limits = c(0, 0.81)) +
  coord_cartesian(ylim = c(-0.07, 0.81)) +
  theme_minimal() +
  guides(color = guide_legend(ncol = 2)) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )


```
  
```{r plasma logreg - volcano plot}

library(readxl)

##CLASS
data_class <- read_excel("GMBC1_Data_Summary_2025-07-30.xlsx", sheet = "1.Metabo_describe")
colnames(data_class) <- as.character(unlist(data_class[1, ]))
data_class <- data_class[-1, ]
data_class <- data_class %>% 
  select(Class, Metabolite)  %>% 
  mutate(Class = ifelse(Class=="Uremic toxin", "Other", Class))

## CONFOUNDER
metabo_results <- read_excel("Logreg_metabo.plasma_log2.outlier_2025-07-24.xlsx", sheet = "Sheet_confounder_vars") %>% select(-last_col(), -last_col(1))


# Data preparation
results <- metabo_results %>% 
  merge(., data_class, by.x="term", by.y="Metabolite", all.x=TRUE) %>% 
  #create an index number for each class
  mutate(Class_Index = as.numeric(factor(Class))) %>% 
  #mutate if sig or not
  mutate(sig = q.value < 0.25) %>% 
  #remove problematic metabolite from plot
  filter(!term %in% c("LysoPC(17:0)","LysoPE(15:0)","LysoPC(16:0)"))

## Histgram
#build
hist_data <- ggplot_build(
  ggplot(results, aes(x = OR)) +
    geom_histogram(bins = 100)
)$data[[1]]
#rescale y values to negative
hist_data$y <- -hist_data$count / max(hist_data$count) / 10  # adjust height if needed

# Plot
volcano.p.1 <- ggplot() +
  #histogram (bars below 0)
  geom_col(data = hist_data, aes(x = x, y = y), fill = "black", width = hist_data$width[1]) +
  #geom points
  geom_point(data = results, aes(x=OR, y=-log10(q.value), color=Class, alpha = sig),
             size = 2) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.2), guide = "none") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha=1) +
  labs(
    x = NULL,#"Odds Ratio",
    y = NULL,#"-log10(FDR)",
    color = "Class"
  ) +
  scale_x_continuous(limits = c(0.16, 3.33)) +
  #scale_y_continuous(limits = c(0, 0.81)) +
  coord_cartesian(ylim = c(-0.07, 0.81)) +
  theme_minimal() +
  guides(color = guide_legend(ncol = 2)) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )


```


```{r stool logreg - apply}

#then run: iterative modeling - apply
apply_iterative.modeling(
  predictor_vars <- metabo.stool_wide %>% select(-study_id) %>% colnames, 
  adjustment_sets <- get_covariates(),
  dataset.model <- metabo.stool_wide %>% transform_metabo(),
  file_name <- "metabo.stool_log2.outlier"
)

```
  
```{r stool logreg - volcano plot}

library(readxl)

##CLASS
data_class <- read_excel("GMBC1_Data_Summary_2025-07-30.xlsx", sheet = "1.Metabo_describe")
colnames(data_class) <- as.character(unlist(data_class[1, ]))
data_class <- data_class[-1, ]
data_class <- data_class %>% select(Class, Metabolite) %>% 
  mutate(Class = ifelse(Class=="Uremic toxin", "Other", Class))

## UNADJUSTED
metabo_results <- read_excel("Logreg_metabo.stool_log2.outlier_2025-07-24.xlsx", sheet = "Sheet_unadj")


# Data preparation
results <- metabo_results %>% 
  merge(., data_class, by.x="term", by.y="Metabolite", all.x=TRUE) %>% 
  #create an index number for each class
  mutate(Class_Index = as.numeric(factor(Class))) %>% 
  #mutate if sig or not
  mutate(sig = q.value < 0.25) #%>%  # & p.value < 0.05
  #remove problematic metabolite from plot
  #filter(!term %in% c("LysoPC(17:0)","LysoPC(16:0)"))

## Histgram
#build
hist_data <- ggplot_build(
  ggplot(results, aes(x = OR)) +
    geom_histogram(bins = 100)
)$data[[1]]
#rescale y values to negative
hist_data$y <- -hist_data$count / max(hist_data$count) / 2.5  # adjust height if needed

# Plot
volcano.s.0 <- ggplot() +
  #histogram (bars below 0)
  geom_col(data = hist_data, aes(x = x, y = y), fill = "black", width = hist_data$width[1]) +
  #geom points
  geom_point(data = results, aes(x=OR, y=-log10(q.value), color=Class, alpha = sig),
             size = 2) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.2), guide = "none") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha=1) +
  labs(
    x = "Odds Ratio",
    y = "-log10(FDR)",
    color = "Class"
  ) +
  scale_x_continuous(limits = c(0.7, 2.2)) +
  #scale_y_continuous(limits = c(0, 3.1)) +
  coord_cartesian(ylim = c(-0.25, 3.1)) +
  theme_minimal() +
  guides(color = guide_legend(ncol = 2)) +
  theme(
    legend.position = "none",
    text = element_text(size = 14)
  )


```
       
```{r stool logreg - volcano plot}

library(readxl)

##CLASS
data_class <- read_excel("GMBC1_Data_Summary_2025-07-30.xlsx", sheet = "1.Metabo_describe")
colnames(data_class) <- as.character(unlist(data_class[1, ]))
data_class <- data_class[-1, ]
data_class <- data_class %>% select(Class, Metabolite) %>% 
  mutate(Class = ifelse(Class=="Uremic toxin", "Other", Class))

## CONFOUNDER
metabo_results <- read_excel("Logreg_metabo.stool_log2.outlier_2025-07-24.xlsx", sheet = "Sheet_confounder_vars")


# Data preparation
results <- metabo_results %>% 
  merge(., data_class, by.x="term", by.y="Metabolite", all.x=TRUE) %>% 
  #create an index number for each class
  mutate(Class_Index = as.numeric(factor(Class))) %>% 
  #mutate if sig or not
  mutate(sig = q.value < 0.25) #%>% 
  #remove problematic metabolite from plot
  #filter(!term %in% c("LysoPC(17:0)","LysoPC(16:0)"))

## Histgram
#build
hist_data <- ggplot_build(
  ggplot(results, aes(x = OR)) +
    geom_histogram(bins = 100)
)$data[[1]]
#rescale y values to negative
hist_data$y <- -hist_data$count / max(hist_data$count) / 2.5  # adjust height if needed

# Plot
volcano.s.1 <- ggplot() +
  #histogram (bars below 0)
  geom_col(data = hist_data, aes(x = x, y = y), fill = "black", width = hist_data$width[1]) +
  #geom points
  geom_point(data = results, aes(x=OR, y=-log10(q.value), color=Class, alpha = sig),
             size = 2) +
  scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.2), guide = "none") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "black") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha=1) +
  labs(
    x = "Odds Ratio",
    y = NULL, #"-log10(FDR)",
    color = "Metabolite Class"
  ) +
  scale_x_continuous(limits = c(0.7, 2.2)) +
  #scale_y_continuous(limits = c(0, 3.1)) +
  coord_cartesian(ylim = c(-0.25, 3.1)) +
  theme_minimal() +
  guides(color = guide_legend(ncol = 2)) +
  theme(
    #legend.position = "none",
    legend.text = element_text(size = 6.8),
    legend.title = element_text(size = 8),
    legend.position = c(1, 0.98),    # inside top right (x,y in npc)
    legend.justification = c("right", "top"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.4, "lines"),
    #legend.spacing.y = unit(0.1, "cm"), 
    #legend.spacing.x = unit(0.1, "cm"), 
    text = element_text(size = 14)
   ) +
   guides(color = guide_legend(ncol = 2, byrow = TRUE)) # make it compact


```
 
```{r logreg combined plot}

library(cowplot)

png("metabo_logreg_2025-08-18.png", 
    width = 6.5 * 600, height = 6 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    volcano.p.0, volcano.p.1,
    volcano.s.0, volcano.s.1,
    legend,
    nrow = 2,
    ncol = 2,
    rel_widths = c(1,1,1,1)
  )
)

grid.text("A", x = unit(0.11, "npc"), y = unit(0.975, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("B", x = unit(0.58, "npc"), y = unit(0.975, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("C", x = unit(0.1, "npc"), y = unit(0.47, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("D", x = unit(0.58, "npc"), y = unit(0.47, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))


dev.off()

```


```{r metabo OR plot}

library(svglite)
library(openxlsx)
library(gridExtra)
library(grid)
library(ggtext)


## if repeated name, turn to ""


## Read in and format data
metabo_OR <- read.xlsx("OR_plot_2025-08-15.xlsx") %>%
  mutate(
    metabolite = str_replace(metabolite, "^plasma", "*plasma*"),  # italicize plasma
    metabolite = str_replace(metabolite, "^stool", "*stool*"),    # italicize stool
    metabolite = factor(metabolite, levels = c(
      #phytoestrogens
      "*plasma* – Genistein","*stool* – Genistein",
      "*plasma* – Daidzein","*stool* – Daidzein",
      "Glycitein","Matairesinol","Pinoresinol",
      #steroid-hormones
      "Estriol","16a-Hydroxyestrone",
      "*plasma* – DHEA-sulfate","*stool* – DHEA-sulfate",
      "Androsterone 3-sulfate",
      #bile acids
      "Murocholic acid",
      "Dehydrocholic acid","12-Ketochenodeoxycholic acid","7-Ketodeoxycholic acid",
      "Hyodeoxycholic acid","Lithocholic acid","Dioxolithocholic acid",
      "Glycoursodeoxycholic acid",
      "Cholic acid 3-sulfate","Hyocholic acid 3-sulfate",
      "Taurolithocholic acid","Tauroursodeoxycholic acid 3-sulfate",
      "Lithocholic acid 3-sulfate",
      "Deoxycholic acid 3-sulfate","Ursodeoxycholic acid 3-sulfate",
      "*plasma* – Ursodeoxycholic acid 7-sulfate","*stool* – Ursodeoxycholic acid 7-sulfate",
      #carnitines
      "Acetyl-carnitine",
      "Adipoyl-carnitine","Arachidonoyl-carnitine",
      "Butyryl-carnitine",
      "*plasma* – Glutaryl-carnitine","*stool* – Glutaryl-carnitine",
      "Hexanoyl-carnitine",
      "*plasma* – Linoleyl-carnitine","*stool* – Linoleyl-carnitine",
      "3-Methylglutaryl-carnitine","3-Methylcrotonyl-L-carnitine",
      "Octanoyl-carnitine","Oleoyl-carnitine",
      "Palmitoyl-carnitine","Stearoyl-carnitine",
      #other
      "*plasma* – Hypoxanthine","*stool* – Hypoxanthine",
      "*plasma* – Caproic acid","*stool* – Caproic acid",
      "Methionine","LysoPC(15:0)","p-Cresol sulfate"
    )))


## Build the HR plot
or_plot <- ggplot(metabo_OR, aes(y = fct_rev(metabolite), x = OR, shape = group, color = group)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = lower_CI, xmax = upper_CI), height = 0.4) +
  geom_point(size = 2.2) +
  scale_x_continuous(limits = c(0.00, 6), breaks = c(0.0,0.5,1.0,1.5,2,2.5,3,3.5,4)) +
  scale_color_manual(values = c("plasma" = "darkred", "stool" = "darkblue")) +
  scale_shape_manual(values = c("plasma" = 16, "stool" = 15)) +
  # Add annotations for HR and CI values to the right of the CI bars
  geom_text(aes(x = 4.5, 
                label = ifelse(!is.na(OR), sprintf("%.2f (%.2f–%.2f)      %.3f", 
                                                   OR, lower_CI, 
                                                   case_when(
                                                     metabolite=="Acetyl-carnitine" ~ 6.14,
                                                     upper_CI==4.44 ~ 7.92,
                                                     TRUE ~ upper_CI
                                                   ), 
                                                   q_value), "")),
                hjust = 0, size = 2.9, colour = "black") +
  theme_minimal(base_size = 11) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_markdown(hjust=1), 
    #axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9),
    legend.position = c(0.55, 0.95), #"bottom",
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.8, "lines"),
    plot.margin = unit(c(0.75, 0.5, 0.2, 0.5), "cm") 
  ) + 
  labs(color="Metabolite matrix",
       shape="Metabolite matrix")


or_plot


### SAVING THE PLOT

png("metabo_OR_2025-09-03.png", 
    width = 7.5 * 600, height = 8.25 * 600, 
    res = 600, bg = "white")

print(or_plot)
# Add the column titles 
grid.text("OR (95% CI)                                                       OR (95% CI)   q-value", 
          x = 0.685, y = 0.975, gp = gpar(fontsize = 10, fontface = "bold"))


dev.off()


```



# MULTI-OMICS


DIABLO

```{r DIABLO - data prep (all features) }

### DATA PREPARATION

#species selection
X.species <- metaphlan4_wide.t %>% 
  transform_metagenomics() %>% 
  select(study_id, 
         "s__Anaerotruncus_rubiinfantis|t__SGB25416",
          "s__Alistipes_shahii|t__SGB2295",
          "s__Lachnospiraceae_bacterium_AM48_27BH|t__SGB4706",
          "s__Alistipes_finegoldii|t__SGB2301",
          "s__Blautia_obeum|t__SGB4811",
          "s__Sutterella_wadsworthensis|t__SGB9286",
          "s__Alistipes_putredinis|t__SGB2318",
          "s__Mediterraneibacter_massiliensis|t__SGB4595",
          "s__Candidatus_Avimonas_narfia|t__SGB14941",
          "s__Mediterraneibacter_butyricigenes|t__SGB25493",
          "s__Blautia_SGB4815|t__SGB4815",
          "s__Faecalibacillus_intestinalis|t__SGB6754",
          "s__Clostridium_fessum|t__SGB4705",
          "s__Candidatus_Cibionibacter_quicibialis|t__SGB15286",
          "s__Anaerobutyricum_hallii|t__SGB4532",
          "s__GGB3057_SGB4059|t__SGB4059"
         ) %>% 
  drop_na() %>% 
  rename_with(
    ~ .x %>%
      sub("^s__", "", .) %>%                     # remove s__ at start
      sub("\\|t__SGB[0-9]+$", "", .) %>%         # remove |t__SGB### at end
      gsub("_", " ", .),                         # replace underscores with spaces
    .cols = -study_id
  ) %>% 
  rename("L. bacterium AM48-27BH" = "Lachnospiraceae bacterium AM48 27BH",
         "C. Avimonas narfia" = "Candidatus Avimonas narfia",
         "C. Cibionibacter quicibialis" = "Candidatus Cibionibacter quicibialis") %>% 
  arrange(study_id) %>% 
  column_to_rownames("study_id")

#enzyme selection
X.enzymes <- humann.enzymes_wide %>%
  transform_metagenomics() %>%
  select(study_id, 
          "1.8.98.1",
          "3.5.1.81",
          "2.7.1.113",
          "2.5.1.120",
          "3.4.13.21",
          "1.2.99.5", #1.2.7.12
          "6.3.4.15",
          "1.4.99.6",
          "3.4.22.8",
          "4.2.1.115",
          "2.4.1.319",
          "2.4.1.329",
          "2.1.1.173",
          "1.10.9.1", #7.1.1.6
          "5.1.3.8",
          "2.1.1.61"
         ) %>%
  drop_na() %>%
  rename_with(~ paste0("EC ", .x), .cols = -study_id) %>% 
  rename("EC 1.2.7.12" = "EC 1.2.99.5",
         "EC 7.1.1.6" = "EC 1.10.9.1") %>% 
  arrange(study_id) %>%
  column_to_rownames("study_id")

X.plasma  <- metabo.plasma_wide %>%
  transform_metabo(remove_outliers=F, merge=F) %>%
  select(study_id,
         "Genistein","Matairesinol","Murocholic.acid",
         "3-Methylglutaryl-carnitine","16a-Hydroxyestrone",
         "Daidzein","DHEA-sulfate","Hyocholic.acid.3-sulfate",
         "Estriol.(E3)","Hypoxanthine", "Pinoresinol", 
         "Glycoursodeoxycholic.acid","Hexanoyl-carnitine",
         "Linoleyl-carnitine","LysoPC(15:0)","Caproic.acid",
        "Glutaryl-carnitine","Methionine"
         ) %>%
  drop_na() %>% 
  rename_with(~ gsub("\\.", " ", .x), .cols = -study_id) %>% 
  rename("Estriol" = "Estriol (E3)") %>% 
  arrange(study_id) %>%
  column_to_rownames("study_id")

X.stool  <- metabo.stool_wide %>%
  transform_metabo(remove_outliers=F, merge=F) %>%
  select(study_id,
         "Adipoyl-carnitine","Genistein","DHEA-sulfate",
         "Octanoyl-carnitine","Oleoyl-carnitine","Palmitoyl-carnitine",
         "Butyryl-carnitine","Ursodeoxycholic.acid.7-sulfate",
         "Linoleyl-carnitine","Cholic.acid.3-sulfate",
         "3-Methylcrotonyl-L-carnitine","7-Ketodeoxycholic.acid",
         "Androsterone.3-sulfate","Glutaryl-carnitine","Daidzein",
         #Model 1 only, lower down
         "Lithocholic.acid.3-sulfate",
         "Cholic.acid.3-sulfate","Stearoyl-carnitine",
         "Deoxycholic.acid.3-sulfate",
         "Dioxolithocholic.acid","Hyodeoxycholic.acid",
         "12-Ketochenodeoxycholic.acid","Ursodeoxycholic.acid.3-sulfate",
         "Glycoursodeoxycholic.acid"
         ) %>% 
  drop_na() %>%
  rename_with(~ gsub("\\.", " ", .x), .cols = -study_id) %>% 
  arrange(study_id) %>%
  column_to_rownames("study_id")



#enzyme selection
X.pathways <- humann.path_wide %>%
  transform_metagenomics() %>%
  select(study_id, 
    "PWY-7992: superpathway of menaquinol-8 biosynthesis III",
    "RIBOSYN2-PWY: flavin biosynthesis I (bacteria and plants)",
    "PWY-7883: anhydromuropeptides recycling II",
    "PWY-5130: 2-oxobutanoate degradation I",
    "PWY-7761: NAD salvage pathway II (PNC IV cycle)",
    "PWY-6876: isopropanol biosynthesis (engineered)",
    "PWY-6901: superpathway of glucose and xylose degradation",
    "P42-PWY: incomplete reductive TCA cycle",
    "PWY-8131: 5'-deoxyadenosine degradation II",
    "METHGLYUT-PWY: superpathway of methylglyoxal degradation",
    "PWY-7013: (S)-propane-1,2-diol degradation",
    "PWY-6588: pyruvate fermentation to acetone",
    "PWY-7371: 1,4-dihydroxy-6-naphthoate biosynthesis II",
    "PWY-5973: cis-vaccenate biosynthesis",
    "PWY-5920: superpathway of heme b biosynthesis from glycine"
    ) %>%
  drop_na() %>%
  rename_with(~ sub(":.*", "", .x), .cols = -study_id) %>% 
  arrange(study_id) %>%
  column_to_rownames("study_id")

```
 
```{r DIABLO - run and plot }
  
### DATA PREPARATION ####

#Filter so that all contain the sample study_ids
shared_ids <- Reduce(intersect, 
                     list(rownames(X.species), rownames(X.enzymes),
                          rownames(X.pathways),
                          rownames(X.plasma), rownames(X.stool)))

#dropping outlier participants
shared_ids <- setdiff(shared_ids, c("GMBC_130", "GMBC_122"))
X.species <- X.species[shared_ids, ]
X.enzymes <- X.enzymes[shared_ids, ]
X.pathways <- X.pathways[shared_ids, ]
X.plasma <- X.plasma[shared_ids, ]
X.stool <- X.stool[shared_ids, ]

# Combine all dataframes into a single 
X <- list(species = X.species, enzymes = X.enzymes, pathways = X.pathways,
          plasma_metabolites = X.plasma, stool_metabolites = X.stool )

#Outcome (also filter for id)
Y <- gmbc_data %>% 
  filter(study_id %in% shared_ids) %>% 
  pull(case_con) %>% 
  as.factor()


### APPLY multiblock PLS-DA ####

library(mixOmics)  #beware, this overrides select function in dplyr and makes it not work properly
library(dplyr)

#run function
diablo_result <- block.splsda(X, Y, 
                              ncomp = 2, #default, first two PLS comp calculated
                              design = "full", #default, strength all relationships between df maximized
                              scale = TRUE) #default, each bloc standardized to zero means and unit variances


## Performance ####

# run repeated CV performance evaluation
perf.diablo = perf(diablo_result, validation = 'Mfold', 
                   M = 10, nrepeat = 10, 
                   dist = 'centroids.dist') 

perf.diablo$MajorityVote.error.rate
perf.diablo$WeightedVote.error.rate


# Use auroc to get ROC/AUC
# roc_diablo <- auroc(diablo_result, roc.comp = 1)


## DIAGNOSTIC PLOTS ####

# Component Correlation plots
plotDiablo(diablo_result, ncomp=1, object.size=10)
plotDiablo(diablo_result, ncomp=2)

# Plot to visualize sample separation across components
plotIndiv(diablo_result, comp = 1:2, group = Y, legend=T) # plot the samples

 
# Plot to show agreement of each sample across blocks (consistency)
plotArrow(diablo_result, group=Y)

## Correlation Circle Plot ####

#pull coords into df
var_coords <- plotVar(diablo_result, comp = c(1, 2), plot = FALSE)
# Create a single-row data frame to define the unit circle
circle_df <- data.frame(
  x0 = c(0, 0),
  y0 = c(0, 0),
  r  = c(1, 0.55)  # outer and inner circle
)
# Add correlation magnitude
var_coords$r <- sqrt(var_coords$x^2 + var_coords$y^2)
# Binary alpha: light if r < 0.5, darker if >= 0.5
var_coords$alpha <- ifelse(var_coords$r < 0.55, 0.2, 0.8)
#plotting
library(ggforce)
ggplot(var_coords, aes(x = x, y = y, color = Block, shape = Block)) +
  geom_point(aes(alpha = alpha), size = 3) +
  geom_text(aes(label = names), size = 2, vjust = -1, hjust = 0.5) +  # tiny labels
  ggforce::geom_circle(data = circle_df, aes(x0 = x0, y0 = y0, r = r),
                       inherit.aes = FALSE,
                       color = "darkgrey", linetype = "solid") +
  geom_hline(yintercept = 0, color = "grey", linetype = "dotted") +
  geom_vline(xintercept = 0, color = "grey", linetype = "dotted") +
  coord_fixed() +
  theme_minimal() +
  scale_color_manual(values = c("species" = "blue", "plasma_metabolites" = "orange",
                                "stool_metabolites" = "purple", "enzymes" = "darkgreen",
                                "pathways" = "red", "covariates" = 'grey20'
                                )) +
  scale_shape_manual(values = c("species" = 16, "plasma_metabolites" = 17,
                                "stool_metabolites" = 15, "enzymes" = 8,
                                "pathways" = 18, "covariates" = 3
                                )) +
  scale_alpha_identity() +  # use alpha as-is
  labs(x = "Component 1", y = "Component 2") +
  theme(
    #legend.position = "bottom",
    panel.grid = element_blank(),  # remove grid lines
    legend.title = element_blank()
    )




## CIRCOS PLOT  ####

# Circular version of the network plot, good for high-dim data
circosPlot(diablo_result, comp=2, cutoff=0.35, #comp=1 cutoff=0.58
           #line=TRUE,
           color.blocks = c("darkgoldenrod3","darkolivegreen3", "darkgreen",
                            "brown3","chocolate4"
                            ),
           color.cor = c("red","darkblue"),
           #showIntraLinks=TRUE,
           size.labels = 1,
           size.variables = 0.34
           )

# Get the similarity matrix
sim_matrix <- circosPlot(diablo_result, comp = 1, cutoff = 0.55, plot = FALSE)
# Convert matrix to long format
sim_df <- as.data.frame(sim_matrix) %>%
  mutate(Feature1 = rownames(sim_matrix)) %>%
  pivot_longer(
    cols = -Feature1,
    names_to = "Feature2",
    values_to = "Correlation"
  ) %>%
  filter(abs(Correlation) >= 0.55) %>% #apply cut off
  filter(Feature1 != Feature2) %>%  #remove self-correlations
  #Remove duplicate pairs by sorting Feature1/Feature2 alphabetically
  rowwise() %>%
  mutate(
    f1 = min(c(Feature1, Feature2)),
    f2 = max(c(Feature1, Feature2))
  ) %>%
  ungroup() #%>%
sim_df <- sim_df[, !names(sim_df) %in% c("Feature1", "Feature2")]
sim_df <- distinct(sim_df)
 

# Network Plot

network(diablo_result, cutoff=0.5)#,        
        # save = 'jpeg', 
        # name.save = 'DIABLO_network')



## FEATURE SELECTION ####

#Feature selection and interpretation (show top contributing variables per component, per block)
#threshold of importance >0.1

plotLoadings(diablo_result, comp = 1, block="species")
plotLoadings(diablo_result, comp = 2, block="species")

#species
selectVar(diablo_result, comp = 1)$species$value #%>% view()
selectVar(diablo_result, comp = 2)$species$value# %>% view()

selectVar(diablo_result, comp = 1)$enzymes$value #%>% view()
selectVar(diablo_result, comp = 2)$enzymes$value# %>% view()

selectVar(diablo_result, comp = 1)$pathways$value #%>% view()
selectVar(diablo_result, comp = 2)$pathways$value# %>% view()

selectVar(diablo_result, comp = 1)$plasma_metabolite$value #%>% view()
selectVar(diablo_result, comp = 2)$plasma_metabolite$value# %>% view()

selectVar(diablo_result, comp = 1)$stool_metabolite$value #%>% view()
selectVar(diablo_result, comp = 2)$stool_metabolite$value# %>% view()

selectVar(diablo_result, comp = 1)$covariates$value #%>% view()
selectVar(diablo_result, comp = 2)$covariates$value# %>% view()



```


