---
title: "GMBC_data analysis"
author: "AL"
date: "2024-10-22"
output: pdf_document
--- 
    
# DATA READING AND CLEANING
 
   
Set Up 
```{r libraries, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

#General
library(tidyverse)
library(stringr)
library(glue)
library(openxlsx) 

#Plotting
library(ggplot2)
library(ggtext)
library(gridExtra)
library(cowplot)

#Models
library(Publish)
library(purrr)
library(broom)

#Stats
library(tableone)
library(rstatix)

#WGS
library(vegan)

```
  
```{r estrobolome targets, control ids, WGS filter}

## Estrobolome data
estrobolome_targets <- read_csv("Estrobolome targets_2025-09-04.csv")

## List of Plate_IDs for Control Samples
technical_ids <- c("GMBC_00307","GMBC_00412","GMBC_00408", #plate1 012_T1-T3
                   "GMBC_00617","GMBC_00911","GMBC_00807", #plate2 079_T1-T3
                   "GMBC_01204","GMBC_01216","GMBC_01120", #plate3 023_T1-T3 (part 2)
                  "GMBC_01401","GMBC_01315","GMBC_01415",#plate4 012_T4-T6
                  "GMBC_01403","GMBC_01310","GMBC_01402", #plate4 079_T4-T6
                  "GMBC_01409","GMBC_01314","GMBC_01407") #plate4 023_T4-T6.P1 (part 1)

control_ids <- c("GMBC_00508","GMBC_00609","GMBC_00723","GMBC_00416", #plate1 NC_01-04
                  "GMBC_00212","GMBC_00801","GMBC_00305","GMBC_00902", #plate2 NC_05-08
                  "GMBC_01002","GMBC_01114","GMBC_01201", #plate3 NC_09-11
                  "GMBC_01414","GMBC_01317") #plate4 NC_12,NC_13


## Filter WGS data by a meta (epi) variable 
wgs_filter <- function(data, variable, var_value) {
  #convert alpha to a symbol for NSE
  alpha <- rlang::ensym(variable)
  #data modifications
  gmbc_data %>% 
    filter(!is.na(plate_ID)) %>%
    #pull out needed meta from gmbc_data
    select(plate_ID, !!alpha) %>%
    #add onto the metaphlan data
    merge(x=data, y=., by="plate_ID", all.x=TRUE) %>%
    #filter by the added meta variable
    filter(!!alpha==var_value) %>% 
    #drop meta var so only meteagenomic
    select(-!!alpha)
}


## Filter metabolite data by collection part
metabo_filter.p1 <- function(metabo_data) {
  metabo_data %>%
    merge(gmbc_data %>% select(study_id, collection_part), by = "study_id", all.x = TRUE) %>%
    filter(collection_part == "1") %>%
    select(-collection_part)
}


``` 
   
gmbc_data (REDCap, CDHQIII)              
```{r gmbc_data read in, impute}
 
#Read in gmbc_data from Data Cleaning.Rmd  
gmbc_data <- read.csv("gmbc_data_2025-08-22.csv") %>% 
  #ONLY INCLUDE INDIVIDUALS WITH BASELINE + FU DATA AVAILABLE
  drop_na(plate_ID) %>% #drops the 4 missing fu vals
  filter(!study_id %in% c("GMBC_043","GMBC_069","GMBC_120","GMBC_134")) #drops the part 1 vals

## Converting Data to Factor
gmbc_data <- gmbc_data %>%
  mutate(
    #selecting redcap variables only
      across(
      .cols = study_id:t_biol.occur,
      .fns = ~ if (!(cur_column() %in% 
              #numerical variables (for NOT in this list)
              c("s_age","s_height_cm","s_bmi","d_household_adults", "d_household_children","METmin",
              "rh_menarchy_age","s_menopause_age","rh_contraceptive_totalyr","rh_pregnancy_num",
              "rh_pregnancy_live","rh_pregnancy_age_first","rh_breastfeed_months",
               "rh_pregnancy_age_last","t_diameter.avg",
              "serum1_hours.diff","stool1_hours.diff","serum2_hours.diff",
              "stool2_hours.diff","stool_serum1.diff","stool_serum2.diff") 
        )) {
          as.factor(.)
        } else {
          .
        }
    ),
  #re-leveling factors
  ethnic_collapsed = factor(ethnic_collapsed, 
      levels = c("White","Asian","Other")),
  d_education_collapsed = factor(d_education_collapsed, 
    levels = c("<= High school","Post-high school training","Bachelor's","Postgraduate")),
  d_income_collapsed = factor(d_income_collapsed, 
      levels = c("49k","50k","100k","150k","PNA")) 
  )

## Updating GMBC variables for adjustment
gmbc_data <- gmbc_data %>% 
  mutate(
    #updates for control who became a case
    plate_ID = case_when(
      study_id == "GMBC_048" ~ "GMBC_00803", #first sample becomes 2nd run
      study_id == "GMBC_048_E2" ~ "GMBC_01318", #second sample becomes 3rd run
      TRUE ~ plate_ID ),
    #age at first pregnancy
    rh_pregnancy_age_first_cat = case_when(
        (rh_pregnancy_age_first >= 30) ~ as.factor(3), #geriatric pregnancy
        (rh_pregnancy_age_first < 30 & rh_pregnancy_age_first >= 25) ~ as.factor(2),
        (rh_pregnancy_age_first < 25) ~ as.factor(1),
        TRUE ~ as.factor(0) ),
    #smoking never / ever
    sm_summary = case_when(
        (sm_summary == 0) ~ as.factor(0),
        (sm_summary %in% c(1,2)) ~ as.factor(1),
        TRUE ~ NA ),
    #education level
    d_education_collapsed = case_when(
      (d_education_collapsed %in% c("<= High school","Post-high school training") ) ~ as.factor("< Bachelor's"),
      TRUE ~ d_education_collapsed ),
    d_education_collapsed = factor(d_education_collapsed), #refactor to collapse levels
    #ethnicity
    ethnic_collapsed = case_when(
      ethnic_collapsed %in% c("Asian","Other") ~ as.factor("Non-White"),
      TRUE ~ ethnic_collapsed),
    ethnic_collapsed = factor(ethnic_collapsed) #refactor to collapse levels
)


## Cleaning up the Follow-Up data
gmbc_data <- gmbc_data %>% 
  mutate(
    ## RADIATION
    t_rad.cat = case_when(
      t_rads.occur == 1 ~ "Yes",
      t_rads.occur == 0 ~ "No",
      is.na(t_rads.occur) ~ "Control"
     ),
    ## CHEMOTHERAPY
    t_chemo.cat = case_when(
      t_chemo.occur == 1 ~ "Yes",
      t_chemo.occur == 0 ~ "No",
      is.na(t_chemo.occur) ~ "Control"
     ),
    ## ENDOCRINE THERAPY
    t_endo.cat = case_when(
      t_endo.occur == 1 ~ "Yes",
      t_endo.occur == 0 ~ "No",
      is.na(t_endo.occur) ~ "Control"
     ),
    t_endo_drug.tamoxifen = case_when(
      t_endo.cat == "Control" ~ "Control",
      #088 switched to tamoxifen AFTER sample collection
      #135 was only briefly on tamoxifen between two different AIs
      study_id %in% c("GMBC_088","GMBC_088_E2","GMBC_135","GMBC_135_E2") ~ "No",
      t_endo.cat == "Yes" & t_endo_drug___1 == 1 ~ "Yes",
      TRUE ~ "No"
    ),
    t_endo_drug.arominhib = case_when(
      t_endo.cat == "Control" ~ "Control",
      #088 switched to tamoxifen AFTER sample collection
      #135 was only briefly on tamoxifen between two different AIs
      study_id %in% c("GMBC_088","GMBC_088_E2","GMBC_135","GMBC_135_E2") ~ "Yes",
      t_endo.cat == "Yes" & t_endo_drug___2 == 1 ~ "Yes",
      t_endo.cat == "Yes" & t_endo_drug___3 == 1 ~ "Yes",
      t_endo.cat == "Yes" & t_endo_drug___4 == 1 ~ "Yes",
      TRUE ~ "No"
    ),
    t_endo_drug.cat= case_when(
      t_endo.cat == "Control" ~ "Control",
      t_endo_drug.tamoxifen == "Yes" ~ "Tamoxifen",
      t_endo_drug.arominhib == "Yes" ~ "Aromatase_inhibitor",
      TRUE ~ "None"
    ),
    t_endo_current.cat = case_when(
      t_endo.cat == "Control" ~ "Control",
      t_endo.cat == "No" ~ "None",
      study_id %in% c("GMBC_009","GMBC_009_E2","GMBC_037","GMBC_037_E2",
                      "GMBC_098","GMBC_098_E2","GMBC_134","GMBC_134_E2") ~ "No",
      study_id %in% c("GMBC_102","GMBC_102_E2") ~ "Yes",
      t_endo_current == 1 ~ "Yes",
      t_endo_current == 0 ~ "No"
    ),
    
    ## BISPHOSPHATES
    t_bisp.cat = case_when(
      t_bisp.occur == 1 ~ "Yes",
      t_bisp.occur == 0 ~ "No",
      is.na(t_bisp.occur) ~ "Control"
     ),
    
    ## ANTIBIOTICS FU
    fu_antibiotic.clean = ifelse( (fu_antibiotic == 1 | t_antibiotics == 1), 1, 0),
    fu_antibiotic.clean = ifelse( is.na(fu_antibiotic.clean), 0, fu_antibiotic.clean),
    fu_antibiotic.clean = as.factor(fu_antibiotic.clean)
  )


## There is a case who's chemo was misinterpreted
gmbc_data$t_chemo[which(gmbc_data$study_id == "GMBC_020")] <- 0 #chemo 1 to 0
gmbc_data$t_bisp[which(gmbc_data$study_id == "GMBC_020")] <- 1 # add in bisp


## Single Imputation for Alcohol.g
#Identify complete cases for modeling
# complete_rows <- gmbc_data[!is.na(gmbc_data$Alcohol.g), ]
# #Fit a regression model using other covariates
# fit <- lm(Alcohol.g ~ s_age + s_bmi + ethnic_collapsed + d_education_collapsed + hormone_current.p1 +
#             sm_summary + pa_cat + fu_antibiotic.clean + family_history + rh_pregnancy_live + rh_pregnancy_age_first_cat + rh_breastfeed_months + rh_contraceptive + hormone_ever, 
#           data = complete_rows)
# #Predict missing Alcohol.g values
# missing_rows <- gmbc_data[is.na(gmbc_data$Alcohol.g), ]
# gmbc_data$Alcohol.g[is.na(gmbc_data$Alcohol.g)] <- predict(fit, newdata = missing_rows)
#this came up with GMBC_096 = -1.137952, and GMBC_128 = 3.556815
gmbc_data <- gmbc_data %>%
  mutate(Alcohol.g = case_when(
    single_id == "GMBC_096" ~ 0,
    single_id == "GMBC_128" ~ 3.556815,
    TRUE ~ Alcohol.g))

``` 

MetaPhlAn4 (stool)        
```{r metaphlan data read and cleaning}

## Read in the data
metaphlan4_all <- 
  #force three cols to properly read in
  read.table("profiled_metagenome.tsv", sep = "\t", fill = TRUE, col.names = c("plate_ID", "clade_name", "clade_taxid", "relative_abundance", "coverage", "estimated_number_of_reads_from_the_clade"), quote = "", comment.char = "") %>%
  #drop front matter
  slice(-(1:329)) %>% 
  mutate(name = str_extract(clade_name, "[^|]+$"),
         #need to convert to numeric
         across(relative_abundance:estimated_number_of_reads_from_the_clade, as.numeric)) %>% 
  #derive rank
  mutate(rank = case_when(
    grepl("k__", name) ~ "kingdom",
    grepl("p__", name) ~ "phylum",
    grepl("c__", name) ~ "class",
    grepl("o__", name) ~ "order",
    grepl("f__", name) ~ "family",
    grepl("g__", name) ~ "genus",
    grepl("s__", name) ~ "species",
    grepl("t__", name) ~ "subspecies",
    TRUE ~ "other"
  ))


## Data Selection

#Data selection
metaphlan4 <- metaphlan4_all %>% 
  # Drop the negative and technical controls
  filter(!plate_ID %in% control_ids) %>%
  filter(!plate_ID %in% technical_ids) %>% 
  # Drop those with NO part 2 DATA!
  filter(!plate_ID %in% c("GMBC_00615","GMBC_00203","GMBC_01104","GMBC_01223"))

#The third percentile of data = 0.00078
min_abundance_threshold <- quantile(metaphlan4 %>% pull(relative_abundance), probs = 0.03)

#Calculating frequency of taxa present
filtered_taxa <- metaphlan4 %>%
  #binary presence/absence matrix
  mutate(presence = ifelse(is.na(relative_abundance), FALSE, relative_abundance > min_abundance_threshold)) %>% #0
  #calc presence frequency for each taxon
  group_by(name) %>%
  summarize(
    presence_count = sum(presence), #243 is the max (should be 241?)
    frequency = presence_count / n_distinct(metaphlan4$plate_ID) #presence/total count to get %
  )
#Filter taxa based on prevalence thresholds
filtered_taxa10 <- filtered_taxa %>% filter(frequency >= 0.1) %>% pull(name)

#Data Filtering 
metaphlan4_long <- metaphlan4 %>%
  filter(name %in% filtered_taxa10) %>% 
  arrange(plate_ID)


## Rename the subspecies to include the species in the name (more informative)
metaphlan4_long <- metaphlan4_long %>% 
  mutate(name = if_else( rank == "subspecies", #if subspecies level
                         str_extract(clade_name, "s__[^|]+\\|t__[^|]+"), #rename to include species
                         name) ) #otherwise leave


## Transforming to wider form
metaphlan4_wide <- metaphlan4_long %>% 
  filter(rank %in% c("species","genus","subspecies")) %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s

## Transforming to wider form for SPECIES data
metaphlan4_wide.s <- metaphlan4_long %>% 
  filter(rank=="species") %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s

## Transforming to wider form for SUBSPECIES data
metaphlan4_wide.t <- metaphlan4_long %>% 
  filter(rank=="subspecies") %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s

## Transforming to wider form for GENUS data
metaphlan4_wide.g <- metaphlan4_long %>% 
  filter(rank=="genus") %>%
  pivot_wider(., 
    id_cols=c(plate_ID), #sample stays as row
    names_from=name, #name becomes column with rel abun as value
    values_from=relative_abundance,
    values_fill = 0)  #covert NA relative abundance to 0s


```   

HUMAnN3 (stool) 
```{r humann.path_nopipe read in}

## Read In File
humann.path_nopipe <- read.csv("humann.path_nopipe_2025-04-15.csv") %>% 
  rename(relative_abundance = value)  %>% 
  # Remove controls
  filter(!plate_ID%in%technical_ids) %>% #tech ids
  filter(!plate_ID%in%control_ids) %>%  #neg con
  # Drop those with NO part 2 DATA!
  filter(!plate_ID %in% c("GMBC_00615","GMBC_00203","GMBC_01104","GMBC_01223"))


## Data Filtering

#third percentile of data = 8.399325e-07(Robertson 3 × 10−7% = 0.000000003)
min_abundance_threshold <- humann.path_nopipe %>% 
  filter(!name %in% c("UNMAPPED","UNINTEGRATED")) %>% 
  pull(relative_abundance) %>% 
  quantile(., probs = 0.03)


#calculating frequency of taxa present
filtered_path5 <- humann.path_nopipe %>%
  #binary presence/absence matrix
  mutate(presence = ifelse(is.na(relative_abundance), FALSE, relative_abundance > min_abundance_threshold)) %>%
  #calc presence frequency for each taxon
  group_by(name) %>%
  summarize(
    presence_count = sum(presence),  #243 is the max (should be 241?)
    #presence/total count to get %
    frequency = presence_count / n_distinct(humann.path_nopipe$plate_ID)
  ) %>%
  #filter taxa based on the 5% threshold
  filter(frequency >= 0.05) %>%
  #pull out names
  pull(name)


#apply
humann.path_long <- humann.path_nopipe %>% 
  #remove unintegrated and unmapped samples
  filter(!name %in% c("UNMAPPED","UNINTEGRATED")) %>% 
  #filtering dataframe for >5% frequency
  filter(name %in% filtered_path5)

## Transforming to wider form
humann.path_wide <- humann.path_long %>% 
  pivot_wider(id_cols=c(plate_ID), #sample stays as row
              names_from=name, #name becomes column with rel abun as value
              values_from=relative_abundance,
              values_fill = 0)  #covert NA relative abundance to 0s


``` 

```{r humann.enzymes_nopipe read and clean}

## Read and Clean

#Read in nopipe data and clean
humann.enzymes_nopipe <- read.csv("genefamilies_relab_ec_cleaned_2025-01-03.csv") %>% 
  # Remove control samples
  filter(!plate_ID%in%technical_ids) %>% #tech ids
  filter(!plate_ID%in%control_ids) %>%  #neg con
  # Drop those with NO part 2 DATA!
  filter(!plate_ID %in% c("GMBC_00615","GMBC_00203","GMBC_01104","GMBC_01223")) %>% 
  ## Merge Name codes in
  merge(.,
        read.delim("map_ec_name.txt", header = FALSE) %>%
        rename(name = V1, Enzyme_name = V2),
        by = "name") %>% 
  # convert to numeric
  mutate(value = as.numeric(value)) %>% 
  #rename
  rename(relative_abundance = value)
  

## Data Filtering

#third percentile of data
min_abundance_threshold <- quantile(humann.enzymes_nopipe %>% pull(relative_abundance), probs = 0.03)

#calculating frequency of taxa present
filtered_enzyme5 <- humann.enzymes_nopipe %>%
  #binary presence/absence matrix
  mutate(presence = ifelse(is.na(relative_abundance), FALSE, relative_abundance > min_abundance_threshold)) %>%
  #calc presence frequency for each taxon
  group_by(name) %>%
  summarize(
    presence_count = sum(presence),
    #presence/total count to get %
    frequency = presence_count / n_distinct(humann.enzymes_nopipe$plate_ID)
  ) %>% 
  #filter taxa based on the 5% threshold
  filter(frequency >= 0.05) %>%
  #pull out names
  pull(name)

#applying
humann.enzymes_long <- humann.enzymes_nopipe %>%
  #filtering dataframe for >5% frequency
  filter(name %in% filtered_enzyme5)

## Transforming to wider form
humann.enzymes_wide <- humann.enzymes_long %>% 
  pivot_wider(id_cols = c(plate_ID), #sample stays as row
              names_from = name, #Enzyme name has special characters that crash
              values_from = relative_abundance,
              values_fill = 0)  #covert NA relative abundance to 0s


```

Metabolomics (blood & stool)   
```{r metabo plasma - reading and cleaning}

library(openxlsx)

#Set Pathway (file with multiple sheets)
pathway <- "GMBC_Metabolite_Results_2025-04-30.xlsx"

#Read in Metabolite Class Info
metabo_class <- read.xlsx(pathway, sheet = 4) %>%
  select(Class, Metabolite) %>% 
  #replace " " with "." to match
  mutate(Metabolite.match = gsub(" ", ".", Metabolite))
  
#Read in metabolomics abundances
plasma_abund <- read.xlsx(pathway, sheet = 1) 

#Clean up the study codes
plasma_abund <- plasma_abund %>% 
  rename(study_id = "Sample_ID") %>% 
  #remove the added front code
  mutate(study_id =   gsub("^\\d{3}_", "", study_id)) %>% 
  #ensure numeric
  mutate(across(-study_id, as.numeric) ) %>% 
  #removing those who have part 2 but no p1
  filter(!study_id %in% c("GMBC_049_E2","GMBC_137_E2")) %>% 
  #removing those who have p1 but no p2
  filter(!study_id %in% c("GMBC_028","GMBC_043","GMBC_069","GMBC_074","GMBC_120","GMBC_134"))

## Calculating prevalence frequency of metabolites present
filtered_metabo10 <- plasma_abund %>%
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relab") %>%
  mutate(presence = ifelse(is.na(relab), FALSE, relab > 0)) %>%
  group_by(Metabolite) %>%
  summarize(frequency = sum(presence) / n_distinct(plasma_abund$study_id)) %>%
  filter(frequency >= 0.10) %>%
  pull(Metabolite)


## Data Filtering >10% prevalence frequency (metabo_wide is filtered)
metabo.plasma_wide <- plasma_abund %>%
  select(study_id, all_of(filtered_metabo10)) %>% 
  merge(., gmbc_data %>% select(study_id), by="study_id", all.x=T)

## Long form data
metabo.plasma_long <- metabo.plasma_wide %>% 
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relative_abundance")


``` 
 
```{r metabo stool - reading and cleaning}

library(openxlsx)

#Set Pathway (file with multiple sheets)
pathway <- "GMBC_Metabolite_Results_2025-04-30.xlsx"

#Read in Metabolite Class Info
stool_abund <- read.xlsx(pathway, sheet = 3)

#Clean up the study codes
stool_abund <- stool_abund %>%
  rename(study_id = "Sample_ID") %>% 
  #remove the added front code
  mutate(study_id =   gsub("^\\d{3}_", "", study_id)) %>% 
  #ensure numeric
  mutate(across("Fecal.wet.mass.(mg)":"Zymosterol", as.numeric) ) %>% 
  select(!c("Fecal.wet.mass.(mg)","Extractant.volume.(mL)","Dry.mass.(mg)")) %>% 
  #remove those missing follow-up data
  filter(!study_id %in% c("GMBC_043","GMBC_069","GMBC_120","GMBC_134",
    #missing p1
    "GMBC_011_E2"))


## Calculating prevalence frequency of metabolites present
filtered_metabo10 <- stool_abund %>%
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relab") %>%
  mutate(presence = ifelse(is.na(relab), FALSE, relab > 0)) %>%
  group_by(Metabolite) %>%
  summarize(frequency = sum(presence) / n_distinct(stool_abund$study_id)) %>%
  filter(frequency >= 0.10) %>%
  pull(Metabolite)


## Data Filtering >10% prevalence frequency (metabo_wide is filtered)
metabo.stool_wide <- stool_abund %>%
  select(study_id, all_of(filtered_metabo10)) %>% 
  merge(., gmbc_data %>% select(study_id), by="study_id", all.x=T)

## Long form data
metabo.stool_long <- metabo.stool_wide %>% 
  pivot_longer(-study_id, names_to = "Metabolite", values_to = "relative_abundance")


``` 
   


   
# ANALYSES FUNCTIONS
      
Descriptive   
```{r alpha diversity functions }

#alpha is within sample diversity - what and how much is there in a sample
#beta is between sample diversity - how similar is one sample to another
 
## Calculate Alpha Diversity
alpha_calc_data <- function(data_long, metaphlan = FALSE) {
  
  ## Calculate alpha diversity as new cols
  alpha_data <- data_long %>% 
    #all species in a sample
    { if (metaphlan) filter(., rank == "subspecies") else . } %>%
    #plate_id is per sample at each time
    group_by(plate_ID) %>% 
    #calculate alpha metrics
    summarize(shannon = diversity(relative_abundance, "shannon"),
              richness = specnumber(relative_abundance)) %>% 
    #merge alpha metrics onto gmbc_data
    merge(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con)
          , ., by="plate_ID", all.x = TRUE) 
  
  return (alpha_data) 
}
 
## Alpha Div Significant Testing

#cases v controls (wilcoxon rank sum)
alpha_sig_test.groups <- function(alpha_data, col_part) {

  #Filter which collection part
  alpha_data <- alpha_data %>% 
    filter(collection_part == col_part) %>% 
    drop_na()
  
  #SHANNON
  #calculate group medians alpha div
  meds <- alpha_data %>% 
    filter(shannon > 0) %>% #only include where alpha data
    group_by(case_con) %>% #group by cancer status
    summarise(median_alpha = median(shannon), .groups = "drop") #calculate group medians
  #calculating p values using wilcoxon rank sum test
  p.shannon <- wilcox.test(shannon ~ case_con, data = alpha_data%>%filter(shannon>0))
  #print out metrics
  print(glue("The group medians of SHANNON diversity:
           Controls = {meds$median_alpha[meds$case_con == 0]}
           Cases = {meds$median_alpha[meds$case_con == 1]}
           Abs median difference: {abs(diff(meds$median_alpha))}
           The p value (Wilcoxon): {p.shannon$p.value}"))

  #RICHNESS
  #calculate group medians alpha div
  meds <- alpha_data %>% 
    filter(richness > 0) %>% #only include where alpha data
    group_by(case_con) %>% #group by cancer status
    summarise(median_alpha = median(richness), .groups = "drop") #calculate group medians
  #calculating p values using wilcoxon rank sum test
  p.richness <- wilcox.test(richness ~ case_con, data = alpha_data%>%filter(richness>0))
  #print out metrics
  print(glue("The group medians of RICHNESS:
           Controls = {meds$median_alpha[meds$case_con == 0]}
           Cases = {meds$median_alpha[meds$case_con == 1]}
           Abs median difference: {abs(diff(meds$median_alpha))}
           The p value (Wilcoxon): {p.richness$p.value}"))
  
  return(list(p_shannon = p.shannon$p.value, 
            p_richness = p.richness$p.value)) 
}
#baseline v follow up (wilcoxon signed rank)
alpha_sig_test.paired <- function(alpha_data, group) {

  #Filter for whether looking in cases or controls
  alpha_data <- alpha_data %>% 
    filter(case_con == group) %>% 
    filter(shannon > 0, richness > 0)
  
  # Pivot to wide format for paired testing
  alpha_wide <- alpha_data %>%
    select(single_id, collection_part, shannon, richness) %>%
    pivot_wider(names_from = collection_part, values_from = c(shannon, richness), 
                names_prefix = "t")

  # Ensure both timepoints are present
  alpha_wide <- alpha_wide %>% 
    filter(!is.na(shannon_t1), !is.na(shannon_t2), 
           !is.na(richness_t1), !is.na(richness_t2))
  
  # SHANNON
  p.shannon <- wilcox.test(alpha_wide$shannon_t1, alpha_wide$shannon_t2, paired = TRUE)
  meds.shannon <- c(median(alpha_wide$shannon_t1), median(alpha_wide$shannon_t2))
  print(glue(
    "The group medians of SHANNON diversity:
     Baseline = {meds.shannon[1]}
     Follow up = {meds.shannon[2]}
     Abs median difference: {abs(diff(meds.shannon))}
     The p value (Wilcoxon-paired): {p.shannon$p.value}"
  ))

  # RICHNESS
  p.richness <- wilcox.test(alpha_wide$richness_t1, alpha_wide$richness_t2, paired = TRUE)
  meds.richness <- c(median(alpha_wide$richness_t1), median(alpha_wide$richness_t2))
  print(glue(
    "The group medians of RICHNESS:
     Baseline = {meds.richness[1]}
     Follow up = {meds.richness[2]}
     Abs median difference: {abs(diff(meds.richness))}
     The p value (Wilcoxon-paired): {p.richness$p.value}"
  ))

  return(list(p_shannon = p.shannon$p.value, 
              p_richness = p.richness$p.value)) 
}

```

Variable Sets 
```{r adjustment sets }

get_covariates <- function(type){

  adjustment_sets <- list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1", "sm_summary","pa_cat","Alcohol.g",
       "fu_antibiotic.clean"
  )
  )

  return(adjustment_sets)

}

```
 
Logistic Regression      
```{r iterative modeling - function }

library(lme4)
library(broom.mixed)
 
## Build the function to apply across multiple variables 
iterative_modeling <- function(column_name, data_set, model_vars) {
  
  #drop any NA for the variable of interest
  data_set <- data_set %>%
    drop_na(!!sym(column_name)) #turn to symbol to ref
  
  # Calculate sample size N
  N <- nrow(data_set)
  
  #construct the formula with the changing column as the first predictor and covariates
  formula <- as.formula(
    paste0(
      #include backticks to allow special characters in the model formula
      "`", column_name, "` ~ case_con * collection_part + (1 | single_id)",
      #if covariates, add in, otherwise if unadjusted model add nothing
      if (length(model_vars) > 0) paste0("+", paste(model_vars, collapse = " + ")) else ""
      )
    )
  
  #fit the model
  model <- lmerTest::lmer(
    formula, 
    data = data_set
  )
  
  #print model summary for first taxa, enzyme, metabolite
  if (column_name %in% c("s__Escherichia_coli","`Genistein`","`1.1.1.62`","Genistein")) { #"`g_Blautia`",
    print(formula)
    print(summary(model))
  }
  
  #pull out results
  result <- tidy(model, conf.int = TRUE)  %>%

    #filter to pull out rows of interest
    filter(term %in% c("case_con1:collection_part2"))%>% #"case_con1", 
    # Add feature name and clean term label
    mutate(feature = column_name) %>%
    select(feature, term, estimate, conf.low, conf.high, p.value) %>%
    # Create summary column
    mutate(
      estimate_CI = paste0(
        formatC(estimate, format = "f", digits = 2),     " (", 
        formatC(conf.low, format = "f", digits = 2), ", ",
        formatC(conf.high, format = "f", digits = 2), ")" 
      ), 
      term = ifelse(term=="case_con1", 
                    "Case Status", 
                    "Interaction")
    )
  
  #add col for sample size
  result$N <- N
  
  #return result
  return(result) 
  
}

## Build the function to apply across multiple variables 
iterative_modeling.therapy <- function(column_name, data_set, model_vars) {
  
  #Clean data
  data_set <- data_set %>%
    # SELECT CASES ONLY
    filter(case_con == 1) %>% 
    #drop any NA for the variable of interest
    drop_na(!!sym(column_name)) #turn to symbol to ref
  
  #factor data
  data_set$t_endo_drug.cat <- factor(data_set$t_endo_drug.cat,
                                       levels = c("None","Aromatase_inhibitor","Tamoxifen"))
  
  # Calculate sample size N
  N <- nrow(data_set)
  
  #construct the formula with the changing column as the first predictor and covariates
  formula <- as.formula(
    paste0(
      #include backticks to allow special characters in the model formula
      "`", column_name, "` ~ t_endo_drug.cat * collection_part + (1 | single_id)",
      #if covariates, add in, otherwise if unadjusted model add nothing
      if (length(model_vars) > 0) paste0("+", paste(model_vars, collapse = " + ")) else ""
      )
    )
  
  #fit the model
  model <- lmerTest::lmer(
    formula, 
    data = data_set
  )
  
  #print model summary for first taxa, enzyme, metabolite
  if (column_name %in% c("s__Escherichia_coli","`Genistein`","`1.1.1.62`","Genistein")) { #"`g_Blautia`",
    print(formula)
    print(summary(model))
  }
  
  #pull out results
  result <- tidy(model, conf.int = TRUE)  %>%
    #filter to pull out rows of interest
    filter(term %in% c("t_endo_drug.catAromatase_inhibitor:collection_part2", "t_endo_drug.catTamoxifen:collection_part2"))%>%
    # Add feature name and clean term label
    mutate(feature = column_name) %>%
    select(feature, term, estimate, conf.low, conf.high, p.value) %>%
    # Create summary column
    mutate(
      estimate_CI = paste0(
        formatC(estimate, format = "f", digits = 2),     " (", 
        formatC(conf.low, format = "f", digits = 2), ", ",
        formatC(conf.high, format = "f", digits = 2), ")" 
      ), 
      term = ifelse(term=="t_endo_drug.catAromatase_inhibitor:collection_part2", 
                    "Interaction (A Inhibitor)", 
                    "Interaction(ER Blocker)")
    )
  
  #add col for sample size
  result$N <- N
  
  #return result
  return(result) 
  
}

```     
 
```{r iterative modeling - apply }

library(writexl)

## PRE DEFINE VARIABLES

#predictor_vars
#adjustment_sets 
#dataset.model


## APPLY THE FUNCTION 

apply_iterative.modeling <- function(predictor_vars, adjustment_sets, dataset.model, file_name) {
  
  #Create empty results list
  df_results <- list()
  
  #Loop through models for each adjustment level
  for (i in names(adjustment_sets) ) {
    
    #select what to include in the adjustment set
    model_vars = adjustment_sets[[i]]
    
    #Remove NA vars
    dataset.model <- dataset.model %>% 
      filter(if_all(all_of(model_vars), ~ !is.na(.)))
    
    #Apply the function to each column in the selected range and combine results
    log_model.results <- map_dfr(predictor_vars, function(column_name) {
      iterative_modeling(column_name = column_name,  #.therapy
                         data_set = dataset.model, 
                         model_vars = model_vars)
    }) 
    
    #Calculate a q value (FDR adjustment for multiple testing)
    log_model.results <- log_model.results %>% 
      group_by(term) %>% 
      mutate(q.value = round(p.adjust(p.value, method = "BH"), 3)) %>% 
      ungroup() %>%
      #clean up the names
      mutate(feature_clean = feature %>%
               gsub("\\.", " ", .) %>%           # Replace dots with spaces
               gsub("`", "", .) %>%              # Remove backticks
               sub("^[sgt]__", "", .) %>%        # Remove leading s__/g__/t__
               gsub("_", " ", .)                 # Replace underscores with spaces
             ) %>% 
      #sort
      select(feature_clean, term, estimate_CI, p.value, 
             q.value, N, estimate, conf.low, conf.high, feature) %>% 
      arrange(p.value)#term, feature_clean)
    
    #Store the result in the list with a dynamic sheet name
    df_results[[paste0("Sheet_", i)]] <- log_model.results
    
  } 
  
  #Save to Excel (multiple sheets)
  write_xlsx(df_results, sprintf("GLMMs_%s_%s.xlsx", file_name, as.character(Sys.Date())))

}

apply_iterative.modeling.therapy <- function(predictor_vars, adjustment_sets, dataset.model, file_name) {
  
  #Create empty results list
  df_results <- list()
  
  #Loop through models for each adjustment level
  for (i in names(adjustment_sets) ) {
    
    #select what to include in the adjustment set
    model_vars = adjustment_sets[[i]]
    
    #Remove NA vars
    dataset.model <- dataset.model %>% 
      filter(if_all(all_of(model_vars), ~ !is.na(.)))
    
    #Apply the function to each column in the selected range and combine results
    log_model.results <- map_dfr(predictor_vars, function(column_name) {
      iterative_modeling.therapy(column_name = column_name,  #.therapy
                         data_set = dataset.model, 
                         model_vars = model_vars)
    }) 
    
    #Calculate a q value (FDR adjustment for multiple testing)
    log_model.results <- log_model.results %>% 
      group_by(term) %>% 
      mutate(q.value = round(p.adjust(p.value, method = "BH"), 3)) %>% 
      ungroup() %>%
      #clean up the names
      mutate(feature_clean = feature %>%
               gsub("\\.", " ", .) %>%           # Replace dots with spaces
               gsub("`", "", .) %>%              # Remove backticks
               sub("^[sgt]__", "", .) %>%        # Remove leading s__/g__/t__
               gsub("_", " ", .)                 # Replace underscores with spaces
             ) %>% 
      #sort
      select(feature_clean, term, estimate_CI, p.value, 
             q.value, N, estimate, conf.low, conf.high, feature) %>% 
      arrange(p.value)
    
    #Store the result in the list with a dynamic sheet name
    df_results[[paste0("Sheet_", i)]] <- log_model.results
    
  } 
  
  #Save to Excel (multiple sheets)
  write_xlsx(df_results, sprintf("GLMMs_%s_%s.xlsx", file_name, as.character(Sys.Date())))

}


```
 
 

# RESULTS

## Sample Characteristics
```{r tableone}


## Table One Function
table1_vars <- c(
  #demographics
  "s_age","s_bmi",
  "ethnic_collapsed","d_education_collapsed",
  #health
  "health_condition",
  "family_history",
  "rh_menarchy_age","s_menopause_age",
  "rh_contraceptive", "rh_pregnancy_live","rh_pregnancy_age_first",
  "rh_pregnancy_age_first_cat","rh_breastfeed_months",
  "hormone_ever","hormone_current.p1", 
  "fu_antibiotic.clean",
  #risk factors
  "pa_cat","sm_summary","Alcohol.g",
  #Tumor
  "t_er","t_pr","t_her2","t_stage","t_grade",
  "t_rad.cat","t_rad_dose.cat",
  "t_chemo.cat","t_chemo_cycles.cat",
  "t_endo.cat", "t_endo_drug.tamoxifen", "t_endo_drug.arominhib",
  "t_endo_drug.cat",
  "t_endo_current.cat",
  "t_bisp.cat"
)



#Create list of nonnormal vals to run wilcoxon test on (only works for numerical vals, not categorical)
nonnormal_vars <- c(
  "s_age","s_bmi","rh_menarchy_age",
  "s_menopause_age","rh_pregnancy_live","rh_breastfeed_months",
  "Alcohol.g"
)


#running tableone function
gmbc.table1 <- gmbc_data %>% 
  filter(collection_part==1) %>% #so not repeating
  filter(case_con == 1) %>% #cases only
  #filter(t_endo_drug.cat == "None") %>% 
  CreateTableOne(data=., 
                 vars=table1_vars,
                 #strata="case_con",
                 strata="t_endo_drug.cat",
                 test=T, 
                 addOverall=F)
print(gmbc.table1, showAllLevels=T, nonnormal = nonnormal_vars)


```


## METAGENOMICS
  
#### Summary
```{r metaphlan species info - part 2}

## PRE-Filtering Unique Species and Prevalent Phylums
metaphlan4.2 <- wgs_filter(metaphlan4, collection_part, 2)

#overall (pre filtering) - after removal of controls
metaphlan4.2 %>% filter(rank == "genus") %>% distinct(name) %>% nrow()
metaphlan4.2 %>% filter(rank == "species") %>% distinct(name) %>% nrow()
metaphlan4.2 %>% filter(rank == "subspecies") %>% distinct(name) %>% nrow()

#kingdom level data - summing relative abundances for each kingdom across all samples
metaphlan4.2 %>%
  filter(rank == "kingdom") %>% 
  group_by(name) %>%
  summarise(total_abundance = sum(relative_abundance, na.rm = TRUE)) %>%
  mutate(proportion = total_abundance / sum(total_abundance))

metaphlan4.2 %>%
  filter(rank == "species") %>% 
  filter(str_detect(clade_name, "k__Eukaryota|k__Archaea")) %>%
  distinct(clade_name)


## POST-Filtering Unique and Prevalent Species - PART TWO SAMPLES ONLY

metaphlan4_long.2 <- wgs_filter(metaphlan4_long, collection_part, 2)
metaphlan4_long.2 %>% filter(rank=="genus") %>% distinct(name) %>% nrow()

metaphlan4_long %>% filter(rank=="subspecies") %>% distinct(name) %>%  nrow()

#kingdom level data - summing relative abundances for each kingdom across all samples
metaphlan4_long %>%
  filter(rank == "phylum") %>% 
  group_by(name) %>%
  summarise(total_abundance = sum(relative_abundance, na.rm = TRUE)) %>%
  mutate(proportion = total_abundance / sum(total_abundance))

metaphlan4_long %>%
  filter(rank == "species") %>% 
  filter(str_detect(clade_name, "k__Eukaryota|k__Archaea")) %>%
  distinct(clade_name)


#predominant species - sum relabun for each species then sort desc
metaphlan4_wide.s %>% 
  select(!plate_ID) %>% 
  colSums() %>% 
  sort(decreasing = TRUE) %>% 
  head(5)


### Estrobolome Target Coverage

glue("The number of genera-level estrobolome targets: {nrow(estrobolome_targets %>% filter(rank=='genus'))}")
glue("The number of species-level estrobolome targets: {nrow(estrobolome_targets %>% filter(rank=='species'))}")

#PRIOR to data filtering
estrobolome_prefilt <- metaphlan4 %>%  
  wgs_filter(., collection_part, 2) %>% 
  filter(name %in% estrobolome_targets$metaphlan_GMBC) %>%
  distinct(name)

#found
glue("Genera targets found prior to filtering: {nrow(estrobolome_prefilt %>% filter(grepl('g__', name)))}")
glue("Species targets found prior to filtering: {nrow(estrobolome_prefilt %>% filter(grepl('s__', name)))}")
#not found in metaphlan
estrobolome_targets %>% filter(rank=="genus") %>% filter(!metaphlan_GMBC %in% estrobolome_prefilt$name) %>% pull(name)
estrobolome_targets %>% filter(rank=="species") %>% filter(!metaphlan_GMBC %in% estrobolome_prefilt$name) %>% pull(name)


#POST data filtering
estrobolome_postfilt <- metaphlan4_long %>%
  wgs_filter(., collection_part, 2) %>%
  filter(name %in% estrobolome_targets$metaphlan_GMBC) %>%
  distinct(name)

#remaining
glue("Genera targets remaining after filtering: {nrow(estrobolome_postfilt %>% filter(grepl('g__', name)))}")
glue("Species targets remaining after filtering): {nrow(estrobolome_postfilt %>% filter(grepl('s__', name)))}")
#removed
estrobolome_prefilt %>% filter(grepl('g__', name)) %>% filter(!name %in% estrobolome_postfilt$name) %>% pull(name)
estrobolome_prefilt %>% filter(grepl('s__', name)) %>% filter(!name %in% estrobolome_postfilt$name)%>% pull(name)


```

```{r humann pathways info - part 2}

## RAW DATA DESCRIPTION
glue("The max relabun raw data: {max(humann.path_nopipe$relative_abundance)} (unintegrated)")
glue("The min relabun raw data: {min(humann.path_nopipe$relative_abundance)}")
glue("The max relabun raw data (removing unmapped, unintegrated): {
     humann.path_nopipe %>% filter(!name %in% c('UNINTEGRATED','UNMAPPED')) %>% pull(relative_abundance) %>% max()}")
glue("The avg relabun for unintegrated data (raw): {
     mean(humann.path_nopipe %>% filter(name == 'UNINTEGRATED') %>% pull(relative_abundance)) }")
glue("The avg relabun for unmapped data (raw): {
     mean(humann.path_nopipe %>% filter(name == 'UNMAPPED') %>% pull(relative_abundance)) }")
glue("The number of unique pathways in data (raw): {nrow( humann.path_nopipe %>% distinct(name) ) }")


## FILTERED DATA
glue("The max relabun filtered data: {max(humann.path_long$relative_abundance)}")
glue("The min relabun filtered data: {min(humann.path_long$relative_abundance)}")
glue("The number of unique pathways in filtered data: {nrow( humann.path_long %>% distinct(name) ) }")

#predominant pathways - sum relabun for each path then sort desc
humann.path_wide %>% 
  select(!plate_ID) %>% 
  colSums() %>% 
  sort(decreasing = TRUE) %>% 
  head(5)


## Estrobolome Targets

#overall number
glue("The number of pathway estrobolome targets: {nrow(estrobolome_targets %>% filter(!is.na(Pathway_code)))}")

#found in raw data
estrobolome.path_prefilt <- humann.path_nopipe %>%
  filter(sapply(name, function(x) any(str_detect(x, estrobolome_targets$Pathway_code)))) %>%
  distinct(name) %>% 
  filter(!grepl("GALACT", name))
estrobolome.path_prefilt$name

#found in filtered p1 data
estrobolome.path_postfilt <- humann.path_long %>%
  filter(sapply(name, function(x) any(str_detect(x, estrobolome_targets$Pathway_code)))) %>%
  distinct(name) %>% 
  filter(!grepl("GALACT", name))
estrobolome.path_postfilt$name

#those not found
estrobolome_targets %>% 
    filter(!sapply(Pathway_code, function(x) any(str_detect(x, estrobolome.path_prefilt$name)))) %>%
    distinct(Pathway_code, Pathway_name)

```

```{r humann enzymes info - part 2}
## RAW DATA DESCRIPTION

glue("The max relabun raw data: {max(humann.enzymes_nopipe$relative_abundance)}")
glue("The min relabun raw data: {min(humann.enzymes_nopipe$relative_abundance)}")
glue("The number of unique enzymes in raw samples: {nrow( humann.enzymes_nopipe %>% distinct(name) ) }")

## FILTERED DATA
glue("The max relabun filtered data: {max(humann.enzymes_long$relative_abundance)}")
glue("The min relabun filtered data: {min(humann.enzymes_long$relative_abundance)}")
glue("The number of unique enzymes in filtered data: {nrow(humann.enzymes_long %>% distinct(name))}")

#predominant enzymes - sum relabun for each path then sort desc
humann.enzymes_wide %>% 
  select(!plate_ID) %>% 
  colSums() %>% 
  sort(decreasing = TRUE) %>% 
  head(5) %>% 
  as.data.frame() %>% 
  rownames_to_column("name") %>% 
  merge(., humann.enzymes_nopipe %>% select(name, Enzyme_name),
        by = "name") %>% 
  unique() %>% 
  mutate(names = paste0(Enzyme_name," (", name, ")")) %>% 
  select(-name, -Enzyme_name)



## Estrobolome Targets

#overall number
glue("The number of enzyme estrobolome targets: {nrow(estrobolome_targets %>% filter(!is.na(Enzyme_code)))}")

#found in raw p1 data
estrobolome.enzyme_prefilt <- humann.enzymes_nopipe %>%
  filter(name %in% estrobolome_targets$Enzyme_code) %>%
  distinct(name, Enzyme_name)
glue("The number of estrobolome enzyme targets found in baseline raw data: {nrow(estrobolome.enzyme_prefilt)}")
#those found
estrobolome.enzyme_prefilt %>% distinct(Enzyme_name,name) %>% pull(Enzyme_name, name)
#those not found
estrobolome_targets %>% filter(!Enzyme_code %in% estrobolome.enzyme_prefilt$name) %>% distinct(Enzyme_name) %>% pull(Enzyme_name)


#found in filtered p1 data
estrobolome.enzyme_postfilt <- humann.enzymes_long %>%
  filter(name %in% estrobolome_targets$Enzyme_code) %>%
  distinct(name)
glue("The number of estrobolome enzyme targets found in baseline filtered data: {nrow(estrobolome.enzyme_postfilt)}")
#not found
setdiff(estrobolome.enzyme_prefilt$name, estrobolome.enzyme_postfilt$name)

```


### CASE VS CONTROL
 
#### Diversity Measures

Alpha and Beta Diversity

```{r metaphlan alpha diversity}

library(ggpubr)

## Calculating the alpha diversity measures
alpha_data <- alpha_calc_data(metaphlan4_long, metaphlan=TRUE)

## Significant testing
# 
# #cases vs controls at baseline
# alpha_sig_test.groups(alpha_data, col_part=1)
# #cases vs controls at follow up
# alpha_sig_test.groups(alpha_data, col_part=2)
# 
# #cases baseline vs cases follow up
# alpha_sig_test.paired(alpha_data, group=1)
# #controls baseline vs controls follow up
# alpha_sig_test.paired(alpha_data, group=0)



## Plotting Sig Testing SHANNON ####

metaphlan.shannon <- alpha_data %>% 
  #reshape data
  mutate(case_con_sep = ifelse(case_con == 1, "Case", "Control"),
          group_label = case_when(
            case_con == 1 & collection_part == 1 ~ "B_Case",
            case_con == 1 & collection_part == 2 ~ "F_Case",
            case_con == 0 & collection_part == 1 ~ "B_Control",
            case_con == 0 & collection_part == 2 ~ "F_Control"
          )
         ) %>% 
  filter(!is.na(shannon)) %>% #shannon #richness
  #plotting
  ggplot(aes(x = interaction(collection_part, case_con_sep), 
             y = shannon, #shannon #richness
             color = group_label,
             fill = group_label)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      geom_line(stat = "summary", aes(group = case_con_sep), fun = median, color = "black") +  # Connect medians with lines
      scale_x_discrete(labels = c("1.Case" = "Baseline", 
                                  "2.Case" = "Follow-up", 
                                  "1.Control" = "Baseline", 
                                  "2.Control" = "Follow-up")) +
      scale_color_manual(values = c("B_Case" = "#6ba",#"darkblue",
                                    "F_Case" = "darkblue",#"#6ba",
                                    "B_Control" = "wheat3",#"sienna",
                                    "F_Control" = "sienna"# "wheat3"
                                    ),
                         guide = guide_legend(nrow = 2, byrow = TRUE)) +
      scale_fill_manual(values = c("B_Case" = "#6ba",#"darkblue",
                                    "F_Case" = "darkblue",#"#6ba",
                                    "B_Control" = "wheat3",#"sienna",
                                    "F_Control" = "sienna"# "wheat3"
                                    ))+
      theme_minimal() +
      theme(
        legend.position = "none" ,#"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9),
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.y = element_text(size = 10) 
      ) +
      labs(x = "Time Point",
           y = "Alpha Diversity (Shannon)", #Shannon #Richness
           color = "Group")+
  # 🔹 Add p-value brackets
  stat_compare_means(
    comparisons = list(c("1.Case", "2.Case"), c("1.Control", "2.Control")),
    method = "wilcox.test",
    aes(label = sprintf("p = %.3f", ..p..)),   # format to 3 decimals
    size = 2.5,                                 # smaller text
    #label.y = c(4.4, 4.5),                    # adjust upward as needed
    tip.length = 0.01
  )

## Plotting Sig Testing RICHNESS ####

metaphlan.richness <- alpha_data %>% 
  #reshape data
  mutate(case_con_sep = ifelse(case_con == 1, "Case", "Control"),
          group_label = case_when(
            case_con == 1 & collection_part == 1 ~ "B_Case",
            case_con == 1 & collection_part == 2 ~ "F_Case",
            case_con == 0 & collection_part == 1 ~ "B_Control",
            case_con == 0 & collection_part == 2 ~ "F_Control"
          )
         ) %>% 
  filter(!is.na(richness)) %>% #shannon #richness
  #plotting
  ggplot(aes(x = interaction(collection_part, case_con_sep), 
             y = richness, #shannon #richness
             color = group_label)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      geom_line(stat = "summary", aes(group = case_con_sep), fun = median, color = "black") +  # Connect medians with lines
      scale_x_discrete(labels = c("1.Case" = "Baseline", 
                                  "2.Case" = "Follow-up", 
                                  "1.Control" = "Baseline", 
                                  "2.Control" = "Follow-up")) +
      scale_color_manual(values = c("B_Case" = "darkblue",
                                    "F_Case" = "#6ba",
                                    "B_Control" = "sienna",
                                    "F_Control" = "wheat3"),
                         guide = guide_legend(nrow = 2, byrow = TRUE)) +
      theme_minimal() +
      theme(
        legend.position = "none", #"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9)   # makes the x-axis title smaller
      ) +
      labs(x = "Time Point",
           y = "Richness Index", #Shannon #Richness
           color = "Group")

``` 
 
```{r metaphlan beta diversity - significance testing}

## Regular PERMANOVA

#Cases vs controls at baseline
beta_data <- wgs_filter(metaphlan4_wide.t, collection_part, 1)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part),
                by="plate_ID") %>% 
          pull(case_con) %>% 
          factor(levels = c("0","1"))
adonis2(beta_dist ~ groups)$`Pr(>F)`[1]
adonis2(beta_dist ~ groups)$`R2`[1]


#cases vs controls at follow up
beta_data <- wgs_filter(metaphlan4_wide.t, collection_part, 2)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part),
                by="plate_ID") %>% 
          pull(case_con) %>% 
          factor(levels = c("0","1"))
adonis2(beta_dist ~ groups)$`Pr(>F)`[1]
adonis2(beta_dist ~ groups)$`R2`[1]


## PERMANOVA Accounting for Dependent Samples

#cases baseline vs cases follow up
beta_data <- wgs_filter(metaphlan4_wide.t, case_con, 1)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part, single_id),
                by="plate_ID") 

beta_results <- adonis2(beta_dist ~ 
                          groups %>% pull(collection_part) %>% factor(levels = c("1","2")), 
                        permutations = 999,
                        strata = groups$single_id,
                        method = "bray")
beta_results$`Pr(>F)`[1]
beta_results$`R2`[1]

#controls baseline vs controls follow up
beta_data <- wgs_filter(metaphlan4_wide.s, case_con, 0)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part, single_id),
                by="plate_ID") 

beta_results <- adonis2(beta_dist ~ 
                          groups %>% pull(collection_part) %>% factor(levels = c("1","2")), 
                        permutations = 999,
                        strata = groups$single_id,
                        method = "bray")
beta_results$`Pr(>F)`[1]
beta_results$`R2`[1]


```
    
```{r metaphlan beta diversity - plotting overall}

## Calculate Bray-Curtis Distances
dist_mat <- metaphlan4_wide.t %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan


## Principal Component Analysis

beta_pcoa <- cmdscale(dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#convert eigenvalues to percentage
beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)

#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  merge(gmbc_data, ., by="plate_ID") #merge PC values onto meta data



## Plotting

#create combined grouping var
beta_pcoa.df <- beta_pcoa.df %>% 
  mutate(
    group_label = case_when(
      case_con == 1 & collection_part == 1 ~ "Case, Baseline",
      case_con == 1 & collection_part == 2 ~ "Case, Follow-up",
      case_con == 0 & collection_part == 1 ~ "Control, Baseline",
      case_con == 0 & collection_part == 2 ~ "Control, Follow-up"
    )
  )

#calculate the median coordinates for each group
medians <- beta_pcoa.df %>%
  filter(case_con == 1 ) %>% 
  group_by(group_label) %>%
  summarize(
    med_PC1 = median(PC1), 
    med_PC2 = median(PC2),
    .groups = "drop"
  )

#set colours by group
group_colors <- c(
  "Case, Baseline" = "#6ba",#"darkblue",     # medium blue
  "Case, Follow-up" = "darkblue", #"#6ba",    # light blue
  "Control, Baseline" = "wheat3",#"sienna",  # dark grey
  "Control, Follow-up" = "sienna"#"wheat3"  # light grey
)

## Plotting PCOAs
metaphlan.beta_plot <- beta_pcoa.df %>% 
  # PART TWO SAMPLES ONLY
  filter(case_con == 1 ) %>% 
  #plotting
  ggplot(aes(x = PC1, y = PC2, color = group_label, shape = collection_part)) + 
  geom_point(alpha=0.5) +
  stat_ellipse(aes(group = group_label), level = 0.95, linewidth = 0.6) +
  #stat_ellipse(aes(x = PC1, y = PC2), level = 0.95) +
  scale_color_manual(values = group_colors) +# c("0" = "darkgray", "1" = "blue")) +
  scale_shape_manual(values = c("1" = 17, "2" = 16)) + #circle baseline #triangle FU
  labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
       #y = glue("Beta Diversityn/ PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)"),
         y = paste0(
                "<span style='font-size:10pt'>Beta Diversity of Cases</span><br>",
                "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"
              ),
       color = "Case Status",
       shape = "Time Point") + 
  theme(legend.position = "none", #"right",
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9),   # makes the x-axis title smaller
        axis.title.y = element_markdown(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines")) +
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = beta_pcoa.df %>% 
                 left_join(medians, by = "group_label"),
               aes(x = med_PC1, y = med_PC2, xend = PC1, yend = PC2,
                   color = group_label),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_PC1, y = med_PC2, color = group_label), 
             size = 3, shape = 4, stroke = 1.15)   # Shape 4 is an 'x' marker


```
  
  
```{r humann pathways alpha diversity}

## Calculating the alpha diversity measures
alpha_data <- alpha_calc_data(humann.path_long)

## Significant testing

# #cases vs controls at baseline
# alpha_sig_test.groups(alpha_data, col_part=1)
# #cases vs controls at follow up
# alpha_sig_test.groups(alpha_data, col_part=2)
# 
# #cases baseline vs cases follow up
# alpha_sig_test.paired(alpha_data, group=1)
# #controls baseline vs controls follow up
# alpha_sig_test.paired(alpha_data, group=0)



## Plotting Sig Testing SHANNON ####

path.shannon <- alpha_data %>% 
  #reshape data
  mutate(case_con_sep = ifelse(case_con == 1, "Case", "Control"),
          group_label = case_when(
            case_con == 1 & collection_part == 1 ~ "B_Case",
            case_con == 1 & collection_part == 2 ~ "F_Case",
            case_con == 0 & collection_part == 1 ~ "B_Control",
            case_con == 0 & collection_part == 2 ~ "F_Control"
          )
         ) %>% 
  filter(!is.na(shannon)) %>% #shannon #richness
  #plotting
  ggplot(aes(x = interaction(collection_part, case_con_sep), 
             y = shannon, #shannon #richness
             color = group_label,
             fill = group_label)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      geom_line(stat = "summary", aes(group = case_con_sep), fun = median, color = "black") +  # Connect medians with lines
      scale_x_discrete(labels = c("1.Case" = "Baseline", 
                                  "2.Case" = "Follow-up", 
                                  "1.Control" = "Baseline", 
                                  "2.Control" = "Follow-up")) +
      scale_color_manual(values = c("B_Case" = "#6ba",#"darkblue",
                                    "F_Case" = "darkblue",#"#6ba",
                                    "B_Control" = "wheat3",#"sienna",
                                    "F_Control" = "sienna"# "wheat3"
                                    ),
                         guide = guide_legend(nrow = 2, byrow = TRUE)) +
      scale_fill_manual(values = c("B_Case" = "#6ba",#"darkblue",
                                    "F_Case" = "darkblue",#"#6ba",
                                    "B_Control" = "wheat3",#"sienna",
                                    "F_Control" = "sienna"# "wheat3"
                                    ))+
      theme_minimal() +
      theme(
        legend.position = "none" ,#"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9),
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.y = element_text(size = 10) 
      ) +
      labs(x = "Time Point",
           y = "Alpha Diversity (Shannon)", #Shannon #Richness
           color = "Group") +
  # 🔹 Add p-value brackets
  stat_compare_means(
    comparisons = list(c("1.Case", "2.Case"), c("1.Control", "2.Control")),
    method = "wilcox.test",
    aes(label = sprintf("p = %.3f", ..p..)),   # format to 3 decimals
    size = 2.5,                                 # smaller text
    label.y = c(5.6, 5.65),                    # adjust upward as needed
    tip.length = 0.01
  )

## Plotting Sig Testing RICHNESS ####

path.richness <- alpha_data %>% 
  #reshape data
  mutate(case_con_sep = ifelse(case_con == 1, "Case", "Control"),
          group_label = case_when(
            case_con == 1 & collection_part == 1 ~ "B_Case",
            case_con == 1 & collection_part == 2 ~ "F_Case",
            case_con == 0 & collection_part == 1 ~ "B_Control",
            case_con == 0 & collection_part == 2 ~ "F_Control"
          )
         ) %>% 
  filter(!is.na(richness)) %>% #shannon #richness
  #plotting
  ggplot(aes(x = interaction(collection_part, case_con_sep), 
             y = richness, #shannon #richness
             color = group_label)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      geom_line(stat = "summary", aes(group = case_con_sep), fun = median, color = "black") +  # Connect medians with lines
      scale_x_discrete(labels = c("1.Case" = "Baseline", 
                                  "2.Case" = "Follow-up", 
                                  "1.Control" = "Baseline", 
                                  "2.Control" = "Follow-up")) +
      scale_color_manual(values = c("B_Case" = "darkblue",
                                    "F_Case" = "#6ba",
                                    "B_Control" = "sienna",
                                    "F_Control" = "wheat3"),
                         guide = guide_legend(nrow = 2, byrow = TRUE)) +
      theme_minimal() +
      theme(
        legend.position = "none", #"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9)   # makes the x-axis title smaller
      ) +
      labs(x = "Time Point",
           y = "Richness Index", #Shannon #Richness
           color = "Group")


```

```{r humann.path beta diversity - significance testing}

## Regular PERMANOVA

#Cases vs controls at baseline
beta_data <- wgs_filter(humann.path_wide, collection_part, 1)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part),
                by="plate_ID") %>% 
          pull(case_con) %>% 
          factor(levels = c("0","1"))
adonis2(beta_dist ~ groups)$`Pr(>F)`[1]
adonis2(beta_dist ~ groups)$`R2`[1]


#cases vs controls at follow up
beta_data <- wgs_filter(humann.path_wide, collection_part, 2)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part),
                by="plate_ID") %>% 
          pull(case_con) %>% 
          factor(levels = c("0","1"))
adonis2(beta_dist ~ groups)$`Pr(>F)`[1]
adonis2(beta_dist ~ groups)$`R2`[1]


## PERMANOVA Accounting for Dependent Samples

#cases baseline vs cases follow up
beta_data <- wgs_filter(humann.path_wide, case_con, 1)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part, single_id),
                by="plate_ID") 

beta_results <- adonis2(beta_dist ~ 
                          groups %>% pull(collection_part) %>% factor(levels = c("1","2")), 
                        permutations = 999,
                        strata = groups$single_id,
                        method = "bray")
beta_results$`Pr(>F)`[1]
beta_results$`R2`[1]

#controls baseline vs controls follow up
beta_data <- wgs_filter(humann.path_wide, case_con, 0)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part, single_id),
                by="plate_ID") 

beta_results <- adonis2(beta_dist ~ 
                          groups %>% pull(collection_part) %>% factor(levels = c("1","2")), 
                        permutations = 999,
                        strata = groups$single_id,
                        method = "bray")
beta_results$`Pr(>F)`[1]
beta_results$`R2`[1]


```
 
```{r humann.path beta diversity - plotting overall}

## Calculate Bray-Curtis Distances
dist_mat <- humann.path_wide %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan


## Principal Component Analysis

beta_pcoa <- cmdscale(dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#convert eigenvalues to percentage
beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)

#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  merge(gmbc_data, ., by="plate_ID") #merge PC values onto meta data



## Plotting

#create combined grouping var
beta_pcoa.df <- beta_pcoa.df %>% 
  mutate(
    group_label = case_when(
      case_con == 1 & collection_part == 1 ~ "Case, Baseline",
      case_con == 1 & collection_part == 2 ~ "Case, Follow-up",
      case_con == 0 & collection_part == 1 ~ "Control, Baseline",
      case_con == 0 & collection_part == 2 ~ "Control, Follow-up"
    )
  )

#calculate the median coordinates for each group
medians <- beta_pcoa.df %>%
  filter(case_con == 1 ) %>% 
  group_by(group_label) %>%
  summarize(
    med_PC1 = median(PC1), 
    med_PC2 = median(PC2),
    .groups = "drop"
  )

#set colours by group
group_colors <- c(
  "Case, Baseline" = "#6ba",#"darkblue",     # medium blue
  "Case, Follow-up" = "darkblue", #"#6ba",    # light blue
  "Control, Baseline" = "wheat3",#"sienna",  # dark grey
  "Control, Follow-up" = "sienna"#"wheat3"  # light grey
)

## Plotting PCOAs
path.beta_plot <- beta_pcoa.df %>% 
  filter(case_con == 1 ) %>% 
  ggplot(aes(x = PC1, y = PC2, color = group_label, shape = collection_part)) + 
  geom_point(alpha=0.5) +
  stat_ellipse(aes(group = group_label), level = 0.95, linewidth = 0.6) +
  #stat_ellipse(aes(x = PC1, y = PC2), level = 0.95) +
  scale_color_manual(values = group_colors) +# c("0" = "darkgray", "1" = "blue")) +
  scale_shape_manual(values = c("1" = 17, "2" = 16)) + #circle baseline #triangle FU
  labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
       #y = glue("PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)"),
         y = paste0(
                "<span style='font-size:10pt'>Beta Diversity of Cases</span><br>",
                "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"
              ),
       color = "Case Status",
       shape = "Time Point") + 
  theme(legend.position = "none",#"right",
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9),   # makes the x-axis title smaller
        axis.title.y = element_markdown(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines")) +
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = beta_pcoa.df %>% 
                 left_join(medians, by = "group_label"),
               aes(x = med_PC1, y = med_PC2, xend = PC1, yend = PC2,
                   color = group_label),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_PC1, y = med_PC2, color = group_label), 
             size = 3, shape = 4, stroke = 1.15)  # Shape 4 is an 'x' marker

```
 
        
```{r humann enzymes alpha diversity}

## Calculating the alpha diversity measures
alpha_data <- alpha_calc_data(humann.enzymes_long)

## Significant testing

#cases vs controls at baseline
# alpha_sig_test.groups(alpha_data, col_part=1)
# #cases vs controls at follow up
# alpha_sig_test.groups(alpha_data, col_part=2)
# 
# #cases baseline vs cases follow up
# alpha_sig_test.paired(alpha_data, group=1)
# #controls baseline vs controls follow up
# alpha_sig_test.paired(alpha_data, group=0)


## Plotting Sig Testing SHANNON ####

enzyme.shannon <- alpha_data %>% 
  #reshape data
  mutate(case_con_sep = ifelse(case_con == 1, "Case", "Control"),
          group_label = case_when(
            case_con == 1 & collection_part == 1 ~ "B_Case",
            case_con == 1 & collection_part == 2 ~ "F_Case",
            case_con == 0 & collection_part == 1 ~ "B_Control",
            case_con == 0 & collection_part == 2 ~ "F_Control"
          )
         ) %>% 
  filter(!is.na(shannon)) %>% #shannon #richness
  #plotting
  ggplot(aes(x = interaction(collection_part, case_con_sep), 
             y = shannon, #shannon #richness
             color = group_label,
             fill = group_label)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      geom_line(stat = "summary", aes(group = case_con_sep), fun = median, color = "black") +  # Connect medians with lines
      scale_x_discrete(labels = c("1.Case" = "Baseline", 
                                  "2.Case" = "Follow-up", 
                                  "1.Control" = "Baseline", 
                                  "2.Control" = "Follow-up")) +
      scale_color_manual(values = c("B_Case" = "#6ba",#"darkblue",
                                    "F_Case" = "darkblue",#"#6ba",
                                    "B_Control" = "wheat3",#"sienna",
                                    "F_Control" = "sienna"# "wheat3"
                                    ),
                         guide = guide_legend(nrow = 2, byrow = TRUE)) +
      scale_fill_manual(values = c("B_Case" = "#6ba",#"darkblue",
                                    "F_Case" = "darkblue",#"#6ba",
                                    "B_Control" = "wheat3",#"sienna",
                                    "F_Control" = "sienna"# "wheat3"
                                    ))+
      theme_minimal() +
      theme(
        legend.position = "none" ,#"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9),
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.y = element_text(size = 10) 
      ) +
      labs(x = "Time Point",
           y = "Alpha Diversity (Shannon)", #Shannon #Richness
           color = "Group") +
  # 🔹 Add p-value brackets
  stat_compare_means(
    comparisons = list(c("1.Case", "2.Case"), c("1.Control", "2.Control")),
    method = "wilcox.test",
    aes(label = sprintf("p = %.3f", ..p..)),   # format to 3 decimals
    size = 2.5,                                 # smaller text
    label.y = c(6.55, 6.65),                    # adjust upward as needed
    tip.length = 0.01
  )

## Plotting Sig Testing RICHNESS ####

enzyme.richness <- alpha_data %>% 
  #reshape data
  mutate(case_con_sep = ifelse(case_con == 1, "Case", "Control"),
          group_label = case_when(
            case_con == 1 & collection_part == 1 ~ "B_Case",
            case_con == 1 & collection_part == 2 ~ "F_Case",
            case_con == 0 & collection_part == 1 ~ "B_Control",
            case_con == 0 & collection_part == 2 ~ "F_Control"
          )
         ) %>% 
  filter(!is.na(richness)) %>% #shannon #richness
  #plotting
  ggplot(aes(x = interaction(collection_part, case_con_sep), 
             y = richness, #shannon #richness
             color = group_label)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      geom_line(stat = "summary", aes(group = case_con_sep), fun = median, color = "black") +  # Connect medians with lines
      scale_x_discrete(labels = c("1.Case" = "Baseline", 
                                  "2.Case" = "Follow-up", 
                                  "1.Control" = "Baseline", 
                                  "2.Control" = "Follow-up")) +
      scale_color_manual(values = c("B_Case" = "darkblue",
                                    "F_Case" = "#6ba",
                                    "B_Control" = "sienna",
                                    "F_Control" = "wheat3"),
                         guide = guide_legend(nrow = 2, byrow = TRUE)) +
      theme_minimal() +
      theme(
        legend.position = "none", #"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 9)   # makes the x-axis title smaller
      ) +
      labs(x = "Time Point",
           y = "Richness Index", #Shannon #Richness
           color = "Group")

```

```{r humann.enzyme beta diversity - significance testing}

## Regular PERMANOVA

#Cases vs controls at baseline
beta_data <- wgs_filter(humann.enzymes_wide, collection_part, 1)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part),
                by="plate_ID") %>% 
          pull(case_con) %>% 
          factor(levels = c("0","1"))
adonis2(beta_dist ~ groups)$`Pr(>F)`[1]
adonis2(beta_dist ~ groups)$`R2`[1]


#cases vs controls at follow up
beta_data <- wgs_filter(humann.enzymes_wide, collection_part, 2)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part),
                by="plate_ID") %>% 
          pull(case_con) %>% 
          factor(levels = c("0","1"))
adonis2(beta_dist ~ groups)$`Pr(>F)`[1]
adonis2(beta_dist ~ groups)$`R2`[1]


## PERMANOVA Accounting for Dependent Samples

#cases baseline vs cases follow up
beta_data <- wgs_filter(humann.enzymes_wide, case_con, 1)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part, single_id),
                by="plate_ID") 

beta_results <- adonis2(beta_dist ~ 
                          groups %>% pull(collection_part) %>% factor(levels = c("1","2")), 
                        permutations = 999,
                        strata = groups$single_id,
                        method = "bray")
beta_results$`Pr(>F)`[1]
beta_results$`R2`[1]

#controls baseline vs controls follow up
beta_data <- wgs_filter(humann.enzymes_wide, case_con, 0)

beta_dist <- beta_data %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

groups <- merge(beta_data %>% select(plate_ID),
                gmbc_data %>% select(plate_ID, case_con, collection_part, single_id),
                by="plate_ID") 

beta_results <- adonis2(beta_dist ~ 
                          groups %>% pull(collection_part) %>% factor(levels = c("1","2")), 
                        permutations = 999,
                        strata = groups$single_id,
                        method = "bray")
beta_results$`Pr(>F)`[1]
beta_results$`R2`[1]


```
   
```{r humann.enzyme beta diversity - plotting overall}

## Calculate Bray-Curtis Distances
dist_mat <- humann.enzymes_wide %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan


## Principal Component Analysis

beta_pcoa <- cmdscale(dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#convert eigenvalues to percentage
beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)

#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  merge(gmbc_data, ., by="plate_ID") #merge PC values onto meta data



## Plotting

#calculate the median coordinates for each group
medians <- beta_pcoa.df %>%
  filter(case_con == 1) %>% 
  group_by(collection_part) %>%
  summarize(
    med_PC1 = median(PC1), 
    med_PC2 = median(PC2),
    .groups = "drop"
  )

## Plotting PCOAs
enzyme.beta_plot <- beta_pcoa.df %>% 
  filter(case_con == 1) %>% 
  ggplot(aes(x = PC1, y = PC2, color = collection_part, shape = collection_part)) + 
  geom_point(alpha=0.5, stroke = 0, size = 2) +
  stat_ellipse(aes(group = collection_part), level = 0.95, linewidth = 0.6) +
  scale_shape_manual(values = c("1" = 17, "2" = 16), guide = "none") + #circle baseline #triangle FU
  scale_color_manual(
    name = "Time Point",           # legend title
    values = c("1" = "#6ba", "2" = "darkblue"),
    labels = c("1" = "Baseline", "2" = "Follow-up")
  ) +
  labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
       #y = glue("PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)")
      y = paste0(
      "<span style='font-size:10pt'>Beta Diversity of Cases</span><br>",
      "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"
    ),
       ) + 
  ylim(-0.2, NA) +
  theme(
    axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
    axis.title.x = element_text(size = 9),   # makes the x-axis title smaller
    axis.title.y = element_markdown(),
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines"),
        #LEGEND
        legend.text = element_text(size = 6.8),
        legend.title = element_text(size = 8),
        legend.position = c(0.99, 0.01),    # inside top left (x,y in npc)
        legend.box = "horizontal",
        legend.justification = c("right", "bottom"),
        legend.background = element_rect(color = "black", fill = "white"),
        legend.key.size = unit(0.5, "lines"),
        legend.spacing.y = unit(0.1, "cm"), 
        legend.spacing.x = unit(0.1, "cm"),
        legend.margin = margin(t = 0.1, r = 0.1, b = 0.1, l = 0.1, unit = "cm")
       ) +
  #add lines from each point to the median of its group with matching colors
  geom_segment(
    data = beta_pcoa.df %>% 
           filter(case_con == 1) %>%  # only follow-up points
           left_join(medians, by = "collection_part"),
    aes(x = med_PC1, y = med_PC2, xend = PC1, yend = PC2, color = collection_part),
    linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE
  ) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_PC1, y = med_PC2, color = collection_part), 
             size = 3, shape = 4, stroke = 1.15) 


```


Temporal Changes in Bray-Curtis Dissimilarities
        
```{r metaphlan beta diversity - intra-individual dissimilarity}

### Calculate Bray-Curtis Distances
dist_mat <- metaphlan4_wide.t %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

### DATA PREPARATION

# Convert distance object to full matrix
beta_matrix <- as.matrix(dist_mat)


# Create a long-form dataframe of all pairwise distances

#Get all pairwise combinations
pairwise_df <- expand.grid(sample1 = rownames(beta_matrix), sample2 = colnames(beta_matrix)) %>%
    mutate(across(everything(), as.character)) %>%  # convert both to character
    filter(sample1 < sample2)  # avoid duplicates and diagonals

# Add distance values
pairwise_df$distance <- mapply(function(x, y) beta_matrix[x, y], 
                                pairwise_df$sample1, pairwise_df$sample2)


# Add subject IDs, collection_part, case_con for each sample
pairwise_df <- pairwise_df %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con), by = c("sample1" = "plate_ID")) %>%
  rename(singleid1 = single_id,
         time1 = collection_part,
         case_con1 = case_con
         ) %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con), by = c("sample2" = "plate_ID")) %>%
  rename(singleid2 = single_id,
         time2 = collection_part,
         case_con2 = case_con
         )



### CALCULATE WITHIN: Intra-Individual Dissimilarity (microbiome stability within indv over time)

# Filter out where ids match but different time points (data is symmetric)
intra_individual_df <- pairwise_df %>% 
  filter(singleid1 == singleid2, time1 != time2)

# Convert case_con to factor for nicer labels
intra_individual_df$case_con1 <- factor(intra_individual_df$case_con1, levels = c(1,0), labels = c("Cases", "Controls"))

# Plot boxplot
metaphlan.intraplot <- ggplot(intra_individual_df, aes(x = case_con1, y = distance, fill = case_con1, color=case_con1)) +
  geom_point(position = position_jitter(width = 0.15), size = 2, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  labs(x = NULL, #"Species", 
       y = "Intra-individual Dissimilarity") + #Bray-Curtis 
  scale_fill_manual(values = c("Controls" = "sienna", "Cases" = "darkblue")) +
  scale_color_manual(values = c("Controls" = "sienna", "Cases" = "darkblue")) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 10),
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.y = element_text(size = 10) 
        ) +
  stat_compare_means(
    comparisons = list(c("Controls", "Cases")),
    method = "wilcox.test",    # or "t.test"
    aes(label = sprintf("p = %.3f", ..p..)),  # three decimals
    tip.length = 0.01,
    size = 2.5  
  )

metaphlan.intraplot
# Statistical comparison
wilcox.test(distance ~ case_con1, data = intra_individual_df)

# Median and IQR values

intra_individual_df %>%
  group_by(case_con1) %>%
  summarise(
    n = n(),
    median = median(distance),
    IQR = IQR(distance),
    q25 = quantile(distance, 0.25),
    q75 = quantile(distance, 0.75)
  )



```
   
```{r humann.path beta diversity - intra-individual dissimilarity}

### Calculate Bray-Curtis Distances
dist_mat <- humann.path_wide %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

### DATA PREPARATION

# Convert distance object to full matrix
beta_matrix <- as.matrix(dist_mat)


# Create a long-form dataframe of all pairwise distances

#Get all pairwise combinations
pairwise_df <- expand.grid(sample1 = rownames(beta_matrix), sample2 = colnames(beta_matrix)) %>%
    mutate(across(everything(), as.character)) %>%  # convert both to character
    filter(sample1 < sample2)  # avoid duplicates and diagonals

# Add distance values
pairwise_df$distance <- mapply(function(x, y) beta_matrix[x, y], 
                                pairwise_df$sample1, pairwise_df$sample2)


# Add subject IDs, collection_part, case_con for each sample
pairwise_df <- pairwise_df %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con), by = c("sample1" = "plate_ID")) %>%
  rename(singleid1 = single_id,
         time1 = collection_part,
         case_con1 = case_con
         ) %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con), by = c("sample2" = "plate_ID")) %>%
  rename(singleid2 = single_id,
         time2 = collection_part,
         case_con2 = case_con
         )



### CALCULATE WITHIN: Intra-Individual Dissimilarity (microbiome stability within indv over time)

# Filter out where ids match but different time points (data is symmetric)
intra_individual_df <- pairwise_df %>% 
  filter(singleid1 == singleid2, time1 != time2)

# Convert case_con to factor for nicer labels
intra_individual_df$case_con1 <- factor(intra_individual_df$case_con1, levels = c(1,0), labels = c("Cases", "Controls"))

# Plot boxplot
path.intraplot <- ggplot(intra_individual_df, aes(x = case_con1, y = distance, fill = case_con1, color=case_con1)) +
  geom_point(position = position_jitter(width = 0.15), size = 2, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  labs(x = NULL, #"Pathways", 
       y = "Intra-individual Dissimilarity") + #"Intra-individual Bray-Curtis Dissimilarity"
  scale_fill_manual(values = c("Controls" = "sienna", "Cases" = "darkblue")) +
  scale_color_manual(values = c("Controls" = "sienna", "Cases" = "darkblue")) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 10),
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.y = element_text(size = 10) 
        ) +
  stat_compare_means(
    comparisons = list(c("Controls", "Cases")),
    method = "wilcox.test",    # or "t.test"
    aes(label = sprintf("p = %.3f", ..p..)),  # three decimals
    tip.length = 0.01,
    size = 2.5  
  )
  # annotate(
  #   "text", x = 1.5,
  #   y = min(intra_individual_df$distance) + 0.00,  # adjust Y as needed
  #   label = "p = 0.168",
  #   size = 3, fontface = "italic"
  # ) 
path.intraplot

# Statistical comparison
wilcox.test(distance ~ case_con1, data = intra_individual_df)

# Median and IQR values

intra_individual_df %>%
  group_by(case_con1) %>%
  summarise(
    n = n(),
    median = median(distance),
    IQR = IQR(distance),
    q25 = quantile(distance, 0.25),
    q75 = quantile(distance, 0.75)
  )



```
  
```{r humann.enzyme beta diversity - intra-individual dissimilarity}
library(ggpubr)

### Calculate Bray-Curtis Distances
dist_mat <- humann.enzymes_wide %>% 
    column_to_rownames("plate_ID") %>% #turning column values to name
    as.matrix() %>% #convert to matrix
    vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

### DATA PREPARATION

# Convert distance object to full matrix
beta_matrix <- as.matrix(dist_mat)


# Create a long-form dataframe of all pairwise distances

#Get all pairwise combinations
pairwise_df <- expand.grid(sample1 = rownames(beta_matrix), sample2 = colnames(beta_matrix)) %>%
    mutate(across(everything(), as.character)) %>%  # convert both to character
    filter(sample1 < sample2)  # avoid duplicates and diagonals

# Add distance values
pairwise_df$distance <- mapply(function(x, y) beta_matrix[x, y], 
                                pairwise_df$sample1, pairwise_df$sample2)


# Add subject IDs, collection_part, case_con for each sample
pairwise_df <- pairwise_df %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con), by = c("sample1" = "plate_ID")) %>%
  rename(singleid1 = single_id,
         time1 = collection_part,
         case_con1 = case_con
         ) %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, case_con), by = c("sample2" = "plate_ID")) %>%
  rename(singleid2 = single_id,
         time2 = collection_part,
         case_con2 = case_con
         )



### CALCULATE WITHIN: Intra-Individual Dissimilarity (microbiome stability within indv over time)

# Filter out where ids match but different time points (data is symmetric)
intra_individual_df <- pairwise_df %>% 
  filter(singleid1 == singleid2, time1 != time2)

# Convert case_con to factor for nicer labels
intra_individual_df$case_con1 <- factor(intra_individual_df$case_con1, levels = c(1,0), labels = c("Cases", "Controls"))

# Plot boxplot
enzyme.intraplot <- ggplot(intra_individual_df, aes(x = case_con1, y = distance, fill = case_con1, color=case_con1)) +
  geom_point(position = position_jitter(width = 0.15), size = 2, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  labs(x = NULL, #"Enzymes", 
       y = "Intra-individual Dissimilarity") +#"Intra-individual Bray-Curtis Dissimilarity"
  scale_fill_manual(values = c("Controls" = "sienna", "Cases" = "darkblue")) +
  scale_color_manual(values = c("Controls" = "sienna", "Cases" = "darkblue")) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.x = element_text(size = 10),
        axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
        axis.title.y = element_text(size = 10) 
        )  +
  stat_compare_means(
    comparisons = list(c("Controls", "Cases")),
    method = "wilcox.test",    # or "t.test"
    aes(label = sprintf("p = %.3f", ..p..)),  # three decimals
    tip.length = 0.01,
    size = 2.5  
  )
  # annotate(
  #   "text", x = 1.5,
  #   y = 1 - min(intra_individual_df$distance) + 0.00,  # adjust Y as needed
  #   label = "p = 0.078",
  #   size = 3, fontface = "italic"
  # )
enzyme.intraplot

# Statistical comparison
wilcox.test(distance ~ case_con1, data = intra_individual_df)

# Median and IQR values

intra_individual_df %>%
  group_by(case_con1) %>%
  summarise(
    n = n(),
    median = median(distance),
    IQR = IQR(distance),
    q25 = quantile(distance, 0.25),
    q75 = quantile(distance, 0.75)
  )



```

Combined Plots

```{r COMBINED PLOTTING }

library(cowplot) 
library(grid)

png("microbiome_overall2_2025-10-14.png", 
    width = 6.5 * 600, height = 7 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #row 1 - alpha diversity measures
    plot_grid(
      metaphlan.shannon, metaphlan.intraplot, metaphlan.beta_plot, 
      nrow = 1,
      ncol = 3,
      rel_widths = c(1.1,0.7,1.3)
    ),
    #row 3 - Bray-curtis dissimilaries
    plot_grid(
           path.shannon, path.intraplot, path.beta_plot,
      #arguments
      nrow = 1,
      ncol = 3,
      rel_widths = c(1.1,0.7,1.2)
    ),
    #row 2 - beta diversity
    plot_grid(
       enzyme.shannon, enzyme.intraplot, enzyme.beta_plot,
      nrow = 1,
      ncol = 3,
      rel_widths = c(1.1,0.7,1.25)
    ),
    #arguments
    nrow = 3,
    ncol = 1,
    rel_heights = c(1,1,1)
  )
)


grid.text("Species", x = unit(0.01, "npc"), y = unit(0.987, "npc"), #0.099 x
          just = "left",  
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Pathways", x = unit(0.01, "npc"), y = unit(0.675, "npc"), #0.348
          just = "left",  
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Enzymes", x = unit(0.01, "npc"), y = unit(0.33, "npc"),
          just = "left",  
          gp = gpar(fontsize = 15, fontface = "bold"))

# grid.text("Cases", x = unit(0.15, "npc"), y = unit(0.993, "npc"), #0.099 x
#           gp = gpar(fontsize = 9))
# 
# grid.text("Controls", x = unit(0.305, "npc"), y = unit(0.993, "npc"), #0.099 x
#           gp = gpar(fontsize = 9))


dev.off()

```



#### GLMMs
 
Running Models
```{r GLMMs - genera}

apply_iterative.modeling(
  predictor_vars = metaphlan4_wide.g %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = metaphlan4_wide.g %>% transform_metagenomics(),
  file_name = "genus_long"
)

```

```{r GLMMs - species}

apply_iterative.modeling(
  predictor_vars = metaphlan4_wide.t %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = metaphlan4_wide.t %>% transform_metagenomics(),
  file_name = "species_long"
)

```

```{r GLMMs - enzymes}

apply_iterative.modeling(
  predictor_vars = humann.enzymes_wide %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = humann.enzymes_wide %>% transform_metagenomics(),
  file_name = "enzymes_long"
)

```
 
```{r GLMMs - pathways}
 
apply_iterative.modeling(
  predictor_vars = humann.path_wide %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = humann.path_wide %>% transform_metagenomics(),
  file_name = "pathways_long"
)

```
 
Volcano Plots
```{r GLMM genera - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_genus_long_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = if_else(p.value < 0.05, "1", "2")
    )

## Run plot
volcano.genus <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(color = sig_color), alpha = 0.4, size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
  #geom_hline(yintercept = Inf, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey70") +
  scale_color_manual(values = c("1" = "darkblue", "2" = "grey70")) +
  scale_x_continuous(limits = c(-2.2, 2)) +
  scale_y_continuous(limits = c(0, 3.1)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Genera"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

#volcano.genus

```
 
```{r GLMM species - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_species_long_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = if_else(p.value < 0.05, "1", "2")
    )

## Run plot
volcano.species <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(color = sig_color), alpha = 0.4, size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
  #geom_hline(yintercept = Inf, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray70") +
  scale_color_manual(values = c("1" = "darkblue", "2" = "grey70")) +
  scale_x_continuous(limits = c(-2.2, 2)) +
  scale_y_continuous(limits = c(0, 3.1)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

volcano.species

```

```{r GLMM pathways - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_pathways_long_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = if_else(p.value < 0.05, "1", "2")
    )

## Run plot
volcano.path <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(color = sig_color), alpha = 0.4, size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
  #geom_hline(yintercept = Inf, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray70") +
  scale_color_manual(values = c("1" = "darkblue", "2" = "grey70")) +
  scale_x_continuous(limits = c(-2.2, 2)) +
  scale_y_continuous(limits = c(0, 3.1)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Pathways"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

#volcano.path

```
 
```{r GLMM enzymes - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_enzymes_long_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = if_else(p.value < 0.05, "1", "2")
    )

## Run plot
volcano.enzyme <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(color = sig_color), alpha = 0.4, size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
  #geom_hline(yintercept = Inf, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray70") +
  scale_color_manual(values = c("1" = "darkblue", "2" = "grey70")) +
  scale_x_continuous(limits = c(-2.2, 2)) +
  scale_y_continuous(limits = c(0, 3.1)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Enzymes"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

#volcano.enzyme

```
 
 
```{r COMBINED - volcano plot}
 
 
library(cowplot)

png("GLMMs_volcano_2025-11-10.png", 
    width = 6.5 * 600, height = 8.5 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #plots
    volcano.genus, volcano.species, 
    volcano.path, volcano.enzyme,
    volcano.plasma, volcano.stool,
    #arguments
    nrow = 3,
    ncol = 2,
    rel_widths = c(1,1),
    rel_heights = c(1,1,1)
  )
)

#grid.text("Species", x = unit(0.45, "npc"), y = unit(0.985, "npc"), #0.099 x
          #gp = gpar(fontsize = 15, fontface = "bold"))


dev.off()

```
 

Relative Abundance Plot

```{r Relative abundance plots}

# Relevant Features
features <- c(
    "g__Christensenellaceae_unclassified", 
    "g__Faecalicatena",
    "s__Bifidobacterium_adolescentis|t__SGB17244_group", 
    "s__Coprococcus_eutactus|t__SGB5117",
    "s__Intestinibacter_bartlettii|t__SGB6140", 
    "1.1.1.64",
    "3.4.13.21",
    "3.5.1.82",
    "PWY-5850: superpathway of menaquinol-6 biosynthesis",
    "PWY-5860: superpathway of demethylmenaquinol-6 biosynthesis I",
    "PWY-5896: superpathway of menaquinol-10 biosynthesis"
  )


# select from each datatype
estro.t <- metaphlan4_wide.t %>% select(plate_ID, any_of(features))
estro.g <- metaphlan4_wide.g %>% select(plate_ID, any_of(features))
estro.e <- humann.enzymes_wide %>% select(plate_ID, any_of(features))
estro.p <- humann.path_wide %>% select(plate_ID, any_of(features))
#merge all together
estro.full <- estro.t %>% 
  left_join(estro.g, by="plate_ID") %>% 
  left_join(estro.e, by="plate_ID") %>% 
  left_join(estro.p, by="plate_ID")
    
# Prep data
estro.full <- estro.full %>% 
  #clean up names
  rename(
    "Christensenellaceae unclass." = "g__Christensenellaceae_unclassified", 
    "Faecalicatena" = "g__Faecalicatena",
    "Bifidobacterium adolescentis" = "s__Bifidobacterium_adolescentis|t__SGB17244_group", 
    "Coprococcus eutactus" = "s__Coprococcus_eutactus|t__SGB5117", 
    "Intestinibacter bartlettii" = "s__Intestinibacter_bartlettii|t__SGB6140", 
    "Testosterone dehyd. (1.1.1.64)"="1.1.1.64",
    "Dipeptidase E (3.4.13.21)" = "3.4.13.21",
    "N-a-D-G deacylase (3.5.1.82)" = "3.5.1.82",
    "menaquinol-6 B (5850)" = "PWY-5850: superpathway of menaquinol-6 biosynthesis",
    "demethylmenaquinol-6 B (5860)" = "PWY-5860: superpathway of demethylmenaquinol-6 biosynthesis I",
    "menaquinol-10 B (5896)" = "PWY-5896: superpathway of menaquinol-10 biosynthesis"
  ) %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con, collection_part), by="plate_ID") %>% 
  mutate(
    group_label = case_when(
      case_con == 1 & collection_part == 1 ~ "Case, Baseline",
      case_con == 1 & collection_part == 2 ~ "Case, Follow-up",
      case_con == 0 & collection_part == 1 ~ "Control, Baseline",
      case_con == 0 & collection_part == 2 ~ "Control, Follow-up"
    )
  ) %>% 
  #long form data
  pivot_longer(
    cols = -c(plate_ID, group_label, case_con, collection_part),     # which columns to reshape
    names_to = "Feature",              # new column for old column names
    values_to = "relative_abundance"             # new column for values
  ) %>% 
  #remove zeros so not clogging the plot
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance))

## Colours
color_values = c(
      "Case, Baseline" = "#6ba",
      "Case, Follow-up" = "darkblue",
      "Control, Baseline" = "wheat3",
      "Control, Follow-up" = "sienna"
    )

## Order for the plot
estro.full <- estro.full %>% 
  mutate(Feature = factor(Feature, levels = 
      c("Christensenellaceae unclass.", "Faecalicatena",
        "Bifidobacterium adolescentis", "Coprococcus eutactus","Intestinibacter bartlettii",
        "Testosterone dehyd. (1.1.1.64)","Dipeptidase E (3.4.13.21)","N-a-D-G deacylase (3.5.1.82)",
        "menaquinol-6 B (5850)","demethylmenaquinol-6 B (5860)","menaquinol-10 B (5896)"
        )
   ))

## Beta value
estro.full <- estro.full %>% 
  mutate(beta_coef = case_when(
    Feature == "Christensenellaceae unclass." ~ "β = 1.79", 
    Feature == "Faecalicatena" ~ "β = -1.64",
    Feature == "Bifidobacterium adolescentis" ~ "β = -2.15",
    Feature == "Coprococcus eutactus" ~ "β = -0.85",
    Feature == "Intestinibacter bartlettii" ~ "β = 1.39",
    Feature == "Testosterone dehyd. (1.1.1.64)" ~ "β = -0.88",
    Feature == "Dipeptidase E (3.4.13.21)" ~ "β = -1.81",
    Feature == "N-a-D-G deacylase (3.5.1.82)" ~ "β = 1.71",
    Feature == "menaquinol-6 B (5850)" ~ "β = -1.33",
    Feature == "demethylmenaquinol-6 B (5860)" ~ "β = -1.26",
    Feature == "menaquinol-10 B (5896)" ~ "β = -1.25",
    TRUE ~ NA ))

## Beta value positions
B_position <- estro.full %>% 
  group_by(Feature, beta_coef) %>% 
  summarize(x_pos = max(log_rel_abundance, na.rm = TRUE) + 0.15, .groups = "drop")


## Plotting
rel_abun.plot <- ggplot(estro.full, aes(y=fct_rev(Feature), x=log_rel_abundance, color=group_label)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.2) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(Feature, group_label), fill = group_label),
               position = position_dodge(width = 0.9),
               alpha = 0.35,
               outlier.shape = NA) +
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = color_values) +
  scale_fill_manual(name = NULL, values = color_values) +
  labs(y=NULL, x="log10(Relative Abundance)") +
  xlim(NA, 2.25) +
  # Add annotations for metadata term
  geom_text(data = B_position,
            aes(x = x_pos, y = Feature, label = beta_coef),
            inherit.aes = FALSE,
            hjust = 0, size = 3, color = "black") +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_markdown(size=9),
        axis.title.x = element_text(size = 9),
        legend.text = element_markdown(),
        legend.position = c(0.025, 0.82),
        legend.justification = c(0, 1), 
        legend.background = element_rect(color="black", fill = NA),
        legend.margin = margin(t=1, r=3, b=3)
  )

rel_abun.plot
## Save file
ggsave(
  filename = "/Users/ashlarnder/Downloads/relabun_2025-10-27.png",   # file name
  plot = rel_abun.plot,                   # which plot to save
  width = 6.5,                  # width
  height = 6,                 # height
  units = "in",               # units: "in", "cm", or "mm"
  dpi = 600                   # resolution
)

```


### ENDOCRINE SUBGROUP ANALYSIS

#### Diversity Measures
 
Alpha Diversity          
```{r stratified alpha diversity - species}
 
library(ggpubr)

## DATA PREP ####

## Calculating the alpha diversity measures
alpha_data <- alpha_calc_data(metaphlan4_long, metaphlan=TRUE) %>% 
   #Add in therapy type data
  merge(., gmbc_data %>% select(plate_ID, t_endo_drug.cat), by="plate_ID") %>% 
  #Filter to cases only
  filter(case_con == 1) %>% 
  #get it so groups are labeled as I want
  mutate(
    t_endo_drug.cat = factor(t_endo_drug.cat, 
                               levels = c("None", "Aromatase_inhibitor", "Tamoxifen"),
                               labels = c("None", "AI", "Tamoxifen")),
    collection_part = factor(collection_part, labels = c("Baseline", "Follow-up"))
  )

## Significant testing

# format wider
delta_df <- alpha_data %>%
  filter(!is.na(shannon)) %>%
  select(single_id, t_endo_drug.cat, collection_part, shannon) %>%
  pivot_wider(names_from = collection_part, values_from = shannon) %>%
  mutate(delta = `Follow-up` - Baseline)

kruskal_res <- kruskal.test(delta ~ t_endo_drug.cat, data = delta_df)
p_val <- kruskal_res$p.value

## Plotting SHANNON two rows ####

metaphlan.shannon <- ggplot(alpha_data, aes(x = t_endo_drug.cat, y = shannon, 
                                            color = t_endo_drug.cat, fill = t_endo_drug.cat)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      #stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      facet_wrap(~ collection_part, ncol = 1) +  # stack Baseline on top, Follow-up below
  
      #geom_line(stat = "summary", aes(group = t_endo_drug.cat), fun = median, color = "black") +  # Connect medians with lines
      scale_color_manual(values = c(
        "None" = "darkgrey",
        "A Inhibitor" = "sienna",
        "ER Blocker" = "darkblue"
      )) +
      scale_fill_manual(values = c(
        "None" = "darkgrey",
        "A Inhibitor" = "sienna",
        "ER Blocker" = "darkblue"
      )) +
      theme_minimal() +
      theme(
        legend.position = "none" ,#"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 8),   # makes the tick labels smaller
        axis.title.x = element_text(size = 10),   # makes the x-axis title smaller
        axis.title.y = element_text(size = 10) 
      ) +
      labs(x = "Species",
           y = "Shannon Index")

#metaphlan.shannon


## Plotting Delta SHANNON ####

dshannon.species <- ggplot(delta_df, 
       aes(x = t_endo_drug.cat, y = delta, 
           color = t_endo_drug.cat, fill = t_endo_drug.cat)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.2) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with
  scale_color_manual(values = c(
        "None" = "darkgrey",
        "AI" = "sienna",
        "Tamoxifen" = "darkblue"
      )) +
  scale_fill_manual(values = c(
        "None" = "darkgrey",
        "AI" = "sienna",
        "Tamoxifen" = "darkblue"
      )) +
  annotate("text",
           x = 2,
           y = min(delta_df$delta) + 0.15,  # slightly above your data
           label = paste0("Kruskal–Wallis p = ", signif(p_val, 3)),
           size = 2.5) +
  labs(x = NULL, y = "Δ Shannon Index") +
  theme_minimal() +
  theme(
    legend.position = "none" ,#"inside",
    #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
    #legend.justification = c("right", "bottom")
    axis.text.x  = element_text(size = 7.5),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 8),   # makes the tick labels smaller
    axis.title.x = element_text(size = 10),   # makes the x-axis title smaller
    axis.title.y = element_text(size = 10) 
  )

``` 
    
```{r stratified alpha diversity - pathways}
 
## DATA PREP #### 

## Calculating the alpha diversity measures
alpha_data <- alpha_calc_data(humann.path_long) %>% 
   #Add in therapy type data
  merge(., gmbc_data %>% select(plate_ID, t_endo_drug.cat), by="plate_ID") %>% 
  #Filter to cases only
  filter(case_con == 1) %>% 
  #get it so groups are labeled as I want
  mutate(
    t_endo_drug.cat = factor(t_endo_drug.cat, 
                               levels = c("None", "Aromatase_inhibitor", "Tamoxifen"),
                               labels = c("None", "AI", "Tamoxifen")),
    collection_part = factor(collection_part, labels = c("Baseline", "Follow-up"))
  )

## Significant testing

# format wider
delta_df <- alpha_data %>%
  filter(!is.na(shannon)) %>%
  select(single_id, t_endo_drug.cat, collection_part, shannon) %>%
  pivot_wider(names_from = collection_part, values_from = shannon) %>%
  mutate(delta = `Follow-up` - Baseline)

kruskal_res <- kruskal.test(delta ~ t_endo_drug.cat, data = delta_df)
p_val <- kruskal_res$p.value

## Plotting Sig Testing SHANNON ####

path.shannon <- ggplot(alpha_data, aes(x = t_endo_drug.cat, y = shannon, 
                                            color = t_endo_drug.cat, fill = t_endo_drug.cat)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      #stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      facet_wrap(~ collection_part, ncol = 1) +  # stack Baseline on top, Follow-up below
  
      #geom_line(stat = "summary", aes(group = t_endo_drug.cat), fun = median, color = "black") +  # Connect medians with lines
      scale_color_manual(values = c(
        "None" = "darkgrey",
        "A Inhibitor" = "sienna",
        "ER Blocker" = "darkblue"
      )) +
      scale_fill_manual(values = c(
        "None" = "darkgrey",
        "A Inhibitor" = "sienna",
        "ER Blocker" = "darkblue"
      )) +
      theme_minimal() +
      theme(
        legend.position = "none" ,#"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 8),   # makes the tick labels smaller
        axis.title.x = element_text(size = 10),   # makes the x-axis title smaller
        axis.title.y = element_text(size = 10) 
      ) +
      labs(x = "Pathways",
           y = "Shannon Index")

#path.shannon


## Plotting Delta SHANNON ####

dshannon.path <- ggplot(delta_df, 
       aes(x = t_endo_drug.cat, y = delta, 
           color = t_endo_drug.cat, fill = t_endo_drug.cat)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.2) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with
  scale_color_manual(values = c(
        "None" = "darkgrey",
        "AI" = "sienna",
        "Tamoxifen" = "darkblue"
      )) +
  scale_fill_manual(values = c(
        "None" = "darkgrey",
        "AI" = "sienna",
        "Tamoxifen" = "darkblue"
      )) +
  annotate("text",
           x = 2,
           y = min(delta_df$delta) - 0.025,  # slightly above your data
           label = paste0("Kruskal–Wallis p = ", signif(p_val, 3)),
           size = 2.5) +
  labs(x = NULL, y = "Δ Shannon Index") +
  theme_minimal() +
  theme(
    legend.position = "none" ,#"inside",
    #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
    #legend.justification = c("right", "bottom")
    axis.text.x  = element_text(size = 7.5),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 8),   # makes the tick labels smaller
    axis.title.x = element_text(size = 10),   # makes the x-axis title smaller
    axis.title.y = element_text(size = 10) 
  )

dshannon.path

``` 
   
```{r stratified alpha diversity - enzymes}
 
## DATA PREP ####

## Calculating the alpha diversity measures
alpha_data <- alpha_calc_data(humann.enzymes_long) %>% 
   #Add in therapy type data
  merge(., gmbc_data %>% select(plate_ID, t_endo_drug.cat), by="plate_ID") %>% 
  #Filter to cases only
  filter(case_con == 1) %>% 
  #get it so groups are labeled as I want
  mutate(
    t_endo_drug.cat = factor(t_endo_drug.cat, 
                               levels = c("None", "Aromatase_inhibitor", "Tamoxifen"),
                               labels = c("None", "AI", "Tamoxifen")),
    collection_part = factor(collection_part, labels = c("Baseline", "Follow-up"))
  )

## Significant testing

# format wider
delta_df <- alpha_data %>%
  filter(!is.na(shannon)) %>%
  select(single_id, t_endo_drug.cat, collection_part, shannon) %>%
  pivot_wider(names_from = collection_part, values_from = shannon) %>%
  mutate(delta = `Follow-up` - Baseline)
 
kruskal_res <- kruskal.test(delta ~ t_endo_drug.cat, data = delta_df)
p_val <- kruskal_res$p.value
 
## Plotting Sig Testing SHANNON ####

enzyme.shannon <- ggplot(alpha_data, aes(x = t_endo_drug.cat, y = shannon, 
                                            color = t_endo_drug.cat, fill = t_endo_drug.cat)) + 
      geom_jitter(width = 0.2, height = 0, alpha = 0.2) +  # Jitter the points along the x-axis
      geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with semi-transparent fill
      #stat_summary(fun = median, geom = "point", size = 3, color = "black", shape = 18) +  # Median points
      facet_wrap(~ collection_part, ncol = 1) +  # stack Baseline on top, Follow-up below
  
      #geom_line(stat = "summary", aes(group = t_endo_drug.cat), fun = median, color = "black") +  # Connect medians with lines
      scale_color_manual(values = c(
        "None" = "darkgrey",
        "A Inhibitor" = "sienna",
        "ER Blocker" = "darkblue"
      )) +
      scale_fill_manual(values = c(
        "None" = "darkgrey",
        "A Inhibitor" = "sienna",
        "ER Blocker" = "darkblue"
      )) +
      theme_minimal() +
      theme(
        legend.position = "none" ,#"inside",
        #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
        #legend.justification = c("right", "bottom")
        axis.text.x  = element_text(size = 8),   # makes the tick labels smaller
        axis.title.x = element_text(size = 10),   # makes the x-axis title smaller
        axis.title.y = element_text(size = 10) 
      ) +
      labs(x = "Enzymes",
           y = "Shannon Index")

#enzyme.shannon


## Plotting Delta SHANNON ####

dshannon.enzyme <- ggplot(delta_df, 
       aes(x = t_endo_drug.cat, y = delta, 
           color = t_endo_drug.cat, fill = t_endo_drug.cat)) +
  geom_jitter(width = 0.2, height = 0, alpha = 0.2) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +  # Boxplots on top, with
  scale_color_manual(values = c(
        "None" = "darkgrey",
        "AI" = "sienna",
        "Tamoxifen" = "darkblue"
      )) +
  scale_fill_manual(values = c(
        "None" = "darkgrey",
        "AI" = "sienna",
        "Tamoxifen" = "darkblue"
      )) +
  annotate("text",
           x = 2,
           y = min(delta_df$delta) - 0.025,  # slightly above your data
           label = paste0("Kruskal–Wallis p = ", signif(p_val, 3)),
           size = 2.5) +
  labs(x = NULL, y = "Δ Shannon Index") +
  scale_y_continuous(
    limits = c(NA, NA),      # extend slightly above 0.1
    breaks = c(-0.1, 0, 0.1)     # custom tick marks
  )+
  theme_minimal() +
  theme(
    legend.position = "none" ,#"inside",
    #legend.position.inside = c(0.99, 0.00),  # x and y (from bottom-left)
    #legend.justification = c("right", "bottom")
    axis.text.x  = element_text(size = 7.5),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 8),   # makes the tick labels smaller
    axis.title.x = element_text(size = 10),   # makes the x-axis title smaller
    axis.title.y = element_text(size = 10) 
  )

dshannon.enzyme

``` 
   
Beta Diversity    
```{r stratified beta diversity - species}
 
## Data Prep ####

## Calculate Bray-Curtis Distances
dist_mat <- metaphlan4_wide.t %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con, collection_part), by="plate_ID") %>% 
  filter(case_con == 1, collection_part==2) %>% #select cases only at follow-up
  select(!c(case_con, collection_part)) %>% 
  arrange(plate_ID) %>% 
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

## Significance testing
levels(gmbc_data$t_endo_drug.cat) <- c("AI", "None", "Tamoxifen")
groups <- gmbc_data %>%
  arrange(plate_ID) %>%  
  filter(case_con == 1, collection_part == 2) %>%
  pull(t_endo_drug.cat)
adonis2(dist_mat ~ groups)$`Pr(>F)`[1]
adonis2(dist_mat ~ groups)$`R2`[1]


## Principal Component Analysis
beta_pcoa <- cmdscale(dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#convert eigenvalues to percentage
beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)
#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  merge(., gmbc_data, by="plate_ID") #merge PC values onto meta data


# Rename factor levels
beta_pcoa.df$t_endo_drug.cat = as.factor(beta_pcoa.df$t_endo_drug.cat)
levels(beta_pcoa.df$t_endo_drug.cat) <- c("AI", "None", "Tamoxifen")

# Calculate the median coordinates for each group
medians <- beta_pcoa.df %>%
  group_by(t_endo_drug.cat) %>% #t_endo_drug.cat #t_rad_dose.cat
  summarize(
    med_PC1 = median(PC1), 
    med_PC2 = median(PC2),
    .groups = "drop"
  )

## Plotting PCOAs ####

metaphlan.beta_plot <- beta_pcoa.df %>% 
  ggplot(aes(x = PC1, y = PC2, color = t_endo_drug.cat)) + 
  geom_point(alpha=0.5, stroke = 0, size = 2) +
  stat_ellipse(aes(group = t_endo_drug.cat), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = c(
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue",
    "None" = "darkgrey"
  )) +
  labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
       #y = glue("PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)"),
       y = paste0(
       "<span style='font-size:10pt'>Beta Diversity of Cases</span><br>",
       "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"),
       color = "Endocrine Therapy") + 
  theme(
    axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
    axis.title.x = element_text(size = 8),   # makes the x-axis title smaller
    axis.title.y = element_markdown(),
    legend.position = "none",#"right",
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(0.5, "lines")
    ) +
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = beta_pcoa.df %>% 
                 left_join(medians, by = "t_endo_drug.cat"),
               aes(x = med_PC1, y = med_PC2, xend = PC1, yend = PC2,
                   color = t_endo_drug.cat),
               linewidth = 0.3, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_PC1, y = med_PC2, color = t_endo_drug.cat), 
             size = 3, shape = 4, stroke = 1.15)  # Shape 4 is an 'x' marker

```

```{r stratified beta diversity - path}
 
## DATA PREP ####

## Calculate Bray-Curtis Distances
dist_mat <- humann.path_wide %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con, collection_part), by="plate_ID") %>% 
  filter(case_con == 1, collection_part==2) %>% #select cases only at follow-up
  select(!c(case_con, collection_part)) %>% 
  arrange(plate_ID) %>% 
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

## Significance testing
levels(gmbc_data$t_endo_drug.cat) <- c("AI", "None", "Tamoxifen")
groups <- gmbc_data %>%
  arrange(plate_ID) %>%  
  filter(case_con == 1, collection_part == 2) %>%
  pull(t_endo_drug.cat)
adonis2(dist_mat ~ groups)$`Pr(>F)`[1]
adonis2(dist_mat ~ groups)$`R2`[1]


## Principal Component Analysis
beta_pcoa <- cmdscale(dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#convert eigenvalues to percentage
beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)
#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  merge(., gmbc_data, by="plate_ID") #merge PC values onto meta data


# Rename factor levels
beta_pcoa.df$t_endo_drug.cat = as.factor(beta_pcoa.df$t_endo_drug.cat)
levels(beta_pcoa.df$t_endo_drug.cat) <- c("AI", "None", "Tamoxifen")

# Calculate the median coordinates for each group
medians <- beta_pcoa.df %>%
  group_by(t_endo_drug.cat) %>% #t_endo_drug.cat #t_rad_dose.cat
  summarize(
    med_PC1 = median(PC1), 
    med_PC2 = median(PC2),
    .groups = "drop"
  )

## Plotting PCOAs ####

path.beta_plot <- beta_pcoa.df %>% 
  ggplot(aes(x = PC1, y = PC2, color = t_endo_drug.cat)) + 
  geom_point(alpha=0.5, stroke = 0, size = 2) +
  stat_ellipse(aes(group = t_endo_drug.cat), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = c(
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue",
    "None" = "darkgrey"
  )) +
  labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
       #y = glue("PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)"),
       y = paste0(
       "<span style='font-size:10pt'>Beta Diversity of Cases</span><br>",
       "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"),
       color = "Endocrine Therapy") + 
  theme(
    axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
    axis.title.x = element_text(size = 8),   # makes the x-axis title smaller
    axis.title.y = element_markdown(),
    legend.position = "none",#"right",
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(0.5, "lines")
    ) +
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = beta_pcoa.df %>% 
                 left_join(medians, by = "t_endo_drug.cat"),
               aes(x = med_PC1, y = med_PC2, xend = PC1, yend = PC2,
                   color = t_endo_drug.cat),
               linewidth = 0.3, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_PC1, y = med_PC2, color = t_endo_drug.cat), 
             size = 3, shape = 4, stroke = 1.15)  # Shape 4 is an 'x' marker

```
  
```{r stratified beta diversity - enzyme}
 
## DATA PREP ####

## Calculate Bray-Curtis Distances
dist_mat <- humann.enzymes_wide %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con, collection_part), by="plate_ID") %>% 
  filter(case_con == 1, collection_part==2) %>% #select cases only at follow-up
  select(!c(case_con, collection_part)) %>% 
  arrange(plate_ID) %>% 
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

## Significance testing
levels(gmbc_data$t_endo_drug.cat) <- c("AI", "None", "Tamoxifen")
groups <- gmbc_data %>%
  arrange(plate_ID) %>%  
  filter(case_con == 1, collection_part == 2) %>%
  pull(t_endo_drug.cat)
adonis2(dist_mat ~ groups)$`Pr(>F)`[1]
adonis2(dist_mat ~ groups)$`R2`[1]


## Principal Component Analysis
beta_pcoa <- cmdscale(dist_mat, #pcoa is list with $points (PCs per sample)
                   k = 2, #N-1 PCs (default 2)
                   eig = TRUE, # $eig (explained variation), $x (null), $ac, $GOF
                   add = TRUE) #added add=T from Sandeep
#convert eigenvalues to percentage
beta_perc_exp <- 100 * beta_pcoa$eig / sum(beta_pcoa$eig)
#turn pcoa points from list to dataframe
beta_pcoa.df <- as.data.frame(beta_pcoa$points) %>% #turn list component to dataframe
  rownames_to_column(var = "plate_ID") %>% #study id stored as row_name
  rename_with(~paste0("PC", gsub("V", "", .)), starts_with("V")) %>% #col names from V to PC
  merge(., gmbc_data, by="plate_ID") #merge PC values onto meta data


# Rename factor levels
beta_pcoa.df$t_endo_drug.cat = as.factor(beta_pcoa.df$t_endo_drug.cat)
levels(beta_pcoa.df$t_endo_drug.cat) <- c("AI", "None", "Tamoxifen")

# Calculate the median coordinates for each group
medians <- beta_pcoa.df %>%
  group_by(t_endo_drug.cat) %>% #t_endo_drug.cat #t_rad_dose.cat
  summarize(
    med_PC1 = median(PC1), 
    med_PC2 = median(PC2),
    .groups = "drop"
  )

## Plotting PCOAs ####
enzyme.beta_plot <- beta_pcoa.df %>% 
  ggplot(aes(x = PC1, y = PC2, color = t_endo_drug.cat)) + 
  geom_point(alpha=0.5, stroke = 0, size = 2) +
  stat_ellipse(aes(group = t_endo_drug.cat), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = c(
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue",
    "None" = "darkgrey"
  )) +
  labs(x = glue("PC 1 ({round(beta_perc_exp[1], digits = 1)}% exp var)"),
       #y = glue("PC 2 ({round(beta_perc_exp[2], digits = 1)}% exp var)"),
       y = paste0(
       "<span style='font-size:10pt'>Beta Diversity of Cases</span><br>",
       "<span style='font-size:8pt'>PC 2 (", round(beta_perc_exp[2], 1), "% exp var)</span>"),
       color = "Endocrine Therapy") + 
  ylim(NA, 0.25) +
  theme(
    axis.text.x  = element_text(size = 7),   # makes the tick labels smaller
    axis.text.y  = element_text(size = 7),   # makes the tick labels smaller
    axis.title.x = element_text(size = 8),   # makes the x-axis title smaller
    axis.title.y = element_markdown(),
    #legend.position = "none",#"right",
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(0.5, "lines"),
    #LEGEND
    legend.text = element_text(size = 6.8),
    legend.title = element_text(size = 7),
    legend.position = c(0.99, 0.99),    # inside top left (x,y in npc)
    legend.box = "horizontal",
    legend.justification = c("right", "top"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.5, "lines"),
    legend.spacing.y = unit(0.1, "cm"), 
    legend.spacing.x = unit(0.1, "cm"),
    legend.margin = margin(t = 0.1, r = 0.1, b = 0.1, l = 0.1, unit = "cm")
    ) +
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = beta_pcoa.df %>% 
                 left_join(medians, by = "t_endo_drug.cat"),
               aes(x = med_PC1, y = med_PC2, xend = PC1, yend = PC2,
                   color = t_endo_drug.cat),
               linewidth = 0.3, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_PC1, y = med_PC2, color = t_endo_drug.cat), 
             size = 3, shape = 4, stroke = 1.15)  # Shape 4 is an 'x' marker

```
    
Beta Dissimilarity
```{r stratified beta dissimilarity - species}
 
### DATA PREPARATION ####

# Calculate Bray-Curtis Distances (cases only)
dist_mat <- metaphlan4_wide.t %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con), by="plate_ID") %>% 
  filter(case_con == 1) %>% #select cases only at follow-up
  select(!c(case_con)) %>% 
  arrange(plate_ID) %>% 
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

# Convert distance object to full matrix
beta_matrix <- as.matrix(dist_mat)

# Get all pairwise combinations
pairwise_df <- expand.grid(sample1 = rownames(beta_matrix), sample2 = colnames(beta_matrix)) %>%
    mutate(across(everything(), as.character)) %>%  # convert both to character
    filter(sample1 < sample2)  # avoid duplicates and diagonals

# Add distance values
pairwise_df$distance <- mapply(function(x, y) beta_matrix[x, y], 
                                pairwise_df$sample1, pairwise_df$sample2)

# Add subject IDs, collection_part, case_con for each sample
pairwise_df <- pairwise_df %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, t_endo_drug.cat), 
            by = c("sample1" = "plate_ID")) %>%
  rename(singleid1 = single_id,
         time1 = collection_part,
         t_endo_drug.cat1 = t_endo_drug.cat) %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, t_endo_drug.cat), 
            by = c("sample2" = "plate_ID")) %>%
  rename(singleid2 = single_id,
         time2 = collection_part,
         t_endo_drug.cat2 = t_endo_drug.cat)



### CALCULATE WITHIN: Intra-Individual Dissimilarity (microbiome stability within indv over time)

# Filter out where ids match but different time points (data is symmetric)
intra_individual_df <- pairwise_df %>% 
  filter(singleid1 == singleid2, time1 != time2)

# Convert case_con to factor for nicer labels
intra_individual_df$t_endo_drug.cat1 <- as.factor(intra_individual_df$t_endo_drug.cat1)
levels(intra_individual_df$t_endo_drug.cat1) <- c("AI", "None", "Tamoxifen")
intra_individual_df <- intra_individual_df %>% 
  mutate(t_endo_drug.cat1 = factor(t_endo_drug.cat1, levels = 
        c("None", "AI", "Tamoxifen")
     ))

# Statistical comparison
kruskal_res <- kruskal.test(distance ~ t_endo_drug.cat1, data = intra_individual_df)
p_val <- kruskal_res$p.value


# Plot boxplot ####

metaphlan.intraplot <- ggplot(intra_individual_df, aes(x = t_endo_drug.cat1, y = distance, fill = t_endo_drug.cat1, color=t_endo_drug.cat1)) +
  geom_point(position = position_jitter(width = 0.1), size = 2, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  labs(x = NULL, y = "Intra-individual Dissimilarity") +
  scale_fill_manual(values = c(
    "None" = "darkgrey", 
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue"
  )) +
  scale_color_manual(values = c(
    "None" = "darkgrey", 
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue")) +
  annotate("text",
           x = 2,
           y = max(intra_individual_df$distance) + 0.05,  # slightly above your data
           label = paste0("Kruskal–Wallis p = ", signif(p_val, 3)),
           size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 7),
        axis.text.y  = element_text(size = 8),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10) 
        )

metaphlan.intraplot



# Median and IQR values ####

intra_individual_df %>%
  group_by(t_endo_drug.cat1) %>%
  summarise(
    n = n(),
    median = median(distance),
    IQR = IQR(distance),
    q25 = quantile(distance, 0.25),
    q75 = quantile(distance, 0.75)
  )

```

```{r stratified beta dissimilarity - path}

### DATA PREPARATION ####

# Calculate Bray-Curtis Distances (cases only)
dist_mat <- humann.path_wide %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con), by="plate_ID") %>% 
  filter(case_con == 1) %>% #select cases only at follow-up
  select(!c(case_con)) %>% 
  arrange(plate_ID) %>% 
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

# Convert distance object to full matrix
beta_matrix <- as.matrix(dist_mat)

# Get all pairwise combinations
pairwise_df <- expand.grid(sample1 = rownames(beta_matrix), sample2 = colnames(beta_matrix)) %>%
    mutate(across(everything(), as.character)) %>%  # convert both to character
    filter(sample1 < sample2)  # avoid duplicates and diagonals

# Add distance values
pairwise_df$distance <- mapply(function(x, y) beta_matrix[x, y], 
                                pairwise_df$sample1, pairwise_df$sample2)

# Add subject IDs, collection_part, case_con for each sample
pairwise_df <- pairwise_df %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, t_endo_drug.cat), 
            by = c("sample1" = "plate_ID")) %>%
  rename(singleid1 = single_id,
         time1 = collection_part,
         t_endo_drug.cat1 = t_endo_drug.cat) %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, t_endo_drug.cat), 
            by = c("sample2" = "plate_ID")) %>%
  rename(singleid2 = single_id,
         time2 = collection_part,
         t_endo_drug.cat2 = t_endo_drug.cat)



### CALCULATE WITHIN: Intra-Individual Dissimilarity (microbiome stability within indv over time)

# Filter out where ids match but different time points (data is symmetric)
intra_individual_df <- pairwise_df %>% 
  filter(singleid1 == singleid2, time1 != time2)

# Convert case_con to factor for nicer labels
intra_individual_df$t_endo_drug.cat1 <- as.factor(intra_individual_df$t_endo_drug.cat1)
levels(intra_individual_df$t_endo_drug.cat1) <- c("AI", "None", "Tamoxifen")
intra_individual_df <- intra_individual_df %>% 
  mutate(t_endo_drug.cat1 = factor(t_endo_drug.cat1, levels = 
        c("None", "AI", "Tamoxifen")
     ))



# Statistical comparison

kruskal_res <- kruskal.test(distance ~ t_endo_drug.cat1, data = intra_individual_df)
p_val <- kruskal_res$p.value

# Plot boxplot ####

path.intraplot <- ggplot(intra_individual_df, aes(x = t_endo_drug.cat1, y = distance, fill = t_endo_drug.cat1, color=t_endo_drug.cat1)) +
  geom_point(position = position_jitter(width = 0.1), size = 2, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  labs(x = NULL, y = "Intra-individual Dissimilarity") +
  scale_fill_manual(values = c(
    "None" = "darkgrey", 
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue"
  )) +
  scale_color_manual(values = c(
    "None" = "darkgrey", 
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue")) +
  annotate("text",
           x = 2,
           y = max(intra_individual_df$distance) + 0.025,  # slightly above your data
           label = paste0("Kruskal–Wallis p = ", signif(p_val, 3)),
           size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 7),
        axis.text.y  = element_text(size = 8),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10) 
        )

path.intraplot


# Median and IQR values ####

intra_individual_df %>%
  group_by(t_endo_drug.cat1) %>%
  summarise(
    n = n(),
    median = median(distance),
    IQR = IQR(distance),
    q25 = quantile(distance, 0.25),
    q75 = quantile(distance, 0.75)
  )

```
 
```{r stratified beta dissimilarity - enzyme}
 
### DATA PREPARATION ####

# Calculate Bray-Curtis Distances (cases only)
dist_mat <- humann.enzymes_wide %>% 
  merge(., gmbc_data %>% select(plate_ID, case_con), by="plate_ID") %>% 
  filter(case_con == 1) %>% #select cases only at follow-up
  select(!c(case_con)) %>% 
  arrange(plate_ID) %>% 
  column_to_rownames("plate_ID") %>% #turning column values to name
  as.matrix() %>% #convert to matrix
  vegdist(method = "bray") #calc bray-curtis matrix of distances using vegan

# Convert distance object to full matrix
beta_matrix <- as.matrix(dist_mat)

# Get all pairwise combinations
pairwise_df <- expand.grid(sample1 = rownames(beta_matrix), sample2 = colnames(beta_matrix)) %>%
    mutate(across(everything(), as.character)) %>%  # convert both to character
    filter(sample1 < sample2)  # avoid duplicates and diagonals

# Add distance values
pairwise_df$distance <- mapply(function(x, y) beta_matrix[x, y], 
                                pairwise_df$sample1, pairwise_df$sample2)

# Add subject IDs, collection_part, case_con for each sample
pairwise_df <- pairwise_df %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, t_endo_drug.cat), 
            by = c("sample1" = "plate_ID")) %>%
  rename(singleid1 = single_id,
         time1 = collection_part,
         t_endo_drug.cat1 = t_endo_drug.cat) %>%
  left_join(gmbc_data %>% select(plate_ID, single_id, collection_part, t_endo_drug.cat), 
            by = c("sample2" = "plate_ID")) %>%
  rename(singleid2 = single_id,
         time2 = collection_part,
         t_endo_drug.cat2 = t_endo_drug.cat)



### CALCULATE WITHIN: Intra-Individual Dissimilarity (microbiome stability within indv over time)

# Filter out where ids match but different time points (data is symmetric)
intra_individual_df <- pairwise_df %>% 
  filter(singleid1 == singleid2, time1 != time2)

# Convert case_con to factor for nicer labels
intra_individual_df$t_endo_drug.cat1 <- as.factor(intra_individual_df$t_endo_drug.cat1)
levels(intra_individual_df$t_endo_drug.cat1) <- c("AI", "None", "Tamoxifen")
intra_individual_df <- intra_individual_df %>% 
  mutate(t_endo_drug.cat1 = factor(t_endo_drug.cat1, levels = 
        c("None", "AI", "Tamoxifen")
     ))


# Statistical comparison

kruskal_res <- kruskal.test(distance ~ t_endo_drug.cat1, data = intra_individual_df)
p_val <- kruskal_res$p.value 

# Plot boxplot ####

enzyme.intraplot <- ggplot(intra_individual_df, aes(x = t_endo_drug.cat1, y = distance, fill = t_endo_drug.cat1, color=t_endo_drug.cat1)) +
  geom_point(position = position_jitter(width = 0.1), size = 2, alpha = 0.3) +
  geom_boxplot(outlier.shape = NA, alpha = 0.3) +
  labs(x = NULL, y = "Intra-individual Dissimilarity") +
  scale_fill_manual(values = c(
    "None" = "darkgrey", 
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue"
  )) +
  scale_color_manual(values = c(
    "None" = "darkgrey", 
    "AI" = "sienna", 
    "Tamoxifen" = "darkblue")) +
  annotate("text",
           x = 2,
           y = max(intra_individual_df$distance) + 0.025,  # slightly above your data
           label = paste0("Kruskal–Wallis p = ", signif(p_val, 3)),
           size = 2.5) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 7),
        axis.text.y  = element_text(size = 8),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10) 
        )

enzyme.intraplot


# Median and IQR values ####

intra_individual_df %>%
  group_by(t_endo_drug.cat1) %>%
  summarise(
    n = n(),
    median = median(distance),
    IQR = IQR(distance),
    q25 = quantile(distance, 0.25),
    q75 = quantile(distance, 0.75)
  )

```

Combined Plot
```{r COMBINED PLOTTING }

library(cowplot) 
library(grid)

png("/Users/ashlarnder/Downloads/microbiome_therapy_2025-11-10.png", 
    width = 6.5 * 600, height = 7 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #row 1 - alpha diversity measures
    plot_grid(
      dshannon.species, metaphlan.intraplot, metaphlan.beta_plot, 
      nrow = 1,
      ncol = 3,
      rel_widths = c(1.0,1.0,1.3)
    ),
    #row 3 - Bray-curtis dissimilaries
    plot_grid(
           dshannon.path, path.intraplot, path.beta_plot,
      #arguments
      nrow = 1,
      ncol = 3,
      rel_widths = c(1.0,1.0,1.3)
    ),
    #row 2 - beta diversity
    plot_grid(
       dshannon.enzyme, enzyme.intraplot, enzyme.beta_plot,
      nrow = 1,
      ncol = 3,
      rel_widths = c(1.0,1.0,1.3)
    ),
    #arguments
    nrow = 3,
    ncol = 1,
    rel_heights = c(1,1,1)
  )
)


grid.text("Species", x = unit(0.01, "npc"), y = unit(0.987, "npc"), #0.099 x
          just = "left",  
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Pathways", x = unit(0.01, "npc"), y = unit(0.66, "npc"), #0.348
          just = "left",  
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Enzymes", x = unit(0.01, "npc"), y = unit(0.325, "npc"),
          just = "left",  
          gp = gpar(fontsize = 15, fontface = "bold"))

# grid.text("Cases", x = unit(0.15, "npc"), y = unit(0.993, "npc"), #0.099 x
#           gp = gpar(fontsize = 9))
# 
# grid.text("Controls", x = unit(0.305, "npc"), y = unit(0.993, "npc"), #0.099 x
#           gp = gpar(fontsize = 9))


dev.off()

```



#### Stratified GLMMs

Running Models
```{r GLMMs - genera}

apply_iterative.modeling.therapy(
  predictor_vars = metaphlan4_wide.g %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = metaphlan4_wide.g %>% transform_metagenomics(),
  file_name = "genus_long-therapy"
)

```

```{r GLMMs - species}

apply_iterative.modeling.therapy(
  predictor_vars = metaphlan4_wide.t %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = metaphlan4_wide.t %>% transform_metagenomics(),
  file_name = "species_long-therapy"
)

```

```{r GLMMs - enzymes}

apply_iterative.modeling.therapy(
  predictor_vars = humann.enzymes_wide %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = humann.enzymes_wide %>% transform_metagenomics(),
  file_name = "enzymes_long-therapy"
)

```
 
```{r GLMMs - pathways}
 
apply_iterative.modeling.therapy(
  predictor_vars = humann.path_wide %>% select(-plate_ID) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
     )
  ),
  dataset.model = humann.path_wide %>% transform_metagenomics(),
  file_name = "pathways_long-therapy"
)

```

Volcano Plotting
```{r GLMM genera - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_genus_long-therapy_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    term = ifelse(term == "Interaction (A Inhibitor)", 
                  "Interaction (AI)",
                  "Interaction (tamoxifen)"),
    sig_color = case_when(
      (term == "Interaction (AI)" & p.value < 0.05) ~ "1",
      (term == "Interaction (tamoxifen)" & p.value < 0.06) ~ "2",
      TRUE ~ "3"),
    alpha_val = ifelse(q.value < 0.25, 0.9, 0.4)
    )

## Run plot
volcano.genus <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value),
                                           shape = term) ) +
  geom_point(aes(color = sig_color, alpha = alpha_val), 
             size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", alpha=0.8, color = "gray80") +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey70") +
  scale_color_manual(values = c("1" = "sienna", 
                                "2" = "darkblue",
                                "3" = "grey70")) +
  scale_alpha_identity() +  # use exact alpha values
  scale_x_continuous(limits = c(-2.9, 4.48)) +
  scale_y_continuous(limits = c(0, 3.6)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Genera"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

volcano.genus

```

```{r GLMM species - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_species_long-therapy_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    term = ifelse(term == "Interaction (A Inhibitor)", 
                  "Interaction (AI)",
                  "Interaction (tamoxifen)"),
    sig_color = case_when(
      (term == "Interaction (AI)" & p.value < 0.05) ~ "1",
      (term == "Interaction (tamoxifen)" & p.value < 0.06) ~ "2",
      TRUE ~ "3"),
    alpha_val = ifelse(q.value < 0.25, 0.9, 0.4)
    )

## Run plot
volcano.species <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value),
                                           shape = term) ) +
  geom_point(aes(color = sig_color, alpha = alpha_val), 
             size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", alpha=0.8, color = "gray80") +
  geom_hline(yintercept = -log10(0.001), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey70") +
  scale_color_manual(values = c("1" = "sienna", 
                                "2" = "darkblue",
                                "3" = "grey70")) +
  scale_alpha_identity() +  # use exact alpha values
  scale_x_continuous(limits = c(-2.9, 4.48)) +
  scale_y_continuous(limits = c(0, 3.6)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

volcano.species

```

```{r GLMM pathways - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_pathways_long-therapy_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    term = ifelse(term == "Interaction (A Inhibitor)", 
                  "Interaction (AI)",
                  "Interaction (tamoxifen)"),
    sig_color = case_when(
      (term == "Interaction (AI)" & p.value < 0.05) ~ "1",
      (term == "Interaction (tamoxifen)" & p.value < 0.06) ~ "2",
      TRUE ~ "3"),
    alpha_val = ifelse(q.value < 0.25, 0.9, 0.4),
    estimate = ifelse(feature_clean=="PWY3O-4107: NAD salvage pathway V (PNC V cycle)" & term=="Interaction (tamoxifen)", 2.6, estimate)
    )

## Run plot
volcano.path <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value),
                                           shape = term) ) +
  geom_point(aes(color = sig_color, alpha = alpha_val), 
             size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", alpha=0.8, color = "gray80") +
  geom_hline(yintercept = -log10(0.00168), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey70") +
  scale_color_manual(values = c("1" = "sienna", 
                                "2" = "darkblue",
                                "3" = "grey70")) +
  scale_alpha_identity() +  # use exact alpha values
  scale_x_continuous(limits = c(-2.9, 4.48)) +
  scale_y_continuous(limits = c(0, 3.6)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    title = "Pathways"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    legend.position = "none",
    text = element_text(size = 9)
  )

volcano.path

```

```{r GLMM enzymes - volcano plot}

## Read in data
glmms_results <- readxl::read_excel("GLMMs_enzymes_long-therapy_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    term = ifelse(term == "Interaction (A Inhibitor)", 
                  "Interaction (AI)",
                  "Interaction (tamoxifen)"),
    sig_color = case_when(
      (term == "Interaction (AI)" & p.value < 0.05) ~ "Interaction (AI)",
      (term == "Interaction (tamoxifen)" & p.value < 0.06) ~ "Interaction (tamoxifen)",
      TRUE ~ NA_character_),
      #TRUE ~ "p > 0.05"),
    alpha_val = ifelse(q.value < 0.25, 0.9, 0.4)
    )

## Run plot
volcano.enzyme <- ggplot(glmms_results, aes(x = estimate, 
                                            y = -log10(p.value),
                                           shape = term) ) +
    # grey nonsignificant points
  geom_point(
    data = subset(glmms_results, is.na(sig_color)),
    color = "grey70", shape = 16, size = 2, alpha = 0.4
  ) +
  geom_point(
    data = subset(glmms_results, !is.na(sig_color)),
    aes(color = sig_color, alpha = alpha_val), 
             size = 2) +
  
  geom_vline(xintercept = 0, linetype = "solid", alpha=0.8, color = "gray80") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey70") +
  # scale_color_manual(values = c("1" = "sienna", 
  #                               "2" = "darkblue",
  #                               "3" = "grey70")) +
  scale_color_manual(values = c(
      "Interaction (AI)" = "sienna",
      "Interaction (tamoxifen)" = "darkblue")
  ) +
  scale_shape_manual(values = c(
      "Interaction (AI)" = 16,
      "Interaction (tamoxifen)" = 17)
  ) +
  scale_alpha_identity() +  # use exact alpha values
  scale_x_continuous(limits = c(-2.9, 4.48)) +
  scale_y_continuous(limits = c(0, 3.625)) +
  labs(
    x = "log(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, 
    shape = NULL,
    title = "Enzymes"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    text = element_text(size = 9),
    #legend
    legend.text = element_text(size = 6.8),
    legend.title = element_text(size = 8),
    legend.position = c(-0.1, 1.1),    # inside top left (x,y in npc)
    #legend.justification = c("right", "bottom"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.5, "lines"),
    legend.spacing.y = unit(0.1, "cm"), 
    legend.spacing.x = unit(0.1, "cm"),
    legend.margin = margin(t = 0.1, r = 0.1, b = 0.1, l = 0.1, unit = "cm")
  ) +
  # Combine legends properly
  guides(
    color = guide_legend(override.aes = list(size = 3, alpha = 1))
  )

volcano.enzyme

```
 
Combined Volcano Plot
```{r COMBINED - volcano plot}
 
 
library(cowplot)

png("GLMMs-therapy_volcano_2025-09-23.png", 
    width = 6.5 * 600, height = 6.5 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #plots
    volcano.genus,volcano.species, 
    volcano.path, volcano.enzyme,
    #arguments
    nrow = 2,
    ncol = 2,
    rel_widths = c(1,1),
    rel_heights = c(1,1)
  )
)

#grid.text("Species", x = unit(0.45, "npc"), y = unit(0.985, "npc"), #0.099 x
          #gp = gpar(fontsize = 15, fontface = "bold"))


dev.off()



```
 

Relative Abundance Plot

```{r Relative abundance plots - NEW}

library(patchwork)

# Relevant Features
features <- c(
  "s__GGB33928_SGB15225|t__SGB15225",  
  "PYRIDNUCSAL-PWY: NAD salvage pathway I (PNC VI cycle)",
  "PWY3O-4107: NAD salvage pathway V (PNC V cycle)",
  "P105-PWY: TCA cycle IV (2-oxoglutarate decarboxylase)"
)

# Select from each datatype
estro.t <- metaphlan4_wide.t %>% select(plate_ID, any_of(features))
estro.e <- humann.enzymes_wide %>% select(plate_ID, any_of(features))
estro.p <- humann.path_wide %>% select(plate_ID, any_of(features))

# Merge all together
estro.full <- estro.t %>% 
  left_join(estro.e, by="plate_ID") %>% 
  left_join(estro.p, by="plate_ID")

# Clean and reshape
estro.long <- estro.full %>% 
  #clean up the data
  rename(
    "GGB33928 SGB15225" = "s__GGB33928_SGB15225|t__SGB15225",  
    "TCA cycle IV (P105-PWY)"= "P105-PWY: TCA cycle IV (2-oxoglutarate decarboxylase)",
    "NAD salvage pathway I (PYRIDNUCSAL-PWY)"= "PYRIDNUCSAL-PWY: NAD salvage pathway I (PNC VI cycle)",
    "NAD salvage pathway V (PWY3O-4107)" = "PWY3O-4107: NAD salvage pathway V (PNC V cycle)"
  ) %>% 
  merge(., gmbc_data %>% 
          select(plate_ID, single_id, case_con, collection_part, t_endo_drug.cat), 
        by="plate_ID") %>% 
  filter(case_con == "1") %>% 
  mutate(t_endo_drug.cat1 = factor(t_endo_drug.cat, 
                  levels = c("None", "Aromatase_inhibitor", "Tamoxifen"),
                  labels = c("None", "AI", "Tamoxifen"))
  ) %>% 
  # Pivot to long form to prepare for delta calculation
  pivot_longer(
    cols = -c(plate_ID,single_id, case_con, collection_part, t_endo_drug.cat, t_endo_drug.cat1),
    names_to = "Feature",
    values_to = "relative_abundance"
  ) %>% 
  filter(relative_abundance > 0)

# Pivot wider
estro.delta <- estro.long %>%
  select(single_id, Feature, relative_abundance, collection_part, t_endo_drug.cat1) %>%
  pivot_wider(
    id_cols = c(single_id, Feature, t_endo_drug.cat1),
    names_from = collection_part,
    values_from = relative_abundance,
    names_prefix = "Time"
  ) %>%
  mutate(
    Time1 = ifelse( is.na(Time1), 0, Time1),
    Time2 = ifelse( is.na(Time2), 0, Time2),
    delta_abundance = Time2 - Time1
    ) %>% 
  mutate(Group = ifelse(Feature == "GGB33928 SGB15225", "Species", "Pathways"))

# Factor levels for Features (for plotting)
estro.delta <- estro.delta %>% 
  mutate(Feature = factor(Feature, levels = c(
    "TCA cycle IV (P105-PWY)",
    "NAD salvage pathway I (PYRIDNUCSAL-PWY)",
    "NAD salvage pathway V (PWY3O-4107)",
    "GGB33928 SGB15225"
  ))) 

# Colours
color_values = c(
  "None" = "grey",
  "AI" = "sienna",
  "Tamoxifen" = "darkblue"
)



# Subset your data by facet variable
top_data <- subset(estro.delta, Group == "Pathways")   # replace with your actual facet name
bottom_data <- subset(estro.delta, Group == "Species")

  
p1 <- ggplot(top_data, 
       aes(y = fct_rev(Feature), x = delta_abundance,
           color = t_endo_drug.cat1, fill = t_endo_drug.cat1)) +
  geom_boxplot(alpha = 0.35, outlier.shape = NA) +
  facet_wrap(~Group, scales = "free", ncol=1) +
  xlim(c(-7e-05,7e-05)) +
  scale_color_manual(values = color_values, name = NULL) +
  scale_fill_manual(values = color_values, name = NULL) +
  labs(y = NULL, x = "Δ Relative Abdundance (Follow-up − Baseline)") +
  theme_classic() +
  theme(
    axis.text.y = element_markdown(size = 8),
    axis.title.x = element_blank(),
    legend.text = element_markdown(),
    legend.position = c(0.8, 0.95),
    legend.justification = c(0, 1),
    legend.background = element_rect(color = "black", fill = NA),
    legend.margin = margin(t = 1, r = 3, b = 3),
    strip.text = element_blank()
  ) +
  guides(
    fill = guide_legend(reverse = TRUE),
    color = guide_legend(reverse = TRUE)
  )

# reuse same plot but with other facet data
p2 <- p1 %+% bottom_data  + 
  scale_x_continuous(limits = NULL) +   # removes inherited xlim()
  theme(legend.position = "none",
        axis.title.x = element_text(size = 8)) 

# Combine with heights 3:1
combined_plot <- p1 / p2 + plot_layout(heights = c(3, 1))

combined_plot



## Save file
ggsave(
  filename = "relabun_2025-11-13.png",   # file name
  plot = combined_plot,                   # which plot to save
  width = 8,                  # width
  height = 4.5,                 # height
  units = "in",               # units: "in", "cm", or "mm"
  dpi = 600                   # resolution
)

```



## METABOLOMICS

#### CASE VS CONTROLS

Summary 
```{r Summary}

## Characterizing metabolites (zero-count, median non-zero val)

#Function 
metabo_characterize  <- function(metabo_data) {
  #Total number of metabolites
  print(glue("The total number of metabolites: {ncol(metabo_data)-1}"))
  
  #Calculate number zeros per metabolite and non-zero val
  metabo_data %>%
    #add in case_con val from metadata
    left_join(gmbc_data %>% select(study_id, case_con), by="study_id") %>% 
    #reshape
    pivot_longer(-c(study_id, case_con), names_to = "Metabolite", values_to = "relab") %>%
    group_by(Metabolite) %>%
    summarise(
      # Median of non-zero values
      median_nonzero = median(relab[relab > 0], na.rm = TRUE),
      # Count of zero values
      zero_count = sum(relab == 0, na.rm = TRUE),
      # Count non-zero in cases and controls
      n_case_nonzero = sum(relab > 0 & case_con == 1, na.rm = TRUE),
      n_control_nonzero = sum(relab > 0 & case_con == 0, na.rm = TRUE),
      #mean and SD
      mean_nonzero = mean(relab, na.rm = TRUE),
      sd_nonzero = sd(relab, na.rm = TRUE),
    ) %>%
    arrange(desc(zero_count)) %>% 
    mutate(across(where(is.character), ~ gsub("\\.", " ", .))) %>% 
  select(Metabolite, zero_count, n_case_nonzero, mean_nonzero, sd_nonzero) %>% 
    view()
}

#Apply
metabo_characterize(plasma_abund)
metabo_characterize(stool_abund)




## Comparing between plasma and fecal samples

#In plasma, not in stool
plasma_only <- setdiff(names(metabo.plasma_wide), names(metabo.stool_wide))  # Columns in df1 but not in df2
view(plasma_only)
#In stool, not in plasma
stool_only <- setdiff(names(metabo.stool_wide), names(metabo.plasma_wide))  # Columns in df2 but not in df1
view(stool_only)
#in both
in_both <- intersect(names(metabo.plasma_wide), names(metabo.stool_wide))  # Common columns
view(in_both)


```


Multi-level PLS-DA   
```{r Multi-level PLS-DA - plasma }

#library(ropls)
library(mixOmics)
library(dplyr)

## Prepare Data ####

#dataset
oplsda <- metabo.plasma_wide %>% 
  transform_metabo(., remove_outliers = F, merge = F) %>% 
  merge(., gmbc_data %>% dplyr::select(study_id, single_id, case_con, collection_part), by="study_id")

# predictors (metabolites)
X <- oplsda %>% dplyr::select(-c(case_con, study_id, single_id, collection_part)) %>% as.matrix()
# outcome variable (e.g., case vs. control)
Y <- as.factor(oplsda$case_con)  
# subject ids for repeated measures
subject <- oplsda$single_id

## Run multilevel PLS-DA (accounts for repeated measures)
plsda_model <- plsda(X, Y, multilevel = subject, ncomp = 2)


## REPORT FEATURES/PERFORMANCE ####

# ## Important Variables
# vipVn.plasma <- vip(plsda_model) %>%   # Variable Importance in Projection (VIP) scores
#   as.data.frame() %>% 
#   arrange(desc(.))
# # write.csv(vipVn.plasma, "/Users/ashlarnder/Downloads/vip.plasma.csv", row.names = TRUE)
# head(vipVn.plasma, 10)


## Explained variance
plsda_model$prop_expl_var$X

# Run repeated cross-validation for mPLS-DA
perf_mplsda <- perf(
  plsda_model,
  validation = "Mfold",
  folds = 5,
  nrepeat = 5,
  dist = "centroids.dist"
)

# Classification error rates
# Overall classification error per component
perf_mplsda$error.rate$overall
# Balanced error rate (BER)
perf_mplsda$error.rate$BER


## Scores Plotting ####

#predictive component scores
scores <- plsda_model$variates$X
#including group labels
scores <- data.frame(
    comp1 = scores[, 1],
    comp2 = scores[, 2],
    case_con = Y,
    time = oplsda$collection_part,
    ID = subject
  ) %>% 
  mutate(group_label = case_when(
      case_con == 1 & time == 1 ~ "Case, Baseline",
      case_con == 1 & time == 2 ~ "Case, Follow-up",
      case_con == 0 & time == 1 ~ "Control, Baseline",
      case_con == 0 & time == 2 ~ "Control, Follow-up"
  ))
#get explained variance for predictive components
expl_var <- plsda_model$prop_expl_var$X

#calculate the median coordinates for each group
medians <- scores %>%
  group_by(group_label) %>%
  summarize(
    med_comp1 = median(comp1), 
    med_comp2 = median(comp2),
    .groups = "drop"
  )

#set colours by group
group_colors <- c(
  "Case, Baseline" = "darkblue",     # medium blue
  "Case, Follow-up" = "#6ba",    # light blue
  "Control, Baseline" = "sienna",  # dark grey
  "Control, Follow-up" = "wheat3" 
)


#build the plot
pls_plot.plasma <- ggplot(scores, aes(x = comp1, y = comp2, 
                                      color = group_label, shape = time)) +
  geom_point(alpha=0.6) +
  stat_ellipse(aes(group = group_label), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = group_colors) +# c("0" = "darkgray", "1" = "blue")) +
  scale_shape_manual(values = c("1" = 16, "2" = 17)) + #circle baseline #triangle FU
  theme_minimal() +
  labs(   
    x = paste0("Component 1 (", round(expl_var[[1]], 2), "% exp var)"),
    y = paste0("Component 2 (", round(expl_var[[2]], 2), "% exp var)"),
    color = "Case_Status",
    shape = "Time Point"
  ) +
  theme(legend.position = "none",#"right",
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines")) +
  
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = scores %>% 
                 left_join(medians, by = "group_label"),
               aes(x = med_comp1, y = med_comp2, xend = comp1, yend = comp2,
                   color = group_label),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_comp1, y = med_comp2, color = group_label), 
             size = 2.5, shape = 4, stroke = 1.2)  # Shape 4 is an 'x' marker

pls_plot.plasma


```
 
```{r Multi-level PLS-DA - stool }

#library(ropls)
library(mixOmics)
library(dplyr)

## Prepare Data ####

#dataset
oplsda <- metabo.stool_wide %>% 
  transform_metabo(., remove_outliers = F, merge = F) %>% 
  merge(., gmbc_data %>% dplyr::select(study_id, single_id, case_con, collection_part), by="study_id")

# predictors (metabolites)
X <- oplsda %>% dplyr::select(-c(case_con, study_id, single_id, collection_part)) %>% as.matrix()
# outcome variable (e.g., case vs. control)
Y <- as.factor(oplsda$case_con)  
# subject ids for repeated measures
subject <- oplsda$single_id


## Run multilevel PLS-DA (accounts for repeated measures)
plsda_model <- plsda(X, Y, multilevel = subject, ncomp = 2)

## PRINT FEATURES / PERFORMANCE ####

# ## Important Variables
# vipVn.plasma <- vip(plsda_model) %>%   # Variable Importance in Projection (VIP) scores
#   as.data.frame() %>% 
#   arrange(desc(.))
# write.csv(vipVn.plasma, "/Users/ashlarnder/Downloads/vip.stool.csv", row.names = TRUE)
# head(vipVn.plasma, 10)


## Explained variance
plsda_model$prop_expl_var$X

# Run repeated cross-validation for mPLS-DA
perf_mplsda <- perf(
  plsda_model,
  validation = "Mfold",
  folds = 5,
  nrepeat = 5,
  dist = "centroids.dist"
)

# Classification error rates
# Overall classification error per component
perf_mplsda$error.rate$overall
# Balanced error rate (BER)
perf_mplsda$error.rate$BER


## Scores Plotting ####

#predictive component scores
scores <- plsda_model$variates$X
#including group labels
scores <- data.frame(
    comp1 = scores[, 1],
    comp2 = scores[, 2],
    case_con = Y,
    time = oplsda$collection_part,
    ID = subject
  ) %>% 
  mutate(group_label = case_when(
      case_con == 1 & time == 1 ~ "Case, Baseline",
      case_con == 1 & time == 2 ~ "Case, Follow-up",
      case_con == 0 & time == 1 ~ "Control, Baseline",
      case_con == 0 & time == 2 ~ "Control, Follow-up"
  ))
#get explained variance for predictive components
expl_var <- plsda_model$prop_expl_var$X

#calculate the median coordinates for each group
medians <- scores %>%
  group_by(group_label) %>%
  summarize(
    med_comp1 = median(comp1), 
    med_comp2 = median(comp2),
    .groups = "drop"
  )

#set colours by group
group_colors <- c(
  "Case, Baseline" = "darkblue",     # medium blue
  "Case, Follow-up" = "#6ba",    # light blue
  "Control, Baseline" = "sienna",  # dark grey
  "Control, Follow-up" = "wheat3" 
)


#build the plot
pls_plot.stool <- ggplot(scores, aes(x = comp1, y = comp2, 
                                     color = group_label, shape = time)) +
  geom_point(alpha=0.6) +
  stat_ellipse(aes(group = group_label), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = group_colors) +
  scale_shape_manual(values = c("1" = 16, "2" = 17)) +  # baseline = circle, FU = triangle
  theme_minimal() +
  ylim(-17, NA) +
  labs(   
    x = paste0("Component 1 (", round(expl_var[[1]], 2), "% exp var)"),
    y = paste0("Component 2 (", round(expl_var[[2]], 2), "% exp var)"),
    color = NULL,
    shape = NULL  # remove separate shape title
  ) +
  theme(
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(0.5, "lines"),
    # LEGEND
    legend.text = element_text(size = 6.5),
    legend.title = element_text(size = 8),
    legend.position = c(0.18, 0.14), #0.01, 0.999
    legend.box = "horizontal",
    legend.justification = c("left", "top"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.margin  = margin(t = 2, r = 2, b = 2, l = 2),
    legend.key.size = unit(0.7, "lines"),
    legend.spacing.y = unit(0.1, "cm"), 
    legend.spacing.x = unit(0.1, "cm")
  ) +
  guides(
    color = guide_legend(
      ncol = 2, byrow = TRUE, order = 1,
      override.aes = list(shape = c(16, 17)) # all shapes same in legend
    ),
    shape = "none"  # remove separate shape legend
  ) +
  geom_segment(data = scores %>% left_join(medians, by = "group_label"),
               aes(x = med_comp1, y = med_comp2, xend = comp1, yend = comp2,
                   color = group_label),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  geom_point(data = medians, 
             aes(x = med_comp1, y = med_comp2, color = group_label), 
             size = 2.5, shape = 4, stroke = 1.2)

pls_plot.stool



```
 
```{r combined plot}

library(cowplot)
library(grid)


png("metabo_mpls_2025-09-23.png", 
    width = 6.5 * 600, height = 3.25 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    pls_plot.plasma, pls_plot.stool,
    nrow = 1,
    rel_widths = c(1, 1)
  )
)
  
grid.text("Plasma", x = unit(0.14, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Stool", x = unit(0.63, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

dev.off()

```

     
GLMMs
```{r plasma metabolites - run GLMMs }

apply_iterative.modeling(
  predictor_vars = metabo.plasma_wide %>% select(-study_id) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean" 
    )
  ),
  dataset.model = metabo.plasma_wide %>% transform_metabo(),
  file_name = "metabo.plasma_long"
)

```

```{r stool metabolites - run GLMMs }

apply_iterative.modeling(
  predictor_vars = metabo.stool_wide %>% select(-study_id) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean"
    )
  ),
  dataset.model = metabo.stool_wide %>% transform_metabo(),
  file_name = "metabo.stool_long"
)


```

    
```{r plasma metabolites - volcano plot}

## Read in the data
glmms_results <- readxl::read_excel("GLMMs_metabo.plasma_long_2025-10-08.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = case_when(
      p.value < 0.0045 ~ "0",
      p.value < 0.05 ~ "1", 
      TRUE ~ "2")
    )

## Run plot
volcano.plasma <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(color = sig_color), 
             alpha = case_when(
               glmms_results$sig_color == "0" ~ 0.9,
               glmms_results$sig_color == "1" ~ 0.3,
               glmms_results$sig_color == "2" ~ 0.5
             ),
             size = 2) +
  geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray70") +
  geom_hline(yintercept = -log10(0.0045), linetype = "dashed", color = "black") +
  scale_color_manual(values = c("0" = "darkblue","1" = "darkblue", "2" = "grey70")) +
  scale_x_continuous(limits = c(-3.85, 3.57)) +
  scale_y_continuous(limits = c(0, 4.105)) +
  labs(
    x = "log2(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL,
    title = "Plasma Metabolites"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    text = element_text(size = 9),
    legend.position = "none"
  )


volcano.plasma



```
 
```{r stool metabolites - volcano plot}

## Read in the data
glmms_results <- readxl::read_excel("GLMMs_metabo.stool_long_2025-10-08.xlsx", sheet = 2)

## Format data

glmms_results <- glmms_results %>%
  mutate(
    sig_color = case_when(
      p.value < 0.026 ~ "FDR < 0.25",#"0",
      p.value < 0.05 ~ "p < 0.05",#"1", 
      TRUE ~ "p >= 0.05"), #"2"),
    p.value = ifelse(feature_clean %in% c("Estrone (E1)", "Hypoxanthine"), 0, p.value)
    )

## Run plot - BASIC ####

# volcano.stool <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value))) +
#   geom_point(aes(color = sig_color), alpha = 0.9, size = 2) +
#   geom_vline(xintercept = 0, linetype = "solid", color = "black") +
#   geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray30") +
#   geom_hline(yintercept = -log10(0.026), linetype = "dashed", color = "red") +
#   scale_color_manual(
#     name = NULL,
#     values = c("0" = "skyblue","1" = "darkblue", "2" = "grey70")
#     ) +
#   scale_x_continuous(limits = c(-3.85, 3.57)) +
#   scale_y_continuous(limits = c(0, 4.105)) +
#   labs(
#     x = "log2(Fold Change)", #because of the log transformation
#     y = "-log10(p value)",
#     #color = NULL,
#     title = "Stool Metabolites" 
#   ) +
#   theme_minimal() +
#   theme(
#     plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
#     legend.position = "none",#"right",
#     text = element_text(size = 9)
#   )
# 
# volcano.stool


### ADDING A CUSTOM LEGEND ####


volcano.stool <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value))) +
  # Main points
  geom_point(aes(color = sig_color), 
             alpha = case_when(
               glmms_results$sig_color == "FDR < 0.25" ~ 0.9,
               glmms_results$sig_color == "p < 0.05" ~ 0.3,
               glmms_results$sig_color == "p >= 0.05" ~ 0.5
             ),
             size = 2) +
  
  # Reference lines
  geom_vline(xintercept = 0, linetype = "solid", color = "gray80") +
  #geom_hline(aes(yintercept = -log10(0.05), linetype = "p = 0.05"), color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray70") +
  #geom_hline(aes(yintercept = -log10(0.026), linetype = "FDR = 0.25"), color = "red") +
  geom_hline(yintercept = -log10(0.026), linetype = "dashed", color = "black") +

  # Manual scales
  scale_color_manual(
    name = "Feature",
    values = c(
      "FDR < 0.25" = "darkblue",
      "p < 0.05" = "darkblue",
      "p >= 0.05" = "grey70"
    )
  ) +
  scale_linetype_manual(
    name = "Thresholds",
    values = c("p = 0.05" = "dashed", "FDR = 0.25" = "dashed")
  ) +
  # Combine legends properly
  guides(
    color = guide_legend(
      override.aes = list(shape = 16),
      order = 1
    ),
    linetype = guide_legend(
      override.aes = list(color = c("red","black")),
      order = 2)
  ) +
  # Axes and labels
  scale_x_continuous(limits = c(-3.85, 3.57)) +
  scale_y_continuous(limits = c(0, 4.105)) +
  labs(
    x = "log2(Fold Change)",
    y = "-log10(p value)",
    title = "Stool Metabolites"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    text = element_text(size = 9),
    #LEGEND
    legend.text = element_text(size = 6.8),
    legend.title = element_text(size = 8),
    legend.position = c(-0.1, 1.1),    # inside top left (x,y in npc)
    #legend.justification = c("right", "bottom"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.5, "lines"),
    legend.spacing.y = unit(0.1, "cm"), 
    legend.spacing.x = unit(0.1, "cm"),
    legend.margin = margin(t = 0.1, r = 0.1, b = 0.1, l = 0.1, unit = "cm")
    

  )

volcano.stool

```

```{r COMBINED - volcano plot}
 
 
library(cowplot)

png("GLMMs_volcano.metabo_2025-09-24.png", 
    width = 6.5 * 600, height = 3.25 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #plots
    volcano.plasma, volcano.stool,
    #arguments
    ncol = 2,
    rel_widths = c(1,1)
  )
)


dev.off()



```

 
Relative Abundance Plots
```{r plasma metabolites - relabun plot}

## Read in the data

glmms_results <- readxl::read_excel("GLMMs_metabo.plasma_long_2025-07-15.xlsx") %>% 
  #remove last two cols because model printed info
  select(1:(ncol(.) - 2)) %>% 
  mutate(metadata = fct_recode(term,
                      "Case status" = "case_con1",
                      "Time point" = "collection_part2",
                      "Interaction (case*time)" = "case_con1:collection_part2") )

features <- glmms_results %>% filter(q.value < 0.25) %>% distinct(feature) %>% pull(feature)

## Metaphlan data
glmms_data <- metabo.plasma_long %>% 
  filter(Metabolite %in% features) %>% 
  select(study_id, Metabolite, relative_abundance) %>% 
  merge(., gmbc_data %>% select(study_id, single_id, collection_part, case_con), by="study_id") %>% 
  mutate(case_con = factor(case_con, 
                           levels = c(0, 1), 
                           labels = c('Control', 'Case')),
         collection_part = factor(collection_part, 
                                  levels = c(1, 2), 
                                  labels = c("Baseline", "Follow-up")),
         group = paste(case_con, collection_part)
  ) %>% 
  merge(., glmms_results %>% filter(q.value <0.25) %>% select(feature, metadata), by.x="Metabolite", by.y="feature")


## Select relevant features to plot
glmms_data.plot <- glmms_data %>% 
  filter(Metabolite %in% c(
    #interaction term
    "Estrone.3-sulfate",
    "Zymostenol", 	
    "Adipoyl-carnitine",
    #previous key findings
    "16a-Hydroxyestrone",
    "Matairesinol",
    "Genistein",
    "Pinoresinol",
    "Estriol.(E3)",
    "Glycoursodeoxycholic acid",
    "Hyocholic acid 3-sulfate",
    "3-Methylglutaryl-carnitine"
  )) %>% 
  #clean up names
  mutate(Metabolite = Metabolite %>%
          sub("^[sgt]__", "", .) %>%         # Remove leading s__/g__/t__
          gsub("[_.]", " ", .)                 # Replace underscores with spaces
    ) %>% 
  #filter out zero vals
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance + 1e-6))

## Colours
color_values = c(
      "Control Baseline" = "grey30",
      "Case Baseline" = "blue4",
      "Control Follow-up" = "grey60",
      "Case Follow-up" = "skyblue"
    )


# Label positions per metabolite
metabolite_labels <- glmms_data.plot %>%
  group_by(Metabolite, metadata) %>%
  summarize(x_pos = max(log_rel_abundance, na.rm = TRUE) + 0.2, .groups = "drop")


## Plotting
ggplot(glmms_data.plot, aes(y=Metabolite, x=log_rel_abundance, color=group)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.25) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(Metabolite, group), fill = group),
               position = position_dodge(width = 0.8),
               alpha = 0.5,
               outlier.shape = NA) +
  # Add annotations for metadata term
  geom_text(data = metabolite_labels,
            aes(x = x_pos, y = Metabolite, label = metadata),
            inherit.aes = FALSE,
            hjust = 0, size = 3, color = "black") +
  labs(y=NULL, x="log10(Relative Abundance)") +
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = color_values) +
  scale_fill_manual(name = NULL, values = color_values) +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        #legend.position = leg_pos,
        legend.background = element_rect(color="black", fill = NA),
        legend.margin = margin(t=1, r=3, b=3)) 
  




```

```{r stool metabolites - relabun plot}

## Read in the data

glmms_results <- readxl::read_excel("GLMMs_metabo.plasma_long_2025-07-15.xlsx") %>% 
  #remove last two cols because model printed info
  select(1:(ncol(.) - 2)) %>% 
  mutate(metadata = fct_recode(term,
                      "Case status" = "case_con1",
                      "Time point" = "collection_part2",
                      "Interaction (case*time)" = "case_con1:collection_part2") )

features <- glmms_results %>% filter(q.value < 0.25) %>% distinct(feature) %>% pull(feature)

## Metaphlan data
glmms_data <- metabo.plasma_long %>% 
  filter(Metabolite %in% features) %>% 
  select(study_id, Metabolite, relative_abundance) %>% 
  merge(., gmbc_data %>% select(study_id, single_id, collection_part, case_con), by="study_id") %>% 
  mutate(case_con = factor(case_con, 
                           levels = c(0, 1), 
                           labels = c('Control', 'Case')),
         collection_part = factor(collection_part, 
                                  levels = c(1, 2), 
                                  labels = c("Baseline", "Follow-up")),
         group = paste(case_con, collection_part)
  ) %>% 
  merge(., glmms_results %>% filter(q.value <0.25) %>% select(feature, metadata), by.x="Metabolite", by.y="feature")


## Select relevant features to plot
glmms_data.plot <- glmms_data %>% 
  filter(Metabolite %in% c(
    #interaction term
    "Estrone.3-sulfate",
    "Zymostenol", 	
    "Adipoyl-carnitine",
    #previous key findings
    "16a-Hydroxyestrone",
    "Matairesinol",
    "Genistein",
    "Pinoresinol",
    "Estriol.(E3)",
    "Glycoursodeoxycholic acid",
    "Hyocholic acid 3-sulfate",
    "3-Methylglutaryl-carnitine"
  )) %>% 
  #clean up names
  mutate(Metabolite = Metabolite %>%
          sub("^[sgt]__", "", .) %>%         # Remove leading s__/g__/t__
          gsub("[_.]", " ", .)                 # Replace underscores with spaces
    ) %>% 
  #filter out zero vals
  filter(relative_abundance > 0) %>% 
  #log transform for visualization on plot
  mutate(log_rel_abundance = log10(relative_abundance + 1e-6))

## Colours
color_values = c(
      "Control Baseline" = "grey30",
      "Case Baseline" = "blue4",
      "Control Follow-up" = "grey60",
      "Case Follow-up" = "skyblue"
    )


# Label positions per metabolite
metabolite_labels <- glmms_data.plot %>%
  group_by(Metabolite, metadata) %>%
  summarize(x_pos = max(log_rel_abundance, na.rm = TRUE) + 0.2, .groups = "drop")



## Plotting
ggplot(glmms_data.plot, aes(y=Metabolite, x=log_rel_abundance, color=group)) +
  #adding in the individual rel_abun points
  geom_jitter(position = position_jitterdodge(dodge.width = 0.8,
                                              jitter.width = 0.1),
              shape=21, alpha = 0.25) +
  #boxplots per group per species
  geom_boxplot(aes(group = interaction(Metabolite, group), fill = group),
               position = position_dodge(width = 0.8),
               alpha = 0.5,
               outlier.shape = NA) +
  # Add annotations for metadata term
  geom_text(data = metabolite_labels,
            aes(x = x_pos, y = Metabolite, label = metadata),
            inherit.aes = FALSE,
            hjust = 0, size = 3, color = "black") +
  labs(y=NULL, x="log10(Relative Abundance)") +
  #manually setting colours of the groups
  scale_color_manual(name = NULL, values = color_values) +
  scale_fill_manual(name = NULL, values = color_values) +
  #overall theme
  theme_classic() +
  theme(axis.text.y = element_markdown(),
        legend.text = element_markdown(),
        #legend.position = leg_pos,
        legend.background = element_rect(color="black", fill = NA),
        legend.margin = margin(t=1, r=3, b=3)) 
  




```




### ENDOCRINE SUBGROUP ANALYSIS


Multi-level PLS-DA   
```{r mPLS-DA therapy - plasma }

#library(ropls)
library(mixOmics)
library(dplyr)

## Prepare Data ####

gmbc_data$t_endo_drug.cat <- factor(gmbc_data$t_endo_drug.cat,
                                       levels = c("None","Aromatase_inhibitor","Tamoxifen"),
                                      labels = c("None","A Inhibitor","ER Blocker"))

#dataset
oplsda <- metabo.plasma_wide %>% 
  transform_metabo(., remove_outliers = F, merge = F) %>% 
  merge(., gmbc_data %>% dplyr::select(study_id, single_id, case_con, collection_part, t_endo_drug.cat), by="study_id") %>% 
  ## FILTER CASES ONLY
  filter(case_con == 1) %>% 
  select(-case_con) 

# predictors (metabolites)
X <- oplsda %>% dplyr::select(-c(study_id, single_id, collection_part,t_endo_drug.cat)) %>% as.matrix()
# outcome variable (e.g., case vs. control)
Y <- as.factor(oplsda$t_endo_drug.cat)  
# subject ids for repeated measures
subject <- oplsda$single_id


## Run multilevel PLS-DA (accounts for repeated measures)
plsda_model <- plsda(X, Y, multilevel = subject, ncom = 2)



## PRINT FEATURES / PERFORMANCE ####

# ## Important Variables
# vipVn.plasma <- vip(plsda_model) %>%   # Variable Importance in Projection (VIP) scores
#   as.data.frame() %>%
#   arrange(desc(.))
# write.csv(vipVn.plasma, "/Users/ashlarnder/Downloads/vip.plasma-therapy.csv", row.names = TRUE)
# head(vipVn.plasma, 10)


## Explained variance
plsda_model$prop_expl_var$X

# Run repeated cross-validation for mPLS-DA
perf_mplsda <- perf(
  plsda_model,
  validation = "Mfold",
  folds = 5,
  nrepeat = 5,
  dist = "centroids.dist"
)

# Classification error rates
# Overall classification error per component
perf_mplsda$error.rate$overall
# Balanced error rate (BER)
perf_mplsda$error.rate$BER

## Scores Plotting ####

#predictive component scores
scores <- plsda_model$variates$X
#including group labels
scores <- data.frame(
    comp1 = scores[, 1],
    comp2 = scores[, 2],
    t_endo_drug.cat = Y,
    time = oplsda$collection_part,
    ID = subject
  ) %>% 
  mutate(group_label = case_when(
      t_endo_drug.cat == "None" & time == 1 ~ "None, Baseline",
      t_endo_drug.cat == "None" & time == 2 ~ "None, Follow-up",
      t_endo_drug.cat == "A Inhibitor" & time == 1 ~ "A Inhibitor, Baseline",
      t_endo_drug.cat == "A Inhibitor" & time == 2 ~ "A Inhibitor, Follow-up",
      t_endo_drug.cat == "ER Blocker" & time == 1 ~ "ER Blocker, Baseline",
      t_endo_drug.cat == "ER Blocker" & time == 2 ~ "ER Blocker, Follow-up",
  ))
#get explained variance for predictive components
expl_var <- plsda_model$prop_expl_var$X

#calculate the median coordinates for each group
medians <- scores %>%
  group_by(group_label) %>%
  summarize(
    med_comp1 = median(comp1), 
    med_comp2 = median(comp2),
    .groups = "drop"
  )

#set colours by group
group_colors <- c(
    "None, Baseline" = "black",
    "None, Follow-up" = "darkgrey",
    "A Inhibitor, Baseline"= "sienna",  
    "A Inhibitor, Follow-up" = "wheat3",
    "ER Blocker, Baseline"= "darkblue", 
    "ER Blocker, Follow-up" = "#6ba"    
)


#build the plot
pls_plot.plasma <- ggplot(scores, aes(x = comp1, y = comp2, 
                                      color = group_label, shape = time)) +
  geom_point(alpha=0.6) +
  stat_ellipse(aes(group = group_label), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = group_colors) +# c("0" = "darkgray", "1" = "blue")) +
  scale_shape_manual(values = c("1" = 16, "2" = 17)) + #circle baseline #triangle FU
  theme_minimal() +
  labs(   
    x = paste0("Component 1 (", round(expl_var[[1]], 2), "% exp var)"),
    y = paste0("Component 2 (", round(expl_var[[2]], 2), "% exp var)"),
    color = "Case_Status",
    shape = "Time Point"
  ) +
  theme(legend.position = "none",#"right",
        panel.background = element_blank(),
        plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "black", fill = NA),
        panel.spacing = unit(0.5, "lines")) +
  
  #add lines from each point to the median of its group with matching colors
  geom_segment(data = scores %>% 
                 left_join(medians, by = "group_label"),
               aes(x = med_comp1, y = med_comp2, xend = comp1, yend = comp2,
                   color = group_label),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  #add median points with matching colors
  geom_point(data = medians, 
             aes(x = med_comp1, y = med_comp2, color = group_label), 
             size = 2.5, shape = 4, stroke = 1.2)  # Shape 4 is an 'x' marker

pls_plot.plasma



```
 
```{r mPLS-DA therapy - stool }

#library(ropls)
library(mixOmics)
library(dplyr)
 
## Prepare Data ####

#dataset
oplsda <- metabo.stool_wide %>% 
  transform_metabo(., remove_outliers = F, merge = F) %>% 
  merge(., gmbc_data %>% dplyr::select(study_id, single_id, case_con, collection_part, t_endo_drug.cat), by="study_id") %>% 
  ## FILTER CASES ONLY
  filter(case_con == 1) %>% 
  select(-case_con) 

# predictors (metabolites)
X <- oplsda %>% dplyr::select(-c(study_id, single_id, collection_part,t_endo_drug.cat)) %>% as.matrix()
# outcome variable (e.g., case vs. control)
Y <- as.factor(oplsda$t_endo_drug.cat)  
# subject ids for repeated measures
subject <- oplsda$single_id


## Run multilevel PLS-DA (accounts for repeated measures)
plsda_model <- plsda(X, Y, multilevel = subject, ncom = 2)



# PRINT FEATURES / PERFORMANCE ####

# ## Important Variables
# vipVn.plasma <- vip(plsda_model) %>%   # Variable Importance in Projection (VIP) scores
#   as.data.frame() %>%
#   arrange(desc(.))
# write.csv(vipVn.plasma, "/Users/ashlarnder/Downloads/vip.stool-therapy.csv", row.names = TRUE)
# head(vipVn.plasma, 10)


## Explained variance
plsda_model$prop_expl_var$X

# Run repeated cross-validation for mPLS-DA
perf_mplsda <- perf(
  plsda_model,
  validation = "Mfold",
  folds = 5,
  nrepeat = 5,
  dist = "centroids.dist"
)

# Classification error rates
# Overall classification error per component
perf_mplsda$error.rate$overall
# Balanced error rate (BER)
perf_mplsda$error.rate$BER

## Scores Plotting ####

#predictive component scores
scores <- plsda_model$variates$X
#including group labels
scores <- data.frame(
    comp1 = scores[, 1],
    comp2 = scores[, 2],
    t_endo_drug.cat = Y,
    time = oplsda$collection_part,
    ID = subject
  ) %>% 
  mutate(group_label = case_when(
      t_endo_drug.cat == "None" & time == 1 ~ "None, Baseline",
      t_endo_drug.cat == "None" & time == 2 ~ "None, Follow-up",
      t_endo_drug.cat == "A Inhibitor" & time == 1 ~ "A Inhibitor, Baseline",
      t_endo_drug.cat == "A Inhibitor" & time == 2 ~ "A Inhibitor, Follow-up",
      t_endo_drug.cat == "ER Blocker" & time == 1 ~ "ER Blocker, Baseline",
      t_endo_drug.cat == "ER Blocker" & time == 2 ~ "ER Blocker, Follow-up",
  ))
#get explained variance for predictive components
expl_var <- plsda_model$prop_expl_var$X

#calculate the median coordinates for each group
medians <- scores %>%
  group_by(group_label) %>%
  summarize(
    med_comp1 = median(comp1), 
    med_comp2 = median(comp2),
    .groups = "drop"
  )

#set colours by group
group_colors <- c(
    "None, Baseline" = "black",
    "None, Follow-up" = "darkgrey",
    "A Inhibitor, Baseline"= "sienna",  
    "A Inhibitor, Follow-up" = "wheat3",
    "ER Blocker, Baseline"= "darkblue", 
    "ER Blocker, Follow-up" = "#6ba"    
)


#build the plot
pls_plot.stool <- ggplot(scores, aes(x = comp1, y = comp2, 
                                     color = group_label, shape = time)) +
  geom_point(alpha=0.6) +
  stat_ellipse(aes(group = group_label), level = 0.95, linewidth = 0.6) +
  scale_color_manual(values = group_colors) +
  scale_shape_manual(values = c("1" = 16, "2" = 17)) +  # baseline = circle, FU = triangle
  theme_minimal() +
  ylim(-7, NA) +
  labs(   
    x = paste0("Component 1 (", round(expl_var[[1]], 2), "% exp var)"),
    y = paste0("Component 2 (", round(expl_var[[2]], 2), "% exp var)"),
    color = NULL,
    shape = NULL  # remove separate shape title
  ) +
  theme(
    panel.background = element_blank(),
    plot.background = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(0.5, "lines"),
    # LEGEND
    legend.text = element_text(size = 6.5),
    legend.title = element_text(size = 8),
    legend.position = c(0.03, 0.2), #0.01, 0.999
    legend.box = "horizontal",
    legend.justification = c("left", "top"),
    legend.background = element_rect(color = "black", fill = "white"),
    legend.margin  = margin(t = 2, r = 2, b = 2, l = 2),
    legend.key.size = unit(0.7, "lines"),
    legend.spacing.y = unit(0.1, "cm"), 
    legend.spacing.x = unit(0.1, "cm")
  ) +
  guides(
    color = guide_legend(
      ncol = 2, byrow = TRUE, order = 1,
      override.aes = list(shape = c(16, 17)) # all shapes same in legend
    ),
    shape = "none"  # remove separate shape legend
  ) +
  geom_segment(data = scores %>% left_join(medians, by = "group_label"),
               aes(x = med_comp1, y = med_comp2, xend = comp1, yend = comp2,
                   color = group_label),
               linewidth = 0.2, alpha = 0.5, inherit.aes = FALSE) +
  geom_point(data = medians, 
             aes(x = med_comp1, y = med_comp2, color = group_label), 
             size = 2.5, shape = 4, stroke = 1.2)

pls_plot.stool 


```
 
```{r combined plot}

library(cowplot)
library(grid)


png("mpls-therapy_2025-09-23.png", 
    width = 6.5 * 600, height = 3.25 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    pls_plot.plasma, pls_plot.stool,
    nrow = 1,
    rel_widths = c(1, 1)
  )
)
  
grid.text("Plasma", x = unit(0.14, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

grid.text("Stool", x = unit(0.64, "npc"), y = unit(0.94, "npc"), 
          gp = gpar(fontsize = 15, fontface = "bold"))

dev.off()

```

 
GLMMs
```{r GLMMs therapy groups - plasma }

gmbc_data$t_endo_drug.cat <- factor(gmbc_data$t_endo_drug.cat,
                                       levels = c("None","Aromatase_inhibitor","Tamoxifen"))

## MAKE SURE YOU TURN THE APPLY FUNCTION TO BE ".therapy" in the modelling function being used!!
apply_iterative.modeling.therapy(
  predictor_vars = metabo.plasma_wide %>% select(-study_id) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean",
        "t_rad.cat","t_chemo.cat" 
    )
  ),
  dataset.model = metabo.plasma_wide %>% transform_metabo(),
  file_name = "metabo.plasma_long-therapy"
)

```

```{r GLMMs therapy groups - stool }

gmbc_data$t_endo_drug.cat <- factor(gmbc_data$t_endo_drug.cat,
                                       levels = c("None","Aromatase_inhibitor","Tamoxifen"))

## MAKE SURE YOU TURN THE APPLY FUNCTION TO BE ".therapy" in the modelling function being used!!
apply_iterative.modeling.therapy(
  predictor_vars = metabo.stool_wide %>% select(-study_id) %>% colnames(), 
  adjustment_sets = list(
    unadj = c(),
    confounder_vars = c(
        "s_age","s_bmi","ethnic_collapsed","d_education_collapsed",
        "hormone_current.p1","sm_summary","pa_cat","Alcohol.g",
        "fu_antibiotic.clean",
        "t_rad.cat","t_chemo.cat" 
    )
  ),
  dataset.model = metabo.stool_wide %>% transform_metabo(),
  file_name = "metabo.stool_long-therapy"
)

```


```{r plasma metabolites - volcano plot}

## Read in the data
glmms_results <- readxl::read_excel("GLMMs_metabo.plasma_long-therapy_2025-09-23.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = case_when(
      (term == "Interaction (A Inhibitor)" & p.value < 0.05) ~ "1",
      (term == "Interaction(ER Blocker)" & p.value < 0.05) ~ "2",
      TRUE ~ "3"),
    term = ifelse(term=="Interaction (A Inhibitor)", "A Inhibitor","ER Blocker")
    )

## Run plot
volcano.plasma <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(shape = term, color = sig_color), alpha = 0.9, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_shape_manual(values = c("A Inhibitor" = 16, "ER Blocker" = 17)) + # circle & triangle
  scale_color_manual(values = c("1" = "sienna", "2" = "darkblue", "3" = "grey70")) +
  scale_x_continuous(limits = c(-6.36, 5.15)) +
  scale_y_continuous(limits = c(0, 4.48)) +
  labs(
    x = "log2(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, #"Significance",
    shape = "Interactions",
    title = "Plasma"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    text = element_text(size = 9),
    legend.position = "none"#"right",
  )

volcano.plasma



```
  
```{r stool metabolites - volcano plot}

## Read in the data
glmms_results <- readxl::read_excel("GLMMs_metabo.stool_long-therapy_2025-09-23.xlsx", sheet = 2)

## Format data
glmms_results <- glmms_results %>%
  mutate(
    sig_color = case_when(
      (term == "Interaction (A Inhibitor)" & p.value < 0.05) ~ "1",
      (term == "Interaction(ER Blocker)" & p.value < 0.05) ~ "2",
      TRUE ~ "3"),
    term = ifelse(term=="Interaction (A Inhibitor)", "A Inhibitor","ER Blocker"),
    p.value = ifelse(feature_clean=="Estrone (E1)" & term=="A Inhibitor", 0, p.value)
  )

## Run plot
volcano.stool <- ggplot(glmms_results, aes(x = estimate, y = -log10(p.value)) ) +
  geom_point(aes(shape = term, color = sig_color), alpha = 0.9, size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  scale_shape_manual(values = c("A Inhibitor" = 16, "ER Blocker" = 17)) + # circle & triangle
  scale_color_manual(values = c("1" = "sienna", "2" = "darkblue", "3" = "grey70")) +
  scale_x_continuous(limits = c(-6.36, 5.15)) +
  scale_y_continuous(limits = c(0, 4.48)) +
  labs(
    x = "log2(Fold Change)", #because of the log transformation
    y = "-log10(p value)",
    color = NULL, #"Significance",
    shape = "Main Effects",
    title = "Stool"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    text = element_text(size = 9),
    #LEGEND
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.position = c(0.84, 0.97),    # inside top right (x,y in npc)
    legend.box = "horizontal",
    legend.background = element_rect(color = "black", fill = "white"),
    legend.key.size = unit(0.7, "lines"),
    legend.spacing.y = unit(0.1, "cm"), 
    legend.spacing.x = unit(0.1, "cm") 
   ) +
  guides(
    shape = guide_legend(override.aes = list(color = c("sienna", "darkblue")) ),  # colors for legend symbols
    color = "none"  # hide color legend
  )

volcano.stool


```

```{r COMBINED - volcano plot}
 
 
library(cowplot)

png("GLMMs_volcano.metabo-therapy_2025-09-24.png", 
    width = 6.5 * 600, height = 3.25 * 600, 
    res = 600, bg = "white")

# Arrange side-by-side with custom widths
print(
  plot_grid(
    #plots
    volcano.plasma, volcano.stool,
    #arguments
    ncol = 2,
    rel_widths = c(1,1)
  )
)


dev.off()



```
